---
title: "Integrated Step Selection Function Models applied to the black lion tamarin - scan movement data"
format: pdf
editor: visual
authors: 
  - name: Felipe Bufalo
  - name: Rodrigo Gonçalves Amaral
  - name: Gabriel Pavan Sabino
  - name: Luca Börger
  - name: Laurence Culot
bibliography: references.bib
---

# Introduction

In this script I will use the environmental, energetic, and social landscapes predicted previously to fit *i*SSF models to investigate what factors are associated with black lion tamarins' (*Leontopithecus chrysopygus -* BLTs) route-planning processes of where and when to move to in a 100 ha seasonal Atlantic Forest Fragment in the municipality of Guareí, in São Paulo, Brazil.

## Loading packages

```{r}
#| output: FALSE
#| echo: TRUE
library(amt)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(tidyr)
library(terra)
```

## Loading data

```{r}
load("2_3_data_SSF_BLT_2025_07_d02.rda")

# List all .tif files in the "rasters" folder
raster_files <- list.files("rasters", pattern = "\\.tif$", full.names = TRUE)

# Loop to load and assign each raster as an individual object
for (file in raster_files) {
  var_name <- tools::file_path_sans_ext(basename(file))
  assign(var_name, rast(file), envir = .GlobalEnv)
}
```

# 1. Create SSF object

Now that all possible predictors have been modeled and landscapes have been created, I can create my ssf object.

I will use steps with 15 minutes of interval between relocations. Also, I will generate 100 random steps for each observed step. Predictor values will be extracted for both start and end locations of each step (observed and random).

I will create one object for the rainy season (ssf_wet) and another for the dry season (ssf_dry).

```{r}
dat <- df
names(dat)
dat <- dat[,-c(1,2,11:14, 17:18)]
dat <- dat %>%
  rename(x = original_x, y = original_y) %>% 
  filter(id == "group")
dat$x <- as.numeric(dat$x)
dat$y <- as.numeric(dat$y)
head(dat)

# Check if all observations are complete
all(complete.cases(dat)) # no action required

# Parse date and time and create time stamps
dat$ts <- as.POSIXct(paste(dat$date, dat$time),
                    format="%d/%m/%y %H:%M:%S",
                    tz = "Etc/GMT+3")

tr <- make_track(dat, x, y, ts, id = id, 
                 behavior = behavior, season = season,
                 crs = 32722)

ssf_wet <- tr %>% 
  filter(season == "wet") %>% 
  track_resample(rate = minutes(15), tolerance = seconds(10)) %>% 
  steps_by_burst() %>% 
  random_steps(n_control = 100) %>% 
  extract_covariates(r_dist_to_border, where = "both") %>% 
  extract_covariates(nt_bot_r, where = "both") %>% 
  extract_covariates(rch_1, where = "both") %>%
  extract_covariates(nft_bot_wet_r, where = "both") %>% 
  extract_covariates(nvft_wet_r, where = "both") %>% 
  extract_covariates(KJ_bot_wet_r, where = "both") %>% 
  extract_covariates(KJ_vft_wet_r, where = "both") %>% 
  extract_covariates(vedba_wet_r, where = "both") %>% 
  extract_covariates(vedba_all_r, where = "both") %>% 
  extract_covariates(r_dist_social_border, where = "both") %>% 
  extract_covariates(UD_cons_r, where = "both") %>%
  extract_covariates(n_lc_wet_r, where = "both") %>% 
  mutate(dist_f_b_start = scale(dist_f_b_start),
         dist_f_b_end = scale(dist_f_b_end),
         nt_bot_start = scale(nt_bot_start), 
         nt_bot_end = scale(nt_bot_end),
         tree_H_start  = scale(tree_H_start),
         tree_H_end = scale(tree_H_end),
         nft_bot_start = scale(nft_bot_start),
         nft_bot_end = scale(nft_bot_end),
         n_vft_start = scale(n_vft_start),
         n_vft_end = scale(n_vft_end),
         KJ_bot_start = scale(KJ_bot_start),
         KJ_bot_end = scale(KJ_bot_end),
         KJ_vft_start = scale(KJ_vft_start),
         KJ_vft_end = scale(KJ_vft_end),
         VeDBA_start = scale(VeDBA_start),
         VeDBA_end = scale(VeDBA_end),
         VeDBA_all_start = scale(VeDBA_all_start),
         VeDBA_all_end = scale(VeDBA_all_end),
         dist_s_b_start = scale(dist_s_b_start),
         dist_s_b_end = scale(dist_s_b_end),
         UDcons_start = scale(UDcons_start),
         UDcons_end = scale(UDcons_end),
         n_lc_start = scale(n_lc_start),
         n_lc_end = scale(n_lc_end),
         cos_ta_ = cos(ta_), 
         log_sl_ = log(sl_)
  ) %>% 
  filter(!is.na(ta_))
names(ssf_wet)

ssf_dry <- tr %>% 
  filter(season == "dry") %>% 
  track_resample(rate = minutes(15), tolerance = seconds(10)) %>% 
  steps_by_burst() %>% 
  random_steps(n_control = 100) %>% 
  extract_covariates(r_dist_to_border, where = "both") %>% 
  extract_covariates(nt_bot_r, where = "both") %>% 
  extract_covariates(rch_1, where = "both") %>%
  extract_covariates(nft_bot_dry_r, where = "both") %>% 
  extract_covariates(nvft_dry_r, where = "both") %>% 
  extract_covariates(KJ_bot_dry_r, where = "both") %>% 
  extract_covariates(KJ_vft_dry_r, where = "both") %>% 
  extract_covariates(inv_bio_dry_r, where = "both") %>% 
  extract_covariates(n_alarm_dry_r, where = "both") %>% 
  extract_covariates(vedba_dry_r, where = "both") %>% 
  extract_covariates(r_dist_social_border, where = "both") %>% 
  extract_covariates(UD_cons_r, where = "both") %>%
  extract_covariates(n_lc_dry_r, where = "both") %>% 
  mutate(dist_f_b_start = scale(dist_f_b_start),
         dist_f_b_end = scale(dist_f_b_end),
         nt_bot_start = scale(nt_bot_start), 
         nt_bot_end = scale(nt_bot_end),
         tree_H_start  = scale(tree_H_start),
         tree_H_end = scale(tree_H_end),
         nft_bot_start = scale(nft_bot_start),
         nft_bot_end = scale(nft_bot_end),
         n_vft_start = scale(n_vft_start),
         n_vft_end = scale(n_vft_end),
         KJ_bot_start = scale(KJ_bot_start),
         KJ_bot_end = scale(KJ_bot_end),
         KJ_vft_start = scale(KJ_vft_start),
         KJ_vft_end = scale(KJ_vft_end),
         inv_bio_start = scale(inv_bio_start),
         inv_bio_end = scale(inv_bio_end),
         n_alarm_start = scale(n_alarm_start),
         n_alarm_end = scale(n_alarm_end),
         VeDBA_start = scale(VeDBA_start),
         VeDBA_end = scale(VeDBA_end),
         dist_s_b_start = scale(dist_s_b_start),
         dist_s_b_end = scale(dist_s_b_end),
         UDcons_start = scale(UDcons_start),
         UDcons_end = scale(UDcons_end),
         n_lc_start = scale(n_lc_start),
         n_lc_end = scale(n_lc_end),
         cos_ta_ = cos(ta_), 
         log_sl_ = log(sl_)
  ) %>% 
  filter(!is.na(ta_))
names(ssf_dry)
```

Make a few adjustments in the data

```{r}
#| warning: false
ssf_wet <- ssf_wet %>% filter(!is.na(n_vft_end))
ssf_dry <- ssf_dry %>% filter(!is.na(tree_H_end))
ssf_wet$UDcons_end[is.na(ssf_wet$UDcons_end)] <- 0
ssf_wet$tree_H_end[is.na(ssf_wet$tree_H_end)] <- 0
ssf_wet$dist_f_b_end[is.na(ssf_wet$dist_f_b_end)] <- 0
ssf_wet$dist_s_b_end[is.na(ssf_wet$dist_s_b_end)] <- 0
ssf_dry$dist_f_b_end[is.na(ssf_dry$dist_f_b_end)] <- 0
ssf_dry$dist_s_b_end[is.na(ssf_dry$dist_s_b_end)] <- 0
ssf_dry$UDcons_end[is.na(ssf_dry$UDcons_end)] <- 0
```

# 2. Check SSF covariates

Check for colinearity between covariates

```{r}
columns_of_interest_wet <- ssf_wet[,c(13:36,6,7,37,38)]
columns_of_interest_dry <- ssf_dry[,c(13:38,6,7,39,40)]
# Convert to data frame if necessary
columns_of_interest_wet <- as.data.frame(columns_of_interest_wet)
columns_of_interest_dry <- as.data.frame(columns_of_interest_dry)
# Perform correlation test
correlation_matrix_wet <- cor(columns_of_interest_wet)
correlation_matrix_dry <- cor(columns_of_interest_dry)
# Convert correlation matrix to long format
correlation_long_wet <- as.data.frame(as.table(correlation_matrix_wet))
names(correlation_long_wet) <- c("Var1", "Var2", "Correlation")
correlation_long_dry <- as.data.frame(as.table(correlation_matrix_dry))
names(correlation_long_dry) <- c("Var1", "Var2", "Correlation")
```

In @fig-corplot_wet we can see the correlation between all covariates in ssf object in the rainy season

```{r}
#| label: fig-corplot_wet
#| fig-width: 7
#| fig-height: 6
#| fig-alt: |
#|   A plot showing the correlation between all covariates in ssf_wet
#| fig-cap: |
#|       Correlation between all covariates in ssf_wet
#| warning: FALSE     
ggplot(correlation_long_wet, aes(Var1, Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 1.8) +  
  scale_fill_gradient2(low = "#565656", mid = "white", high = "#4f7160", 
                       midpoint = 0, 
                       limits = c(-1, 1), name = "Correlation") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Correlation Heatmap - rainy season")+
  xlab("")+
  ylab("")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black",
                                    fill=NA, size=2))+
  theme(plot.title = element_text(hjust = 0.5),
        title=element_text(size=16),
        legend.title = element_blank())
```

In @fig-corplot_dry we can see the correlation between all covariates in ssf object in the dry season

```{r}
#| label: fig-corplot_dry 
#| fig-width: 7
#| fig-height: 6
#| fig-alt: |
#|   A plot showing the correlation between all covariates in ssf_dry
#| fig-cap: |
#|       Correlation between all covariates in ssf object for dry season
#| warning: FALSE     
ggplot(correlation_long_dry, aes(Var1, Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 1.8) +  
  scale_fill_gradient2(low = "#565656", mid = "white", high = "#4f7160", 
                       midpoint = 0, 
                       limits = c(-1, 1), name = "Correlation") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Correlation Heatmap - dry season")+
  xlab("")+
  ylab("")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black",
                                    fill=NA, size=2))+
  theme(plot.title = element_text(hjust = 0.5),
        title=element_text(size=16),
        legend.title = element_blank())
```

### Translating the names used for covariates:

Each covariate has the extracted values for the start and end of each of the observed and randomly generated steps.

| Covariate | Description |
|------------------------------------|------------------------------------|
| `dist_f_b` | Distance to fragment edge (i.e., limits of the forest fragment), in meters. |
| `nt_bot` | Number of trees, computed through vegetation plots. |
| `tree_H` | Canopy height (in meters), extracted from a 1 m resolution raster available from [@tolan2024]. |
| `nft_bot` | Number of fruit-feeding trees, computed through vegetation plots and behavior data. |
| `n_vft` | Number of fruit-feeding trees, computed through behavior data. |
| `KJ_bot` | Fruit energetic content (KJ), computed through vegetation plots, behavior data, and fruit chemical content. |
| `KJ_vft` | Fruit energetic content (KJ), computed through behavior data and fruit chemical content. |
| `dist_s_b` | Distance to the border of the 95% AKDEc estimation of the used area of the BLT study group, in meters - social border. |
| `UDcons` | Summed probability of conspecific Utilization Densities, computed using [Axytrek mini dataloggers from Technosmart](https://www.technosmart.eu/axy-trek-mini/) installed on two individuals from each of the three groups in the area. |
| `n_lc` | Number of inter group long calls, computed through behavior data. |
| `inv_bio` | Invertebrate biomass (in grams), computed through pitfall trap experiments. |
| `n_alarm` | Number of alarm calls, computed through behavior data. |
| `VeDBA` | Average Vectorial Dynamic Body Acceleration for each coordinate, computed using accelerometer data in [Axytrek mini dataloggers from Technosmart](https://www.technosmart.eu/axy-trek-mini/) for each season. |
| `VeDBA_all` | Average Vectorial Dynamic Body Acceleration for each coordinate, computed using accelerometer data in [Axytrek mini dataloggers from Technosmart](https://www.technosmart.eu/axy-trek-mini/) across both seasons. Used just for modeling rainy season. |
| `sl_` | Tentative step length, computed by the `amt` package using a gamma distribution. |
| `ta_` | Tentative turn angle, computed by the `amt` package using a von Mises distribution. |
| `log_sl_` | Log of tentative step length. |
| `cos_ta_` | Cosine of tentative turn angle. |

# 3. *i*SSF models

## 3.1 Rainy season - model fit

I will fit models with interactions with movement parameters (*sl\_*, *log_sl\_*, *ta\_* and *cos_ta\_*) and and without the interactions

### 3.1.1 Forest structure model

```{r}
frst_st_model_int_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ dist_f_b_end + nt_bot_end + tree_H_end +  
               sl_ + log_sl_ + cos_ta_ + 
               dist_f_b_start:(sl_ + log_sl_ + cos_ta_) +
               nt_bot_start:(sl_ + log_sl_ + cos_ta_) +
               tree_H_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

frst_st_model_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ dist_f_b_end + nt_bot_end + tree_H_end + sl_ + log_sl_ + cos_ta_ + 
               strata(step_id_), model=TRUE)

```

### 3.1.2 Fruits model

```{r}
fruits_bot_model_int_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ nft_bot_end + KJ_bot_end + 
               sl_ + log_sl_ + cos_ta_ + 
               KJ_bot_start:(sl_ + log_sl_ + cos_ta_) +
               nft_bot_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

fruits_bot_model_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ nft_bot_end + KJ_bot_end +
               sl_ + log_sl_ + cos_ta_ + strata(step_id_), 
             model=TRUE)

fruits_vis_model_int_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ n_vft_end + KJ_vft_end + 
               sl_ + log_sl_ + cos_ta_ + 
               KJ_vft_start:(sl_ + log_sl_ + cos_ta_) +
               n_vft_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

fruits_vis_model_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ n_vft_end + KJ_vft_end + 
               sl_ + log_sl_ + cos_ta_ + strata(step_id_), 
             model=TRUE)
```

### 3.1.3 VeDBA model

```{r}
vedba_model_int_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ VeDBA_end + sl_ + log_sl_ + cos_ta_ + 
               VeDBA_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

vedbaall_model_int_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ VeDBA_all_end + sl_ + log_sl_ + cos_ta_ + 
               VeDBA_all_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

vedba_model_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ VeDBA_end + sl_ + log_sl_ + cos_ta_ + 
               strata(step_id_), 
             model=TRUE)

vedbaall_model_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ VeDBA_all_end + sl_ + log_sl_ + cos_ta_ + 
               strata(step_id_), 
             model=TRUE)
```

### 3.1.4 Social model

```{r}
conspecifics_model_int_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ UDcons_end + n_lc_end + sl_ + log_sl_ + cos_ta_ + 
               UDcons_start:(sl_ + log_sl_ + cos_ta_) +
               n_lc_end:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

conspecifics_model_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ UDcons_end + n_lc_end + sl_ + log_sl_ + cos_ta_ +
               strata(step_id_), model=TRUE)

social_border_model_int_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ dist_f_b_end + n_lc_end + sl_ + log_sl_ + cos_ta_ + 
               dist_f_b_start:(sl_ + log_sl_ + cos_ta_) +
               n_lc_end:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

social_border_model_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ dist_f_b_end + n_lc_end + sl_ + log_sl_ + cos_ta_ +
               strata(step_id_), model=TRUE)
```

### 3.1.5 Null model

```{r}
null_model_wet <- ssf_wet %>% 
  fit_clogit(case_ ~ 1 + sl_ + log_sl_ + cos_ta_ + strata(step_id_), 
             model=TRUE)

```

### 3.1.6 Compare models

```{r}
conspecifics_model_int_wet_aic <- AIC(conspecifics_model_int_wet)
conspecifics_model_wet_aic <- AIC(conspecifics_model_wet)
fruits_bot_model_int_wet_aic <- AIC(fruits_bot_model_int_wet)
fruits_bot_model_wet_aic <- AIC(fruits_bot_model_wet)
fruits_vis_model_int_wet_aic <- AIC(fruits_vis_model_int_wet)
fruits_vis_model_wet_aic <- AIC(fruits_vis_model_int_wet)
frst_st_model_int_wet_aic <- AIC(frst_st_model_int_wet)
frst_st_model_wet_aic <- AIC(frst_st_model_wet)
vedba_model_int_wet_aic <- AIC(vedba_model_int_wet)
vedbaall_model_int_wet_aic <- AIC(vedbaall_model_int_wet)
vedba_model_wet_aic <- AIC(vedba_model_wet)
vedbaall_model_wet_aic <- AIC(vedbaall_model_wet)
social_border_model_int_wet_aic <- AIC(social_border_model_int_wet)
social_border_model_wet_aic <- AIC(social_border_model_wet)
null_model_wet_aic <- AIC(null_model_wet)

# Compare AIC values
aic_table_wet <- data.frame(Model = c("conspecifics_model_int_wet", 
                                      "conspecifics_model_wet", 
                                      "fruits_bot_model_int_wet", 
                                      "fruits_bot_model_wet",
                                      "fruits_vis_model_int_wet",
                                      "fruits_vis_model_wet",
                                      "frst_st_model_int_wet",
                                      "frst_st_model_wet", 
                                      "vedba_model_int_wet",
                                      "vedbaall_model_int_wet",
                                      "vedba_model_wet",
                                      "vedbaall_model_wet",
                                      "social_border_model_int_wet",
                                      "social_border_model_wet",
                                      "null_model_wet"),
                            AIC_Value = c(conspecifics_model_int_wet_aic, 
                                          conspecifics_model_wet_aic,
                                          fruits_bot_model_int_wet_aic,
                                          fruits_bot_model_wet_aic, 
                                          fruits_vis_model_int_wet_aic,
                                          fruits_vis_model_wet_aic,
                                          frst_st_model_int_wet_aic,
                                          frst_st_model_wet_aic, 
                                          vedba_model_int_wet_aic,
                                          vedbaall_model_int_wet_aic,
                                          vedba_model_wet_aic,
                                          vedbaall_model_wet_aic,
                                          social_border_model_int_wet_aic,
                                          social_border_model_wet_aic,
                                          null_model_wet_aic))

# Find the index of the best model (i.e., the model with the lowest AIC value)
best_model_index_wet <- which.min(aic_table_wet$AIC_Value)
# Subtract lowest AIC_value from all the rest to create a delta_AIC column
aic_table_wet$delta_AIC<-aic_table_wet$AIC_Value-min(aic_table_wet$AIC_Value)
# Sort the AIC table by AIC values in increasing order
aic_table_sorted_wet <- aic_table_wet[order(aic_table_wet$AIC_Value), ]
print(aic_table_sorted_wet)
```

I can see that *vedba_model_wet*, *vedbaall_model_wet*, and *fruits_bot_model_wet*, are worse or equal to *null_model*.

Best model is *fruits_vis_model_int_wet*:

```{r}
summary(fruits_vis_model_int_wet)
```

Inspect other models better than null:

```{r}
summary(fruits_vis_model_wet) 
summary(conspecifics_model_int_wet) 
summary(conspecifics_model_wet) 
summary(social_border_model_wet)
summary(frst_st_model_wet)
summary(social_border_model_int_wet)
summary(fruits_bot_model_int_wet)
summary(frst_st_model_int_wet)
summary(vedbaall_model_int_wet)
summary(vedba_model_int_wet)
```

From the outputs of the models, I will fit more complex models with covariates and interactions so I can choose best model again

```{r}
summary(full_wet_1 <- ssf_wet %>% 
  fit_clogit(case_ ~ n_vft_end + KJ_vft_end + tree_H_end + 
               n_lc_end + VeDBA_all_end + sl_ + log_sl_ + cos_ta_ + 
               n_vft_start:(sl_ + log_sl_ + cos_ta_) + 
               KJ_vft_start:(sl_ + log_sl_ + cos_ta_) +
               tree_H_start:(sl_ + log_sl_ + cos_ta_) +
               n_lc_start:(sl_ + log_sl_ + cos_ta_) +
               VeDBA_all_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), 
             model=TRUE))

summary(full_wet_2 <- ssf_wet %>% 
          fit_clogit(case_ ~ n_vft_end + tree_H_end +  
                       n_lc_end + VeDBA_all_end + sl_ + log_sl_ + cos_ta_ + 
                       n_vft_start:(sl_ + log_sl_ + cos_ta_) + 
                       tree_H_start:(sl_ + log_sl_ + cos_ta_) + 
                       n_lc_start:(sl_ + log_sl_ + cos_ta_) +
                       VeDBA_all_start:(sl_ + log_sl_ + cos_ta_) +
                       strata(step_id_), 
                     model=TRUE))

AIC(full_wet_1)
AIC(full_wet_2)
```

Model *full_wet_2* is simpler and statistically equal to *full_wet_1*. I will use *full_wet_2* for further analyses.

### 3.1.7 Check tentative vs updated movement parameters

#### 3.1.7.1 Step lengths

```{r}
updated_sl <- update_sl_distr(full_wet_2)
# data.frame for plotting
plot_sl <- data.frame(x = rep(NA, 100))

# x-axis is sequence of possible step lengths
plot_sl$x <- seq(from = 0, to = 400, length.out = 100)

# y-axis is the probability density under the given gamma distribution
# For the tentative distribution
plot_sl$tentative <- dgamma(
  x = plot_sl$x, 
  shape = full_wet_2$sl_$params$shape,
  scale = full_wet_2$sl_$params$scale)

# For the updated distribution
plot_sl$updated <- dgamma(
  x = plot_sl$x,
  shape = updated_sl$params$shape,
  scale = updated_sl$params$scale)

# Pivot from wide data to long data
plot_sl <- plot_sl %>% 
  pivot_longer(cols = -x)
```

In @fig-sl_wet we can see that there is not much variation between tentative and updated step lengths. This was expected since our model did not find significance for *sl\_*.

```{r}
#| label: fig-sl_wet 
#| fig-width: 5
#| fig-asp: 1 
#| fig-alt: |
#|   A plot showing tentative and updated step lengths
#| fig-cap: |
#|       Comparison between tentative and updated step lengths based on model full_wet_2
#| warning: FALSE  
ggplot(plot_sl, aes(x = x, y = value, color = factor(name))) +
  geom_line(linewidth = 1) +
  xlab("Step Length (m)") +
  ylab("Probability Density") +
  scale_color_manual(name = "Distribution", 
                     breaks = c("tentative", "updated"),
                     values = c("blue", "orange")) +
  theme_bw()
```

#### 3.1.7.2 Turn angles

```{r}
#| warning: FALSE
updated_ta <- update_ta_distr(full_wet_2)

# data.frame for plotting
plot_ta <- data.frame(x = rep(NA, 100))

# x-axis is sequence of possible step lengths
plot_ta$x <- seq(from = -1 * pi, to = pi, length.out = 100)

# y-axis is the probability density under the given von Mises distribution
# For the tentative distribution
plot_ta$tentative <- circular::dvonmises(
  x = plot_ta$x, 
  mu = full_wet_2$ta_$params$mu,
  kappa = full_wet_2$ta_$params$kappa)

# For the updated distribution
plot_ta$updated <- circular::dvonmises(
  x = plot_ta$x, 
  mu = updated_ta$params$mu,
  kappa = updated_ta$params$kappa)

# Pivot from wide data to long data
plot_ta <- plot_ta %>% 
  pivot_longer(cols = -x)
```

As in @fig-sl_wet, in @fig-ta_wet we can see that there is not much variation between tentative and updated turn angles. This was expected since our model did not find significance for *ta\_*.

```{r}
#| label: fig-ta_wet 
#| fig-width: 5
#| fig-asp: 1 
#| fig-alt: |
#|   A plot showing tentative and updated turn angles
#| fig-cap: |
#|       Comparison between tentative and updated turn angles based on model full_wet_2
#| warning: FALSE  
ggplot(plot_ta, aes(x = x, y = value, color = factor(name))) +
  geom_line(size = 1) +
  coord_cartesian(ylim = c(0, 0.25)) +
  xlab("Relative Turn Angle (radians)") +
  ylab("Probability Density") +
  scale_x_continuous(breaks = c(-pi, -pi/2, 0, pi/2, pi),
                     labels = c(expression(-pi, -pi/2, 0, pi/2, pi))) +
  scale_color_manual(name = "Distribution", 
                     breaks = c("tentative", "updated"),
                     values = c("blue", "orange")) +
theme_bw()
```

### 3.1.8 Calculate Relative selection strengths

To calculate the relative selection strengths for the covariates included in *full_wet_2*, I will create a series of data frames with medium, maximum and minimum values for each of the covariates in *full_wet_2*.

```{r}
# n_vft_end
s1 <- data.frame(
  n_vft_end = seq(from = -2, to = 2, length.out = 200),
  tree_H_end = 0,
  n_lc_end = 0,
  VeDBA_all_end = 0,
  n_vft_start = 0,
  tree_H_start = 0,
  n_lc_start = 0,
  VeDBA_all_start = 0,
  sl_ = median(ssf_wet$sl_),
  log_sl_ = log(median(ssf_wet$sl_)),
  cos_ta_ = median(ssf_wet$cos_ta_))

# data.frame for s2 -> all average
s2 <- data.frame(
  n_vft_end = 0,
  tree_H_end = 0,
  n_lc_end = 0,
  VeDBA_all_end = 0,
  n_vft_start = 0,
  tree_H_start = 0,
  n_lc_start = 0,
  VeDBA_all_start = 0,
  sl_ = median(ssf_wet$sl_),
  log_sl_ = log(median(ssf_wet$sl_)),
  cos_ta_ = median(ssf_wet$cos_ta_))

# Calculate log-RSS with Bootstrap Confidence Intervals
lr1 <- log_rss(full_wet_2, s1, s2, ci = "se", ci_level = 0.95)
foo <- lr1
n_vft_wet_df <- lr1$df

# tree_H 
s3 <- data.frame(
  n_vft_end = 0,
  tree_H_end = seq(from = -2, to = 2, length.out = 200),
  n_lc_end = 0,
  VeDBA_all_end = 0,
  n_vft_start = 0,
  tree_H_start = 0,
  n_lc_start = 0,
  VeDBA_all_start = 0,
  sl_ = median(ssf_wet$sl_),
  log_sl_ = log(median(ssf_wet$sl_)),
  cos_ta_ = median(ssf_wet$cos_ta_))

lr2 <- log_rss(full_wet_2, s3, s2, ci = "se", ci_level = 0.95)
tree_H_wet_df <- lr2$df

# n_lc
s4 <- data.frame(
  n_vft_end = 0,
  tree_H_end = 0,
  n_lc_end = seq(from = -2, to = 2, length.out = 200),
  VeDBA_all_end = 0,
  n_vft_start = 0,
  tree_H_start = 0,
  n_lc_start = 0,
  VeDBA_all_start = 0,
  sl_ = median(ssf_wet$sl_),
  log_sl_ = log(median(ssf_wet$sl_)),
  cos_ta_ = median(ssf_wet$cos_ta_))

lr3 <- log_rss(full_wet_2, s4, s2, ci = "se", ci_level = 0.95)
n_lc_wet_df <- lr3$df

### I will save the data frames with errors for the thesis
df1 <- n_vft_wet_df %>%
  transmute(x = n_vft_end_x1, log_rss, lwr, upr,
            variable = "#Known fruiting trees")

df2 <- tree_H_wet_df %>%
  transmute(x = tree_H_end_x1, log_rss, lwr, upr,
            variable = "Canopy height (m)")

df3 <- n_lc_wet_df %>%
  transmute(x = n_lc_end_x1, log_rss, lwr, upr,
            variable = "Log(#Long calls)")


# ---- Combine data ----
combined_df <- bind_rows(df1, df2, df3)

# ---- Colors ----
curve_colors <- c("#Known fruiting trees" = "burlywood",
                  "Canopy height (m)" = "chocolate4", 
                  "Log(#Long calls)" = "antiquewhite4")

```

Now, in @fig-habitat_wet we can see that there is a positive selection for all three significant covariates selected in model *full_wet_2*.

```{r}
#| label: fig-habitat_wet 
#| fig-width: 5
#| fig-asp: 1 
#| fig-alt: |
#|   A plot showing selection strengths for n_vft, tree_H and n_lc
#| fig-cap: |
#|       Selection strengths for significant covariates in model full_wet_2
#| warning: FALSE  
p1 <- ggplot(combined_df, aes(x = x, y = exp(log_rss), 
                              color = variable, fill = variable)) +
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray30") +
  scale_color_manual(values = curve_colors) +
  scale_fill_manual(values = curve_colors) +
  labs(x = "Predictor (SD)", y = "Log-RSS vs Mean predictor", color = "", fill = "") +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
p1
```

### 3.1.9 Interaction between movement and covariates

As in relative selection strengths, I will create a series of data frames with different values for each covariate.

#### 3.1.9.1 Number of visited feeding trees vs movement parameters

##### 3.1.9.1.1 *n_vft* vs *ta\_*

```{r}
#| warning: FALSE
low_ta_n_vft <- update_vonmises(
  dist = full_wet_2$ta_,
  beta_cos_ta = full_wet_2$model$coefficients["cos_ta_"] +
    -2 * full_wet_2$model$coefficients["cos_ta_:n_vft_start"])

# med turn-angle distribution
med_ta_n_vft <- update_vonmises(
  dist = full_wet_2$ta_,
  beta_cos_ta = full_wet_2$model$coefficients["cos_ta_"] +
    0 * full_wet_2$model$coefficients["cos_ta_:n_vft_start"])

# hi turn-angle distribution
hi_ta_n_vft <- update_vonmises(
  dist = full_wet_2$ta_,
  beta_cos_ta = full_wet_2$model$coefficients["cos_ta_"] +
    2 * full_wet_2$model$coefficients["cos_ta_:n_vft_start"])

#data.frame for plotting
plot_ta_n_vft <- data.frame(x = rep(NA, 100))

# x-axis is sequence of possible step lengths
plot_ta_n_vft$x <- seq(from = -pi, to = pi, length.out = 100)

# y-axis is the probability density under the given gamma distribution
# low
plot_ta_n_vft$low <- circular::dvonmises(x = plot_ta_n_vft$x, 
                                   kappa = low_ta_n_vft$params$kappa,
                                   mu = 0)
# med
plot_ta_n_vft$medium <- circular::dvonmises(x = plot_ta_n_vft$x, 
                                      kappa = med_ta_n_vft$params$kappa,
                                      mu = 0)
# hi
plot_ta_n_vft$high <- circular::dvonmises(x = plot_ta_n_vft$x, 
                                    kappa = hi_ta_n_vft$params$kappa,
                                    mu = 0)

# Pivot from wide to long data
plot_ta_n_vft <- plot_ta_n_vft %>% 
  pivot_longer(cols = -x)
```

Prepare the plot of *n_vft* vs *ta\_*

```{r}
p_ta_n_vft <- ggplot(plot_ta_n_vft, aes(x = x, y = value, color = factor(name))) +
  geom_line(size = 1) +
  scale_color_manual(name = "#Known fruiting trees",
                     breaks = c("low", "medium", "high"),
                     values = c("gray85", "gray50", "gray20")) +
  scale_x_continuous(breaks = c(-pi, -pi/2, 0, pi/2, pi),
                     labels = c(expression(-pi, -pi/2, 0, pi/2, pi))) +
  labs(x = "", y = "Probability Density", 
       title = "#Known fruiting trees") +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        plot.title = element_text(size = 20, hjust = .5),
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_blank(),
        axis.title.y = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
p_ta_n_vft
```

##### 3.1.9.1.2 *n_vft* vs *sl\_*

```{r}
# Low n_vft step-length distribution
low_sl_n_vft <- update_gamma(
  dist = full_wet_2$sl_,
  beta_sl = full_wet_2$model$coefficients["sl_"] +
    -10 * full_wet_2$model$coefficients["sl_:n_vft_start"],
  beta_log_sl = full_wet_2$model$coefficients["log_sl_"] +
    -10 * full_wet_2$model$coefficients["log_sl_:n_vft_start"])

# Medium silence step-length distribution
med_sl_n_vft <- update_gamma(
  dist = full_wet_2$sl_,
  beta_sl = full_wet_2$model$coefficients["sl_"] +
    0 * full_wet_2$model$coefficients["sl_:n_vft_start"],
  beta_log_sl = full_wet_2$model$coefficients["log_sl_"] +
    0 * full_wet_2$model$coefficients["log_sl_:n_vft_start"])

# High silence step-length distribution
hi_sl_n_vft <- update_gamma(
  dist = full_wet_2$sl_,
  beta_sl = full_wet_2$model$coefficients["sl_"] +
    10 * full_wet_2$model$coefficients["sl_:n_vft_start"],
  beta_log_sl = full_wet_2$model$coefficients["log_sl_"] +
    10 * full_wet_2$model$coefficients["log_sl_:n_vft_start"])

# Function to adjust parameters
adjust_params <- function(shape, scale) {
  if (shape <= 0) shape <- abs(shape) + 1e-5  # Make it slightly positive
  if (scale <= 0) scale <- abs(scale) + 1e-5  # Make it slightly positive
  return(list(shape = shape, scale = scale))
}

# Adjust parameters
low_sl_adj_n_vft <- adjust_params(low_sl_n_vft$params$shape, low_sl_n_vft$params$scale)
med_sl_adj_n_vft <- adjust_params(med_sl_n_vft$params$shape, med_sl_n_vft$params$scale)
hi_sl_adj_n_vft <- adjust_params(hi_sl_n_vft$params$shape, hi_sl_n_vft$params$scale)

# Create the data.frame for plotting
plot_sl_n_vft <- data.frame(
  x = seq(from = 0, to = 400, length.out = 100),
  low = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
               shape = low_sl_adj_n_vft$shape, scale = low_sl_adj_n_vft$scale),
  medium = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
                  shape = med_sl_adj_n_vft$shape, scale = med_sl_adj_n_vft$scale),
  high = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
                shape = hi_sl_adj_n_vft$shape, scale = hi_sl_adj_n_vft$scale))

# Verify the structure of the newly created data frame
str(plot_sl_n_vft)

# Pivot from wide to long data
plot_sl_long_n_vft <- plot_sl_n_vft %>%
  pivot_longer(cols = c(low, medium, high), names_to = "n_vft", 
               values_to = "Density")

# Verify the structure of the reshaped data
str(plot_sl_long_n_vft)
```

Prepare the plot of *n_vft* vs *sl\_*

```{r}
#tiff("n_vft_sl_WET_2025_05_d05.tiff", width = 11000, height = 6500, res = 1000)
p_sl_n_vft <- ggplot(plot_sl_long_n_vft, aes(x = x, y = Density, color = n_vft)) +
  geom_line(size = 1) +
  labs(x = "Step Length (m)", y = "Probability Density", 
       title = "#Known fruiting trees") +
  scale_color_manual(name = "",
                     breaks = c("low", "medium", "high"),
                     values = c("gray85", "gray50", "gray20")) +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        plot.title = element_text(size = 20, hjust = .5),
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_blank(),
        #axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_blank(),
        #axis.title.y = element_text(size = 20),
        axis.title.x = element_blank(),
        #axis.title.x = element_text(size = 20),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
p_sl_n_vft
#dev.off()
```

In @fig-n_vft_mov we can see that there is a small significant effect of number of visited fruit trees on the length of the steps, with steps longer (faster movement) where number of trees is higher. There is also an effect on turn angles, with movement more directed where there are fewer visited feeding trees.

```{r}
#| label: fig-n_vft_mov 
#| fig-width: 7
#| fig-asp: 6 
#| fig-alt: |
#|   A plot showing how number of visited fruit trees affected both step lengths and turn angles in the rainy season
#| fig-cap: |
#|       Interaction between number of visited fruit trees and both step lengths and turn angles in the rainy season 
#| warning: FALSE  
grid.arrange(p1, p_sl_n_vft, p_ta_n_vft, ncol=3)
```

#### 3.1.9.2 Canopy height vs movement parameters

##### 3.1.9.2.1 *VeDBA_all* vs *ta\_*

```{r}
#| warning: false
### VeDBA_all_start:
# Create again low, med and high turn-angle distribution objects
low_ta_VeDBA <- update_vonmises(
  dist = full_wet_2$ta_,
  beta_cos_ta = full_wet_2$model$coefficients["cos_ta_"] +
    -2 * full_wet_2$model$coefficients["cos_ta_:VeDBA_all_start"])

# med turn-angle distribution
med_ta_VeDBA <- update_vonmises(
  dist = full_wet_2$ta_,
  beta_cos_ta = full_wet_2$model$coefficients["cos_ta_"] +
    0 * full_wet_2$model$coefficients["cos_ta_:VeDBA_all_start"])

# hi turn-angle distribution
hi_ta_VeDBA <- update_vonmises(
  dist = full_wet_2$ta_,
  beta_cos_ta = full_wet_2$model$coefficients["cos_ta_"] +
    2 * full_wet_2$model$coefficients["cos_ta_:VeDBA_all_start"])

#data.frame for plotting
plot_ta_VeDBA <- data.frame(x = rep(NA, 100))

# x-axis is sequence of possible step lengths
plot_ta_VeDBA$x <- seq(from = -pi, to = pi, length.out = 100)

# y-axis is the probability density under the given gamma distribution
# low
plot_ta_VeDBA$low <- circular::dvonmises(x = plot_ta_VeDBA$x, 
                                   kappa = low_ta_VeDBA$params$kappa,
                                   mu = 0)
# med
plot_ta_VeDBA$medium <- circular::dvonmises(x = plot_ta_VeDBA$x, 
                                      kappa = med_ta_VeDBA$params$kappa,
                                      mu = 0)
# hi
plot_ta_VeDBA$high <- circular::dvonmises(x = plot_ta_VeDBA$x, 
                                    kappa = hi_ta_VeDBA$params$kappa,
                                    mu = 0)

# Pivot from wide to long data
plot_ta_VeDBA <- plot_ta_VeDBA %>% 
  pivot_longer(cols = -x)
```

Prepare the plot of *VeDBA_all* vs *ta\_*

```{r}
#tiff("VeDBA_all_ta_WET_2025_05_d05.tiff", width = 11000, height = 6500, res = 1000)
p_ta_VeDBA <- ggplot(plot_ta_VeDBA, aes(x = x, y = value, color = factor(name))) +
  geom_line(size = 1) +
  scale_color_manual(name = "VeDBA",
                     breaks = c("low", "medium", "high"),
                     values = c("gray85", "gray50", "gray20")) +
  scale_x_continuous(breaks = c(-pi, -pi/2, 0, pi/2, pi),
                     labels = c(expression(-pi, -pi/2, 0, pi/2, pi))) +
  labs(x = "Turn Angle (radians)", y = "Probability Density", 
       title = "sVeDBA") +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        plot.title = element_text(size = 20, hjust = .5),
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
#dev.off()
```

##### 3.1.9.2.2 *VeDBA_all* vs *sl\_*

```{r}
# Low alarm step-length distribution
low_sl_VeDBA <- update_gamma(
  dist = full_wet_2$sl_,
  beta_sl = full_wet_2$model$coefficients["sl_"] +
    -10 * full_wet_2$model$coefficients["sl_:VeDBA_all_start"],
  beta_log_sl = full_wet_2$model$coefficients["log_sl_"] +
    -10 * full_wet_2$model$coefficients["log_sl_:VeDBA_all_start"])

# Medium alarm step-length distribution
med_sl_VeDBA <- update_gamma(
  dist = full_wet_2$sl_,
  beta_sl = full_wet_2$model$coefficients["sl_"] +
    0 * full_wet_2$model$coefficients["sl_:VeDBA_all_start"],
  beta_log_sl = full_wet_2$model$coefficients["log_sl_"] +
    0 * full_wet_2$model$coefficients["log_sl_:VeDBA_all_start"])

# High alarm step-length distribution
hi_sl_VeDBA <- update_gamma(
  dist = full_wet_2$sl_,
  beta_sl = full_wet_2$model$coefficients["sl_"] +
    10 * full_wet_2$model$coefficients["sl_:VeDBA_all_start"],
  beta_log_sl = full_wet_2$model$coefficients["log_sl_"] +
    10 * full_wet_2$model$coefficients["log_sl_:VeDBA_all_start"])

# Function to adjust parameters
adjust_params_VeDBA <- function(shape, scale) {
  if (shape <= 0) shape <- abs(shape) + 1e-5  # Make it slightly positive
  if (scale <= 0) scale <- abs(scale) + 1e-5  # Make it slightly positive
  return(list(shape = shape, scale = scale))
}

# Adjust parameters
low_sl_adj_VeDBA <- adjust_params_VeDBA(low_sl_VeDBA$params$shape, low_sl_VeDBA$params$scale)
med_sl_adj_VeDBA <- adjust_params_VeDBA(med_sl_VeDBA$params$shape, med_sl_VeDBA$params$scale)
hi_sl_adj_VeDBA <- adjust_params_VeDBA(hi_sl_VeDBA$params$shape, hi_sl_VeDBA$params$scale)

# Create the data.frame for plotting
plot_sl_VeDBA <- data.frame(
  x = seq(from = 0, to = 400, length.out = 100),
  low = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
               shape = low_sl_adj_VeDBA$shape, scale = low_sl_adj_VeDBA$scale),
  medium = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
                  shape = med_sl_adj_VeDBA$shape, scale = med_sl_adj_VeDBA$scale),
  high = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
                shape = hi_sl_adj_VeDBA$shape, scale = hi_sl_adj_VeDBA$scale))

# Verify the structure of the newly created data frame
str(plot_sl_VeDBA)

# Pivot from wide to long data
plot_sl_long_VeDBA <- plot_sl_VeDBA %>%
  pivot_longer(cols = c(low, medium, high), names_to = "VeDBA_all", 
               values_to = "Density")

# Verify the structure of the reshaped data
str(plot_sl_long_VeDBA)
```

Prepare the plot of *VeDBA_all* vs *sl\_*

```{r}
#tiff("VeDBA_all_WET_2025_05_d05.tiff", width = 11000, height = 6500, res = 1000)
p_sl_VeDBA <- ggplot(plot_sl_long_VeDBA, aes(x = x, y = Density, color = VeDBA_all)) +
  geom_line(size = 1) +
  labs(x = "Step length (m)", y = "", 
       title = "sVeDBA") +
  scale_color_manual(name = "sVeDBA",
                     breaks = c("low", "medium", "high"),
                     values = c("gray85", "gray50", "gray20")) +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        plot.title = element_text(size = 20, hjust = .5),
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_blank(),
        axis.title.x = element_text(size = 20),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
#dev.off()
```

In @fig-VeDBA_mov_wet we can see that there is a small significant effect of VeDBA on the length of the steps, with steps longer (faster movement) where VeDBA is higher. There is also a marginaly significant effect on turn angles, with movement more directed where VeDBA is higher.

```{r}
#| label: fig-VeDBA_mov_wet 
#| fig-width: 7
#| fig-asp: 6 
#| fig-alt: |
#|   A plot showing how number of visited fruit trees affected both step lengths and turn angles in the rainy season
#| fig-cap: |
#|       Interaction between number of visited fruit trees and both step lengths and turn angles in the rainy season 
#| warning: FALSE  
grid.arrange(p_sl_VeDBA, p_ta_VeDBA, ncol=2)
```

## 3.2 Dry season - model fit

I will fit models with interactions with movement parameters (*sl\_*, *log_sl\_*, *ta\_* and *cos_ta\_*) and and without the interactions

### 3.2.1 Forest structure model

```{r}
frst_st_model_int_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ dist_f_b_end + nt_bot_end + tree_H_end + sl_ + log_sl_ + 
               cos_ta_ + 
               dist_f_b_start:(sl_ + log_sl_ + cos_ta_) +
               nt_bot_start:(sl_ + log_sl_ + cos_ta_) +
               tree_H_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

frst_st_model_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ dist_f_b_end + nt_bot_end + tree_H_end + sl_ + log_sl_ + 
               cos_ta_ + strata(step_id_), model=TRUE)
```

### 3.2.2 Fruits model

```{r}
fruits_bot_model_int_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ nft_bot_end + KJ_bot_end +
               sl_ + log_sl_ + cos_ta_ + 
               nft_bot_start:(sl_ + log_sl_ + cos_ta_) +
               KJ_bot_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

fruits_bot_model_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ nft_bot_end + KJ_bot_end +
               sl_ + log_sl_ + cos_ta_ + strata(step_id_), model=TRUE)

fruits_vis_model_int_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ n_vft_end + KJ_vft_end + 
               sl_ + log_sl_ + cos_ta_ + 
               n_vft_start:(sl_ + log_sl_ + cos_ta_) +
               KJ_vft_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

fruits_vis_model_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ n_vft_end + KJ_vft_end +
               sl_ + log_sl_ + cos_ta_ + strata(step_id_), model=TRUE)
```

### 3.2.3 Invertebrates model

```{r}
insect_model_int_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ inv_bio_end + sl_ + log_sl_ + cos_ta_ + 
               inv_bio_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

insect_model_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ inv_bio_end + sl_ + log_sl_ + cos_ta_ + 
               strata(step_id_), model=TRUE)
```

### 3.2.4 Fear model

```{r}
fear_model_int_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ n_alarm_end + sl_ + log_sl_ + cos_ta_ + 
               n_alarm_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

fear_model_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ n_alarm_end + sl_ + log_sl_ + cos_ta_ + 
               strata(step_id_), 
             model=TRUE)
```

### 3.2.5 VeDBA model

```{r}
vedba_model_int_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ VeDBA_end + sl_ + log_sl_ + cos_ta_ + 
               VeDBA_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

vedba_model_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ VeDBA_end + sl_ + log_sl_ + cos_ta_ + strata(step_id_), 
             model=TRUE)
```

### 3.2.6 Social model

```{r}
conspecifics_model_int_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ UDcons_end + n_lc_end + sl_ + log_sl_ + cos_ta_ + 
               UDcons_start:(sl_ + log_sl_ + cos_ta_) +
               n_lc_start:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

conspecifics_model_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ UDcons_end + n_lc_end + sl_ + log_sl_ + cos_ta_ +
               strata(step_id_), model=TRUE)

social_border_model_int_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ dist_f_b_end + n_lc_end + sl_ + log_sl_ + cos_ta_ + 
               dist_f_b_start:(sl_ + log_sl_ + cos_ta_) +
               n_lc_end:(sl_ + log_sl_ + cos_ta_) +
               strata(step_id_), model=TRUE)

social_border_model_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ dist_f_b_end + n_lc_end + sl_ + log_sl_ + cos_ta_ +
               strata(step_id_), model=TRUE)
```

### 3.2.7 Null model

```{r}
null_model_dry <- ssf_dry %>% 
  fit_clogit(case_ ~ 1 + sl_ + log_sl_ + cos_ta_ + strata(step_id_), 
             model=TRUE)
```

### 3.2.8 Compare models

```{r}
conspecifics_model_int_dry_aic <- AIC(conspecifics_model_int_dry)
conspecifics_model_dry_aic <- AIC(conspecifics_model_dry)
fruits_bot_model_int_dry_aic <- AIC(fruits_bot_model_int_dry)
fruits_bot_model_dry_aic <- AIC(fruits_bot_model_dry)
fruits_vis_model_int_dry_aic <- AIC(fruits_vis_model_int_dry)
fruits_vis_model_dry_aic <- AIC(fruits_vis_model_dry)
insect_model_int_dry_aic <- AIC(insect_model_int_dry)
insect_model_dry_aic <- AIC(insect_model_dry)
frst_st_model_int_dry_aic <- AIC(frst_st_model_int_dry)
frst_st_model_dry_aic <- AIC(frst_st_model_dry)
vedba_model_int_dry_aic <- AIC(vedba_model_int_dry)
vedba_model_dry_aic <- AIC(vedba_model_dry)
fear_model_int_dry_aic <- AIC(fear_model_int_dry)
fear_model_dry_aic <- AIC(fear_model_dry)
social_border_model_int_dry_aic <- AIC(social_border_model_int_dry)
social_border_model_dry_aic <- AIC(social_border_model_dry)
null_model_dry_aic <- AIC(null_model_dry)

# Compare AIC values
aic_table_dry <- data.frame(Model = c("conspecifics_model_int_dry", 
                                      "conspecifics_model_dry", 
                                      "fruits_bot_model_int_dry", 
                                      "fruits_bot_model_dry",
                                      "fruits_vis_model_int_dry", 
                                      "fruits_vis_model_dry",
                                      "insect_model_int_dry",
                                      "insect_model_dry",
                                      "frst_st_model_int_dry",
                                      "frst_st_model_dry", 
                                      "vedba_model_int_dry",
                                      "vedba_model_dry",
                                      "fear_model_int_dry",
                                      "fear_model_dry",
                                      "social_border_model_int_dry",
                                      "social_border_model_dry",
                                      "null_model_dry"),
                            AIC_Value = c(conspecifics_model_int_dry_aic, 
                                          conspecifics_model_dry_aic,
                                          fruits_bot_model_int_dry_aic,
                                          fruits_bot_model_dry_aic,
                                          fruits_vis_model_int_dry_aic,
                                          fruits_vis_model_dry_aic, 
                                          insect_model_int_dry_aic,
                                          insect_model_dry_aic,
                                          frst_st_model_int_dry_aic,
                                          frst_st_model_dry_aic, 
                                          vedba_model_int_dry_aic,
                                          vedba_model_dry_aic,
                                          fear_model_int_dry_aic,
                                          fear_model_dry_aic,
                                          social_border_model_int_dry_aic,
                                          social_border_model_dry_aic,
                                          null_model_dry_aic))


# Find the index of the best model (i.e., the model with the lowest AIC value)
best_model_index_dry <- which.min(aic_table_dry$AIC_Value)
# Subtract lowest AIC_value from all the rest to create a delta_AIC column
aic_table_dry$delta_AIC<-aic_table_dry$AIC_Value-min(aic_table_dry$AIC_Value)
# Sort the AIC table by AIC values in increasing order
aic_table_sorted_dry <- aic_table_dry[order(aic_table_dry$AIC_Value), ]
print(aic_table_sorted_dry)
```

I can see that *vedba_model_int_dry*, *insect_model_int_dry*, *vedba_model_dry*, *social_border_model_int_dry*, *insect_model_dry*, *fruits_bot_model_dry*, are worse or equal to null_model.

Best model is *fear_model_int_dry*:

```{r}
summary(fear_model_int_dry)
```

Inspect other models better than null:

```{r}
summary(fruits_vis_model_int_dry)
summary(fruits_vis_model_dry) 
summary(fear_model_dry)
summary(frst_st_model_int_dry) 
summary(frst_st_model_dry) 
summary(conspecifics_model_int_dry)
summary(fruits_bot_model_int_dry)
summary(conspecifics_model_dry)
summary(social_border_model_dry)
```

From the outputs of the models, I will fit more complex models with covariates and interactions so I can choose best model again

```{r}
summary(full_dry_1 <- ssf_dry %>% 
          fit_clogit(case_ ~ n_vft_end + KJ_bot_end + tree_H_end + 
                       n_alarm_end+ n_lc_end + sl_ + log_sl_ + cos_ta_ +
                       n_vft_start:(sl_ + log_sl_ + cos_ta_) + 
                       tree_H_start:(sl_ + log_sl_ + cos_ta_) +
                       n_alarm_start:(sl_ + log_sl_ + cos_ta_) +
                       n_lc_start:(sl_ + log_sl_ + cos_ta_) +
                       KJ_bot_start:(sl_ + log_sl_ + cos_ta_) +
                       strata(step_id_), 
                     model=TRUE))

summary(full_dry_2 <- ssf_dry %>% 
          fit_clogit(case_ ~ n_vft_end + tree_H_end + 
                       n_alarm_end + n_lc_end + sl_ + log_sl_ + cos_ta_ +
                       n_vft_start:(sl_ + log_sl_ + cos_ta_) + 
                       tree_H_start:(sl_ + log_sl_ + cos_ta_) +
                       n_alarm_start:(sl_ + log_sl_ + cos_ta_) +
                       n_lc_start:(sl_ + log_sl_ + cos_ta_) +
                       strata(step_id_), 
                     model=TRUE))

summary(full_dry_3 <- ssf_dry %>% 
          fit_clogit(case_ ~ n_vft_end + tree_H_end + 
                       n_alarm_end + sl_ + log_sl_ + cos_ta_ +
                       n_vft_start:(sl_ + log_sl_ + cos_ta_) + 
                       tree_H_start:(sl_ + log_sl_ + cos_ta_) +
                       n_alarm_start:(sl_ + log_sl_ + cos_ta_) +
                       strata(step_id_), 
                     model=TRUE))

AIC(full_dry_1)
AIC(full_dry_2)
AIC(full_dry_3)
```

Model *full_dry_1* is the best model. I will use *full_dry_1* for further analyses.

### 3.2.9 Check tentative vs updated movement parameters

#### 3.2.9.1 Step lengths

```{r}
updated_sl <- update_sl_distr(full_dry_1)
# data.frame for plotting
plot_sl <- data.frame(x = rep(NA, 100))

# x-axis is sequence of possible step lengths
plot_sl$x <- seq(from = 0, to = 400, length.out = 100)

# y-axis is the probability density under the given gamma distribution
# For the tentative distribution
plot_sl$tentative <- dgamma(
  x = plot_sl$x, 
  shape = full_dry_1$sl_$params$shape,
  scale = full_dry_1$sl_$params$scale)

# For the updated distribution
plot_sl$updated <- dgamma(
  x = plot_sl$x,
  shape = updated_sl$params$shape,
  scale = updated_sl$params$scale)

# Pivot from wide data to long data
plot_sl <- plot_sl %>% 
  pivot_longer(cols = -x)
```

In @fig-sl_dry we can see that there is not much variation between tentative and updated step lengths. This was expected since our model did not find significance for *sl\_*.

```{r}
#| label: fig-sl_dry
#| fig-width: 5
#| fig-asp: 1 
#| fig-alt: |
#|   A plot showing tentative and updated step lengths
#| fig-cap: |
#|       Comparison between tentative and updated step lengths based on model full_dry_1
#| warning: FALSE  
ggplot(plot_sl, aes(x = x, y = value, color = factor(name))) +
  geom_line(linewidth = 1) +
  xlab("Step Length (m)") +
  ylab("Probability Density") +
  scale_color_manual(name = "Distribution", 
                     breaks = c("tentative", "updated"),
                     values = c("blue", "orange")) +
  theme_bw()
```

#### 3.2.9.2 Turn angles

```{r}
#| warning: FALSE
updated_ta <- update_ta_distr(full_dry_1)

# data.frame for plotting
plot_ta <- data.frame(x = rep(NA, 100))

# x-axis is sequence of possible step lengths
plot_ta$x <- seq(from = -1 * pi, to = pi, length.out = 100)

# y-axis is the probability density under the given von Mises distribution
# For the tentative distribution
plot_ta$tentative <- circular::dvonmises(
  x = plot_ta$x, 
  mu = full_dry_1$ta_$params$mu,
  kappa = full_dry_1$ta_$params$kappa)

# For the updated distribution
plot_ta$updated <- circular::dvonmises(
  x = plot_ta$x, 
  mu = updated_ta$params$mu,
  kappa = updated_ta$params$kappa)

# Pivot from wide data to long data
plot_ta <- plot_ta %>% 
  pivot_longer(cols = -x)
```

As in @fig-sl_dry, in @fig-ta_dry we can see that there is not much variation between tentative and updated turn angles. This was expected since our model did not find significance for *ta\_*.

```{r}
#| label: fig-ta_dry 
#| fig-width: 5
#| fig-asp: 1 
#| fig-alt: |
#|   A plot showing tentative and updated turn angles
#| fig-cap: |
#|       Comparison between tentative and updated turn angles based on model full_dry_1
#| warning: FALSE  
ggplot(plot_ta, aes(x = x, y = value, color = factor(name))) +
  geom_line(size = 1) +
  coord_cartesian(ylim = c(0, 0.4)) +  # ← increased upper limit
  xlab("Relative Turn Angle (radians)") +
  ylab("Probability Density") +
  scale_x_continuous(breaks = c(-pi, -pi/2, 0, pi/2, pi),
                     labels = c(expression(-pi),expression(-pi/2),
                                expression(0), 
                                expression(pi/2), 
                                expression(pi))) +
  scale_color_manual(
    name = "Distribution", 
    breaks = c("tentative", "updated"),
    values = c("blue", "orange")) +
  theme_bw()
```

### 3.2.9 Calculate Relative selection strengths

To calculate the relative selection strengths for the covariates included in *full_wet_2*, I will create a series of data frames with medium, maximum and minimum values for each of the covariates in *full_wet_2*.

```{r}
# Effect of just n_vft
s1 <- data.frame(
  n_vft_end = seq(from = -2, to = 2, length.out = 200),
  tree_H_end = 0,
  n_alarm_end = 0,
  n_lc_end = 0,
  n_vft_start = 0,
  tree_H_start = 0,
  n_alarm_start = 0,
  n_lc_start = 0,
  sl_ = median(ssf_dry$sl_),
  log_sl_ = log(median(ssf_dry$sl_)),
  cos_ta_ = median(ssf_dry$cos_ta_))

# data.frame for s5 -> all average
s2 <- data.frame(
  n_vft_end = 0,
  tree_H_end = 0,
  n_alarm_end = 0,
  n_lc_end = 0,
  n_vft_start = 0,
  tree_H_start = 0,
  n_alarm_start = 0,
  n_lc_start = 0,
  sl_ = median(ssf_dry$sl_),
  log_sl_ = log(median(ssf_dry$sl_)),
  cos_ta_ = median(ssf_dry$cos_ta_))

# Calculate log-RSS with Bootstrap Confidence Intervals
lr1 <- log_rss(full_dry_2, s1, s2, ci = "se", ci_level = 0.95)
foo <- lr1
n_vft_dry_df <- lr1$df

# Tree Height
s3 <- data.frame(
  n_vft_end = 0,
  tree_H_end = seq(from = -2, to = 2, length.out = 200),
  n_alarm_end = 0,
  n_lc_end = 0,
  n_vft_start = 0,
  tree_H_start = 0,
  n_alarm_start = 0,
  n_lc_start = 0,
  sl_ = median(ssf_dry$sl_),
  log_sl_ = log(median(ssf_dry$sl_)),
  cos_ta_ = median(ssf_dry$cos_ta_))

lr2 <- log_rss(full_dry_2, s3, s2, ci = "se", ci_level = 0.95)
tree_H_dry_df <- lr2$df


# Alarm calls
s4 <- data.frame(
  n_vft_end = 0,
  tree_H_end = 0,
  n_alarm_end = seq(from = -2, to = 2, length.out = 200),
  n_lc_end = 0,
  n_vft_start = 0,
  tree_H_start = 0,
  n_alarm_start = 0,
  n_lc_start = 0,
  sl_ = median(ssf_dry$sl_),
  log_sl_ = log(median(ssf_dry$sl_)),
  cos_ta_ = median(ssf_dry$cos_ta_))

lr3 <- log_rss(full_dry_2, s4, s2, ci = "se", ci_level = 0.95)
alarm_dry_df <- lr3$df

df1_dry <- n_vft_dry_df %>%
  transmute(x = n_vft_end_x1, log_rss, lwr, upr, variable = "#Known fruiting trees")

df2_dry <- tree_H_dry_df %>%
  transmute(x = tree_H_end_x1, log_rss, lwr, upr, variable = "Canopy height (m)")

combined_df_dry <- bind_rows(df1_dry, df2_dry)
```

Now, in @fig-habitat_dry we can see that there is a positive selection for number of visited fruit feeding trees and canopy height, the significant covariates selected in model *full_dry_1*.

```{r}
#| label: fig-habitat_dry 
#| fig-width: 5
#| fig-asp: 1 
#| fig-alt: |
#|   A plot showing selection strengths for n_vft and tree_H
#| fig-cap: |
#|       Selection strengths for significant covariates in model full_dry_1
#| warning: FALSE  
p2 <- ggplot(combined_df_dry, aes(x = x, y = exp(log_rss), 
                                  color = variable, fill = variable)) +
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray30") +
  scale_color_manual(values = curve_colors) +
  scale_fill_manual(values = curve_colors) +
  labs(x = "Predictor (SD)", y = "Log-RSS vs Mean predictor", color = "", fill = "") +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
p2
```

## 3.2.10 Interaction between movement and covariates

As in relative selection strengths, I will create a series of data frames with different values for each covariate.

### 3.2.10.1 Number of alarm calls vs movement parameters

#### 3.2.10.1.1 *n_alarm* vs *ta\_*

```{r}
#| warning: false
# Create again low, med and high turn-angle distribution objects
low_ta_n_alarm_d <- update_vonmises(
  dist = full_dry_2$ta_,
  beta_cos_ta = full_dry_2$model$coefficients["cos_ta_"] +
    -2 * full_dry_2$model$coefficients["cos_ta_:n_alarm_start"])

# med turn-angle distribution
med_ta_n_alarm_d <- update_vonmises(
  dist = full_dry_2$ta_,
  beta_cos_ta = full_dry_2$model$coefficients["cos_ta_"] +
    0 * full_dry_2$model$coefficients["cos_ta_:n_alarm_start"])

# hi turn-angle distribution
hi_ta_n_alarm_d <- update_vonmises(
  dist = full_dry_2$ta_,
  beta_cos_ta = full_dry_2$model$coefficients["cos_ta_"] +
    2 * full_dry_2$model$coefficients["cos_ta_:n_alarm_start"])

#data.frame for plotting
plot_ta_n_alarm_d <- data.frame(x = rep(NA, 100))

# x-axis is sequence of possible step lengths
plot_ta_n_alarm_d$x <- seq(from = -pi, to = pi, length.out = 100)

# y-axis is the probability density under the given gamma distribution
# low
plot_ta_n_alarm_d$low <- circular::dvonmises(x = plot_ta_n_alarm_d$x, 
                                   kappa = low_ta_n_alarm_d$params$kappa,
                                   mu = 0)
# med
plot_ta_n_alarm_d$medium <- circular::dvonmises(x = plot_ta_n_alarm_d$x, 
                                      kappa = med_ta_n_alarm_d$params$kappa,
                                      mu = 0)
# hi
plot_ta_n_alarm_d$high <- circular::dvonmises(x = plot_ta_n_alarm_d$x, 
                                    kappa = hi_ta_n_alarm_d$params$kappa,
                                    mu = 0)

# Pivot from wide to long data
plot_ta_n_alarm_d <- plot_ta_n_alarm_d %>% 
  pivot_longer(cols = -x)
```

Prepare the plot of *n_alarm* vs *ta\_*

```{r}
#tiff("alarm_ta_dry_2025_05_d05.tiff", width = 11000, height = 6500, res = 1000)
p_ta_n_alarm_dry <- ggplot(plot_ta_n_alarm_d, aes(x = x, y = value, color = factor(name))) +
  geom_line(size = 1) +
  scale_color_manual(name = "#Alarm calls",
                     breaks = c("low", "medium", "high"),
                     values = c("gray85", "gray50", "gray20")) +
  scale_x_continuous(breaks = c(-pi, -pi/2, 0, pi/2, pi),
                     labels = c(expression(-pi, -pi/2, 0, pi/2, pi))) +
  labs(x = "Turn Angle (radians)", y = "Probability Density", 
       title = "#Alarm calls") +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        plot.title = element_text(size = 20, hjust = .5),
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        legend.position = "none",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
#dev.off()
```

#### 3.2.10.1.2 *n_alarm* vs *sl\_*

```{r}
# Low alarm step-length distribution
low_sl_n_alarm_d <- update_gamma(
  dist = full_dry_2$sl_,
  beta_sl = full_dry_2$model$coefficients["sl_"] +
    -2 * full_dry_2$model$coefficients["sl_:n_alarm_start"],
  beta_log_sl = full_dry_2$model$coefficients["log_sl_"] +
    -2 * full_dry_2$model$coefficients["log_sl_:n_alarm_start"])

# Medium alarm step-length distribution
med_sl_n_alarm_d <- update_gamma(
  dist = full_dry_2$sl_,
  beta_sl = full_dry_2$model$coefficients["sl_"] +
    0 * full_dry_2$model$coefficients["sl_:n_alarm_start"],
  beta_log_sl = full_dry_2$model$coefficients["log_sl_"] +
    0 * full_dry_2$model$coefficients["log_sl_:n_alarm_start"])

# High alarm step-length distribution
hi_sl_n_alarm_d <- update_gamma(
  dist = full_dry_2$sl_,
  beta_sl = full_dry_2$model$coefficients["sl_"] +
    2 * full_dry_2$model$coefficients["sl_:n_alarm_start"],
  beta_log_sl = full_dry_2$model$coefficients["log_sl_"] +
    2 * full_dry_2$model$coefficients["log_sl_:n_alarm_start"])

# Function to adjust parameters
adjust_params_n_alarm_d <- function(shape, scale) {
  if (shape <= 0) shape <- abs(shape) + 1e-5  # Make it slightly positive
  if (scale <= 0) scale <- abs(scale) + 1e-5  # Make it slightly positive
  return(list(shape = shape, scale = scale))
}

# Adjust parameters
low_sl_adj_n_alarm_d <- adjust_params(low_sl_n_alarm_d$params$shape, low_sl_n_alarm_d$params$scale)
med_sl_adj_n_alarm_d <- adjust_params(med_sl_n_alarm_d$params$shape, med_sl_n_alarm_d$params$scale)
hi_sl_adj_n_alarm_d <- adjust_params(hi_sl_n_alarm_d$params$shape, hi_sl_n_alarm_d$params$scale)

# Create the data.frame for plotting
plot_sl_n_alarm_d <- data.frame(
  x = seq(from = 0, to = 400, length.out = 100),
  low = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
               shape = low_sl_adj_n_alarm_d$shape, scale = low_sl_adj_n_alarm_d$scale),
  medium = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
                  shape = med_sl_adj_n_alarm_d$shape, scale = med_sl_adj_n_alarm_d$scale),
  high = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
                shape = hi_sl_adj_n_alarm_d$shape, scale = hi_sl_adj_n_alarm_d$scale))

# Verify the structure of the newly created data frame
str(plot_sl_n_alarm_d)

# Pivot from wide to long data
plot_sl_n_alarm_d <- plot_sl_n_alarm_d %>%
  pivot_longer(cols = c(low, medium, high), names_to = "n_alarm", 
               values_to = "Density")

# Verify the structure of the reshaped data
str(plot_sl_n_alarm_d)
```

Prepare the plot of *n_alarm* vs *sl\_*

```{r}
#tiff("alarm_sl_dry_2025_05_d05.tiff", width = 11000, height = 6500, res = 1000)
p_sl_n_alarm_dry <- ggplot(plot_sl_n_alarm_d, aes(x = x, y = Density, color = n_alarm)) +
  geom_line(size = 1) +
  labs(x = "Step length (m)", y = "", 
       title = "#Alarm calls") +
  scale_color_manual(name = "#Alarms",
                     breaks = c("low", "medium", "high"),
                     values = c("gray85", "gray50", "gray20")) +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        plot.title = element_text(size = 20, hjust = .5),
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_blank(),
        axis.title.x = element_text(size = 20),
        legend.position = "none",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
#dev.off()
```

In @fig-alarm_mov_dry we can see that there is a marginally significant effect of the number of alarm calls on the length of the steps, with steps longer (faster movement) where the number of alarm calls is higher. There is also a significant effect on turn angles, with movement more directed where the number of alarm calls is lower.

```{r}
#| label: fig-alarm_mov_dry
#| fig-width: 7
#| fig-asp: 6 
#| fig-alt: |
#|   A plot showing how number of alarm calls affected both step lengths and turn angles in the rainy season
#| fig-cap: |
#|       Interaction between number of alarm calls and both step lengths and turn angles in the rainy season 
#| warning: FALSE  
grid.arrange(p_sl_n_alarm_dry, p_ta_n_alarm_dry, ncol=2)
```

### 3.2.10.2 Number of inter group long calls vs movement parameters

#### 3.2.10.2.1 *n_lc* vs *ta\_*

```{r}
#| warning: false
# Create again low, med and high turn-angle distribution objects
low_ta_n_lc_d <- update_vonmises(
  dist = full_dry_2$ta_,
  beta_cos_ta = full_dry_2$model$coefficients["cos_ta_"] +
    -2 * full_dry_2$model$coefficients["cos_ta_:n_lc_start"])

# med turn-angle distribution
med_ta_n_lc_d <- update_vonmises(
  dist = full_dry_2$ta_,
  beta_cos_ta = full_dry_2$model$coefficients["cos_ta_"] +
    0 * full_dry_2$model$coefficients["cos_ta_:n_lc_start"])

# hi turn-angle distribution
hi_ta_n_lc_d <- update_vonmises(
  dist = full_dry_2$ta_,
  beta_cos_ta = full_dry_2$model$coefficients["cos_ta_"] +
    2 * full_dry_2$model$coefficients["cos_ta_:n_lc_start"])

#data.frame for plotting
plot_ta_n_lc_d <- data.frame(x = rep(NA, 100))

# x-axis is sequence of possible step lengths
plot_ta_n_lc_d$x <- seq(from = -pi, to = pi, length.out = 100)

# y-axis is the probability density under the given gamma distribution
# low
plot_ta_n_lc_d$low <- circular::dvonmises(x = plot_ta_n_lc_d$x, 
                                   kappa = low_ta_n_lc_d$params$kappa,
                                   mu = 0)
# med
plot_ta_n_lc_d$medium <- circular::dvonmises(x = plot_ta_n_lc_d$x, 
                                      kappa = med_ta_n_lc_d$params$kappa,
                                      mu = 0)
# hi
plot_ta_n_lc_d$high <- circular::dvonmises(x = plot_ta_n_lc_d$x, 
                                    kappa = hi_ta_n_lc_d$params$kappa,
                                    mu = 0)

# Pivot from wide to long data
plot_ta_n_lc_d <- plot_ta_n_lc_d %>% 
  pivot_longer(cols = -x)
```

Prepare the plot of *n_lc* vs *ta\_*

```{r}
#tiff("n_lc_ta_dry_2025_05_d05.tiff", width = 11000, height = 6500, res = 1000)
p_ta_n_lc_dry <- ggplot(plot_ta_n_lc_d, aes(x = x, y = value, color = factor(name))) +
  geom_line(size = 1) +
  scale_color_manual(name = "Log(#Long calls)",
                     breaks = c("low", "medium", "high"),
                     values = c("gray85", "gray50", "gray20")) +
  scale_x_continuous(breaks = c(-pi, -pi/2, 0, pi/2, pi),
                     labels = c(expression(-pi, -pi/2, 0, pi/2, pi))) +
  labs(x = "Turn Angle (radians)", y = "Probability Density", 
       title = "Log(#Long calls)") +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        plot.title = element_text(size = 20, hjust = .5),
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
#dev.off()
```

#### 3.2.10.2.2 *n_lc* vs *sl\_*

```{r}
# Low alarm step-length distribution
low_sl_n_lc_d <- update_gamma(
  dist = full_dry_2$sl_,
  beta_sl = full_dry_2$model$coefficients["sl_"] +
    -2 * full_dry_2$model$coefficients["sl_:n_lc_start"],
  beta_log_sl = full_dry_2$model$coefficients["log_sl_"] +
    -2 * full_dry_2$model$coefficients["log_sl_:n_lc_start"])

# Medium alarm step-length distribution
med_sl_n_lc_d <- update_gamma(
  dist = full_dry_2$sl_,
  beta_sl = full_dry_2$model$coefficients["sl_"] +
    0 * full_dry_2$model$coefficients["sl_:n_lc_start"],
  beta_log_sl = full_dry_2$model$coefficients["log_sl_"] +
    0 * full_dry_2$model$coefficients["log_sl_:n_lc_start"])

# High alarm step-length distribution
hi_sl_n_lc_d <- update_gamma(
  dist = full_dry_2$sl_,
  beta_sl = full_dry_2$model$coefficients["sl_"] +
    2 * full_dry_2$model$coefficients["sl_:n_lc_start"],
  beta_log_sl = full_dry_2$model$coefficients["log_sl_"] +
    2 * full_dry_2$model$coefficients["log_sl_:n_lc_start"])

# Function to adjust parameters
adjust_params_n_lc_d <- function(shape, scale) {
  if (shape <= 0) shape <- abs(shape) + 1e-5  # Make it slightly positive
  if (scale <= 0) scale <- abs(scale) + 1e-5  # Make it slightly positive
  return(list(shape = shape, scale = scale))
}

# Adjust parameters
low_sl_adj_n_lc_d <- adjust_params(low_sl_n_lc_d$params$shape, low_sl_n_lc_d$params$scale)
med_sl_adj_n_lc_d <- adjust_params(med_sl_n_lc_d$params$shape, med_sl_n_lc_d$params$scale)
hi_sl_adj_n_lc_d <- adjust_params(hi_sl_n_lc_d$params$shape, hi_sl_n_lc_d$params$scale)

# Create the data.frame for plotting
plot_sl_n_lc_d <- data.frame(
  x = seq(from = 0, to = 400, length.out = 100),
  low = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
               shape = low_sl_adj_n_lc_d$shape, scale = low_sl_adj_n_lc_d$scale),
  medium = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
                  shape = med_sl_adj_n_lc_d$shape, scale = med_sl_adj_n_lc_d$scale),
  high = dgamma(x = seq(from = 0, to = 400, length.out = 100), 
                shape = hi_sl_adj_n_lc_d$shape, scale = hi_sl_adj_n_lc_d$scale))

# Verify the structure of the newly created data frame
str(plot_sl_n_lc_d)

# Pivot from wide to long data
plot_sl_n_lc_d <- plot_sl_n_lc_d %>%
  pivot_longer(cols = c(low, medium, high), names_to = "n_lc", 
               values_to = "Density")

# Verify the structure of the reshaped data
str(plot_sl_n_lc_d)
```

Prepare the plot of *n_lc* vs *sl\_*

```{r}
#tiff("n_lc_sl_dry_2025_05_d05.tiff", width = 11000, height = 6500, res = 1000)
p_sl_n_lc_dry <- ggplot(plot_sl_n_lc_d, aes(x = x, y = Density, color = n_lc)) +
  geom_line(size = 1) +
  labs(x = "Step length (m)", y = "", 
       title = "Log(#Long calls)") +
  scale_color_manual(name = "Log(#Long calls)",
                     breaks = c("low", "medium", "high"),
                     values = c("gray85", "gray50", "gray20")) +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        plot.title = element_text(size = 20, hjust = .5),
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_blank(),
        axis.title.x = element_text(size = 20),
        legend.position = "none",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
#dev.off()
```

In @fig-lc_mov_dry we can see that there is a marginally significant effect of the number of inter group long calls on the length of the steps, with longer steps (faster movement) where the number of inter group long calls is higher. There is also a significant effect on turn angles, with movement more directed where the number of long calls is lower.

```{r}
#| label: fig-lc_mov_dry
#| fig-width: 7
#| fig-asp: 6 
#| fig-alt: |
#|   A plot showing how number of inter group long calls affected both step lengths and turn angles in the rainy season
#| fig-cap: |
#|       Interaction between number of inter group long calls and both step lengths and turn angles in the rainy season 
#| warning: FALSE  
grid.arrange(p_sl_n_lc_dry, p_ta_n_lc_dry, ncol=2)
```

### 3.2.10.3 Fruit energy content vs movement parameters

#### 3.2.10.3.1 *KJ_bot* vs *ta\_*

```{r}
#| warning: false
## KJ_bot_start
# Create again low, med and high turn-angle distribution objects
low_ta_KJ_bot_d <- update_vonmises(
  dist = full_dry_1$ta_,
  beta_cos_ta = full_dry_1$model$coefficients["cos_ta_"] +
    -2 * full_dry_1$model$coefficients["cos_ta_:KJ_bot_start"])
# med turn-angle distribution
med_ta_KJ_bot_d <- update_vonmises(
  dist = full_dry_1$ta_,
  beta_cos_ta = full_dry_1$model$coefficients["cos_ta_"] +
    0 * full_dry_1$model$coefficients["cos_ta_:KJ_bot_start"])
# hi turn-angle distribution
hi_ta_KJ_bot_d <- update_vonmises(
  dist = full_dry_1$ta_,
  beta_cos_ta = full_dry_1$model$coefficients["cos_ta_"] +
    2 * full_dry_1$model$coefficients["cos_ta_:KJ_bot_start"])
#data.frame for plotting
plot_ta_KJ_bot_d <- data.frame(x = rep(NA, 100))
# x-axis is sequence of possible step lengths
plot_ta_KJ_bot_d$x <- seq(from = -pi, to = pi, length.out = 100)
# y-axis is the probability density under the given gamma distribution
# low
plot_ta_KJ_bot_d$low <- circular::dvonmises(x = plot_ta_KJ_bot_d$x,
                                   kappa = low_ta_KJ_bot_d$params$kappa,
                                   mu = 0)
# med
plot_ta_KJ_bot_d$medium <- circular::dvonmises(x = plot_ta_KJ_bot_d$x,
                                      kappa = med_ta_KJ_bot_d$params$kappa,
                                      mu = 0)

# hi
plot_ta_KJ_bot_d$high <- circular::dvonmises(x = plot_ta_KJ_bot_d$x,
                                    kappa = hi_ta_KJ_bot_d$params$kappa,
                                    mu = 0)
# Pivot from wide to long data
plot_ta_KJ_bot_d <- plot_ta_KJ_bot_d %>%
  pivot_longer(cols = -x)
```

Prepare the plot of *KJ_bot* vs *ta\_*

```{r}
p_ta_KJ_bot_dry <- ggplot(plot_ta_KJ_bot_d, aes(x = x, y = value,
                                          color = factor(name))) +
  geom_line(linewidth = 1) +
  scale_color_manual(name = "Fruit energy",
                     breaks = c("low", "medium", "high"),
                     values = c("gray85", "gray50", "gray20")) +
  scale_x_continuous(breaks = c(-pi, -pi/2, 0, pi/2, pi),
                     labels = c(expression(-pi, -pi/2, 0, pi/2, pi))) +
  labs(x = "Turn Angle (radians)", y = "", 
       title = "Phenology fruits’ energetic content") +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        aspect.ratio = 1,
        plot.title = element_text(size = 20, hjust = .5),
        axis.text.y = element_text(size = 18, color = "black"),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_blank(),
        axis.title.x = element_text(size = 20),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        strip.text = element_text(size = 12),
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0))
```

#### In @fig-KJ_bot_mov_dry we can see that there is a significant effect of the fruits energetic content on the turn angles, with movement more directed where the fruits energetic content is higher.

```{r}
#| label: fig-KJ_bot_mov_dry
#| fig-width: 7
#| fig-asp: 6 
#| fig-alt: |
#|   A plot showing how number of fruits energetic content affected both step lengths and turn angles in the rainy season
#| fig-cap: |
#|       Interaction between number of fruits energetic content (KJ) and both step lengths and turn angles in the rainy season 
#| warning: FALSE  
p_ta_KJ_bot_dry
```

# 4. Conclusions

Our integrated step selection function models, fitted using predicted social and environmental landscapes, revealed season-specific drivers of BLT movement. During the rainy season, we found that BLTs moved toward areas with higher densities of known feeding trees (coeff. = 0.522, *z* = 6.93, *p* \< 0.001), taller canopies (coeff. = 0.200, *z* = 3.29, *p* = 0.001), and higher long call emission (coeff. = 0.202, *z* = 3.57, *p* \< 0.001). When starting from areas with higher densities of known fruiting trees, BLT movement was more tortuous (coeff. = –0.206, *z* = –3.07, *p* = 0.002) and the group took shorter steps, suggesting foraging within resource-rich patches. Additionally, BLT movement steps were longer and marginally more directed when moving from areas associated with higher energy expenditure, suggesting a move away from areas where movement incurs high costs.

During the dry season, we found that BLTs moved toward areas with higher densities of known feeding trees (coeff. = 0.340, *z* = 3.63, *p* \< 0.001) and taller canopies (coeff. = 0.271, *z* = 4.67, *p* \< 0.001). They turned more (coeff. = –0.238, *z* = –3.17, *p* = 0.002) and took shorter steps when moving from areas where alarm calls were emitted more frequently. We also found that BLTs moved with more turns (i.e. change of direction) from areas with lower long call activity (coeff. = –0.223, *z* = –3.00, *p* = 0.003) but showed more straight movements when moving from areas where fruits had higher energetic content (coeff. = 0.343, *z* = 2.02, *p* = 0.044).

Overall, our findings reveal that BLTs integrate spatial memory, structural habitat cues, energetic state, and social signalling to guide movement decisions in a complex and fluctuating environment. In a small, high-density forest fragment, these interacting drivers allow individuals to navigate spatial and temporal variation in both ecological opportunity and social pressure. This study underscores the importance of considering multiple behavioural and ecological dimensions in understanding primate route planning, providing new insight into how cognitive, social, and environmental processes jointly shape movement in seasonal forest fragments.

### References
