---
title: "AKDE_AJP"
author: "Felipe Bufalo"
date: "2023-05-13"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Home range estimation with AKDE for all four areas used for the publication to AJP

## This method is indicated for auto-correlated GPS data. Also, I will implement in the estimates hard boundaries (i.e., forest borders) for the estimates to be forced inside forest areas.

### First, load the packages and data
```{r}
setwd("~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/HR_Guarei-Taquara-Suzano")

library(ctmm)
library(dplyr)
library(viridisLite)
library(sf)
library(rgdal)

load("AKDE.rda")
```


# Starting with Guareí
### Convert to telemetry object (ctmm format)
```{r}
df_guarei <- as.telemetry(df_guarei)
summary(df_guarei)
guarei_sp <- as(guarei_sf, "Spatial")
```


### Convert the shapefiles to spatial objects fo ctmm to read as boundary
```{r}
guarei_sp <- as(guarei_sf, "Spatial")
```

### Plot data for visual inspection
```{r}
plot(df_guarei, col = viridisLite::viridis(6))
```

### Generate a variogram using ctmm's variogram() function
```{r}
vg <- variogram(df_guarei)
plot(vg)
```

### Guesstimate the model's parameters
### This is done to provide initial values for the formal model fitting/selection
```{r}
GUESS_guarei <- ctmm.guess(df_guarei,interactive=FALSE); summary(GUESS_guarei)
```

### Fit and select models
```{r}
FIT_ML <- ctmm.select(df_guarei, GUESS_guarei, method = 'ML'); summary(FIT_ML)
```

### Estimate an animal's home range conditioned on the selected autocorrelation model
### I will use both 
```{r}
UDakdedf_ML <- akde(df_guarei,FIT_ML, SP=guarei_sp, SP.in=TRUE);
summary(UDakdedf_ML)
```

### Look at the estimated home range
```{r}
plot(UDakdedf_ML)
title(expression("Guareí ML - OUF AKDE" ["C"]))
```

### Some parameters for the estimations
```{r}
summary(UDakdedf_ML)$DOF["area"] # effective sample size of guareí GC01, ML
nrow(df_guarei) # absolute sample size
```

### For exporting the shapfiles:
```{r}
#shapefile_polygons <- as.sf(UDakdedf_ML, level.UD=0.95, level=0.95)
#plot(shapefile_polygons)

#middle_polygon <- shapefile_polygons[2,]  #selects the second row which is the
#95% est middle polygon
#plot(middle_polygon)

#class(middle_polygon)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/Guarei/hr_guarei/ctmm/
#guarei_akde95.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon, output_path, layer = "guarei_akde95", driver = "GPKG")

## Exporting shapfile for akde 70%
#shapefile_polygons_70 <- as.sf(UDakdedf_ML, level.UD=0.70, level=0.95)
#plot(shapefile_polygons_70)

#middle_polygon_70 <- shapefile_polygons_70[2,]  #selects the second row which 
#is the 95% est middle polygon
#plot(middle_polygon_70)

#class(middle_polygon_70)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/Guarei/hr_guarei/ctmm/
#guarei_akde70.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon_70, output_path, layer = "guarei_akde70", 
#driver = "GPKG")

## Exporting shapfile for akde 50%
#shapefile_polygons_50 <- as.sf(UDakdedf_ML, level.UD=0.50, level=0.95)
#plot(shapefile_polygons_50)

#middle_polygon_50 <- shapefile_polygons_50[2,]  #selects the second row which 
#is the 95% est middle polygon
#plot(middle_polygon_50)

#class(middle_polygon_50)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/Guarei/hr_guarei/ctmm/
#guarei_akde50.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon_50, output_path, layer = "guarei_akde50", 
#driver = "GPKG")
```


# Suzano
### Convert to telemetry object (ctmm format)
```{r}
df_suzano <- as.telemetry(df_suzano)
summary(df_suzano)
```

### Convert the shapefiles to spatial objects fo ctmm to read as boundary
```{r}
suzano_sp <- as(suzano_sf, "Spatial")
```

### Plot data for visual inspection
```{r}
plot(df_suzano, col = viridisLite::viridis(6))
```

### Generate a variogram using ctmm's variogram() function
```{r}
vg <- variogram(df_suzano)
plot(vg)
```

### Guesstimate the model's parameters
### This is done to provide initial values for the formal model fitting/selection
```{r}
GUESS_suzano <- ctmm.guess(df_suzano,interactive=FALSE); summary(GUESS_suzano)
```

### Fit and select models
```{r}
FIT_ML_suzano <- ctmm.select(df_suzano, GUESS_suzano, method = 'ML')
summary(FIT_ML_suzano)
```


### Estimate an animal's home range conditioned on the selected autocorrelation model
### I will use both 
```{r}
UDakdedf_suzano_ML <- akde(df_suzano,FIT_ML_suzano, SP=suzano_sp, SP.in=TRUE);summary(UDakdedf_suzano_ML)
```

### Look at the estimated home range
```{r}
plot(UDakdedf_suzano_ML)
title(expression("Guareí ML - OUF AKDE" ["C"]))
```

### Some parameters for the estimations
```{r}
summary(UDakdedf_suzano_ML)$DOF["area"] # effective sample size of suzano group,
#ML
nrow(df_suzano) # absolute sample size
```

### For exporting the shapfiles:
```{r}
#shapefile_polygons <- as.sf(UDakdedf_suzano_ML, level.UD=0.95, level=0.95)
#plot(shapefile_polygons)

#middle_polygon <- shapefile_polygons[2,]  #selects the second row which is the
#95% est middle polygon
#plot(middle_polygon)

#class(middle_polygon)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/suzano/hr_suzano/ctmm/
#suzano_akde95.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon, output_path, layer = "suzano_akde95", driver = "GPKG")

## Exporting shapfile for akde 70%
#shapefile_polygons_70 <- as.sf(UDakdedf_suzano_ML, level.UD=0.70, level=0.95)
#plot(shapefile_polygons_70)

#middle_polygon_70 <- shapefile_polygons_70[2,]  #selects the second row which 
#is the 95% est middle polygon
#plot(middle_polygon_70)

#class(middle_polygon_70)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/suzano/hr_suzano/ctmm/
#suzano_akde70.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon_70, output_path, layer = "suzano_akde70", 
#driver = "GPKG")

## Exporting shapfile for akde 50%
#shapefile_polygons_50 <- as.sf(UDakdedf_suzano_ML, level.UD=0.50, level=0.95)
#plot(shapefile_polygons_50)

#middle_polygon_50 <- shapefile_polygons_50[2,]  #selects the second row which 
#is the 95% est middle polygon
#plot(middle_polygon_50)

#class(middle_polygon_50)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/suzano/hr_suzano/ctmm/
#suzano_akde50.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon_50, output_path, layer = "suzano_akde50", 
#driver = "GPKG")
```


# Taquara
### Convert to telemetry object (ctmm format)
```{r}
df_taquara <- as.telemetry(df_taquara)
summary(df_taquara)
```

### Convert the shapefiles to spatial objects fo ctmm to read as boundary
```{r}
taquara_sp <- as(taquara_sf, "Spatial")
```

### Plot data for visual inspection
```{r}
plot(df_taquara, col = viridisLite::viridis(6))
```

### Generate a variogram using ctmm's variogram() function
```{r}
vg <- variogram(df_taquara)
plot(vg)
```

### Guesstimate the model's parameters
### This is done to provide initial values for the formal model fitting/selection
```{r}
GUESS_taquara <- ctmm.guess(df_taquara,interactive=FALSE); 
summary(GUESS_taquara)
```

### Fit and select models
```{r}
FIT_ML_taquara <- ctmm.select(df_taquara, GUESS_taquara, method = 'ML')
summary(FIT_ML_taquara)
```


### Estimate an animal's home range conditioned on the selected autocorrelation model
### I will use both 
```{r}
UDakdedf_taquara_ML <- akde(df_taquara,FIT_ML_taquara, SP=taquara_sp, SP.in=TRUE);summary(UDakdedf_taquara_ML)
```

### Look at the estimated home range
```{r}
plot(UDakdedf_taquara_ML)
title(expression("Guareí ML - OUF AKDE" ["C"]))
```

### Some parameters for the estimations
```{r}
summary(UDakdedf_taquara_ML)$DOF["area"] # effective sample size of taquara 
#group, ML
nrow(df_taquara) # absolute sample size
```

### For exporting the shapfiles:
```{r}
#shapefile_polygons <- as.sf(UDakdedf_taquara_ML, level.UD=0.95, level=0.95)
#plot(shapefile_polygons)

#middle_polygon <- shapefile_polygons[2,]  #selects the second row which is the
#95% est middle polygon
#plot(middle_polygon)

#class(middle_polygon)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/taquara/hr_taquara/ctmm/
#taquara_akde95.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon, output_path, layer = "taquara_akde95", driver = "GPKG")

## Exporting shapfile for akde 70%
#shapefile_polygons_70 <- as.sf(UDakdedf_taquara_ML, level.UD=0.70, level=0.95)
#plot(shapefile_polygons_70)

#middle_polygon_70 <- shapefile_polygons_70[2,]  #selects the second row which 
#is the 95% est middle polygon
#plot(middle_polygon_70)

#class(middle_polygon_70)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/taquara/hr_taquara/ctmm/
#taquara_akde70.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon_70, output_path, layer = "taquara_akde70", 
#driver = "GPKG")

## Exporting shapfile for akde 50%
#shapefile_polygons_50 <- as.sf(UDakdedf_taquara_ML, level.UD=0.50, level=0.95)
#plot(shapefile_polygons_50)

#middle_polygon_50 <- shapefile_polygons_50[2,]  #selects the second row which 
#is the 95% est middle polygon
#plot(middle_polygon_50)

#class(middle_polygon_50)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/taquara/hr_taquara/ctmm/
#taquara_akde50.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon_50, output_path, layer = "taquara_akde50", 
#driver = "GPKG")
```


# Santa Maria (sma)
### Convert to telemetry object (ctmm format)
```{r}
df_sma <- as.telemetry(df_sma)
summary(df_sma)
```

### Convert the shapefiles to spatial objects fo ctmm to read as boundary
```{r}
sma_sp <- as(sma_sf, "Spatial")
```

### Plot data for visual inspection
```{r}
plot(df_sma, col = viridisLite::viridis(6))
```

### Generate a variogram using ctmm's variogram() function
```{r}
vg <- variogram(df_sma)
plot(vg)
```

### Guesstimate the model's parameters
### This is done to provide initial values for the formal model fitting/selection
```{r}
GUESS_sma <- ctmm.guess(df_sma,interactive=FALSE); 
summary(GUESS_sma)
```

### Fit and select models
```{r}
FIT_ML_sma <- ctmm.select(df_sma, GUESS_sma, method = 'ML')
summary(FIT_ML_sma)
```


### Estimate an animal's home range conditioned on the selected autocorrelation model
### I will use both 
```{r}
UDakdedf_sma_ML <- akde(df_sma,FIT_ML_sma, SP=sma_sp, SP.in=TRUE);
summary(UDakdedf_sma_ML)
```

### Look at the estimated home range
```{r}
plot(UDakdedf_sma_ML)
title(expression("Guareí ML - OUF AKDE" ["C"]))
```

### Some parameters for the estimations
```{r}
summary(UDakdedf_sma_ML)$DOF["area"] # effective sample size of sma 
#group, ML
nrow(df_sma) # absolute sample size
```

### For exporting the shapfiles:
```{r}
#shapefile_polygons <- as.sf(UDakdedf_sma_ML, level.UD=0.95, level=0.95)
#plot(shapefile_polygons)

#middle_polygon <- shapefile_polygons[2,]  #selects the second row which is the
#95% est middle polygon
#plot(middle_polygon)

#class(middle_polygon)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/sma/hr_sma/ctmm/
#sma_akde95.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon, output_path, layer = "sma_akde95", driver = "GPKG")

## Exporting shapfile for akde 70%
#shapefile_polygons_70 <- as.sf(UDakdedf_sma_ML, level.UD=0.70, level=0.95)
#plot(shapefile_polygons_70)

#middle_polygon_70 <- shapefile_polygons_70[2,]  #selects the second row which 
#is the 95% est middle polygon
#plot(middle_polygon_70)

#class(middle_polygon_70)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/sma/hr_sma/ctmm/
#sma_akde70.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon_70, output_path, layer = "sma_akde70", 
#driver = "GPKG")

## Exporting shapfile for akde 50%
#shapefile_polygons_50 <- as.sf(UDakdedf_sma_ML, level.UD=0.50, level=0.95)
#plot(shapefile_polygons_50)

#middle_polygon_50 <- shapefile_polygons_50[2,]  #selects the second row which 
#is the 95% est middle polygon
#plot(middle_polygon_50)

#class(middle_polygon_50)

# Specify the path to the output geopackage file
#output_path <- "~/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/sma/hr_sma/ctmm/
#sma_akde50.gpkg"

# Write the "sf" object to a geopackage
#st_write(middle_polygon_50, output_path, layer = "sma_akde50", 
#driver = "GPKG")
```

