---
title: "Distance from locomotion CPs to to borders"
author: "Felipe Bufalo"
date: "2023-05-22"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/febufalo/Library/CloudStorage/GoogleDrive-felipe.bufalo@unesp.br/My Drive/Unesp/Mestrado/Analises/Analises_Publication/CPT_Guarei-Taquara-Suzano/CP_distance to border")
```

# Script to calculate distances from CP points associated to locomotion to the borders of both 95% akde home ranges and core areas (70-75% akde)

## Load packages
```{r}
library(sf)
library(nngeo)
library(ggplot2)
library(maptools)
library(ggspatial)
```

# *Guareí!*
## Read and treat files (shapes and data frame)
```{r}
# Read files
guarei_akde95 <- st_read("guarei_akde95.gpkg")
plot(guarei_akde95)
# convert a polygon geometry to a data frame
guarei_akde95 <- st_coordinates(guarei_akde95)
# Convert to data frame
guarei_akde95 <- as.data.frame(guarei_akde95)
head(guarei_akde95)
# Rename the columns
colnames(guarei_akde95) <- c("x", "y")
head(guarei_akde95)

guarei_akde70 <- st_read("guarei_akde70.gpkg")
plot(guarei_akde70)
# convert a polygon geometry to a data frame
guarei_akde70 <- st_coordinates(guarei_akde70)
# Convert to data frame
guarei_akde70 <- as.data.frame(guarei_akde70)
head(guarei_akde70)
# Rename the columns
colnames(guarei_akde70) <- c("x", "y")
head(guarei_akde70)

df_cp_guarei <- read.csv("CP_distanceborder_guarei.csv")
head(df_cp_guarei)
```

## Filter lines just to have locomotion cps (the ones I want to calculate distance to the borders)
```{r}
df_cp_guarei <- df_cp_guarei[df_cp_guarei$CP_q6 == "Locomotion", ]
head(df_cp_guarei)
```

## Plot the data to have a look at the situation
```{r}
ggplot()+
  geom_polygon(data = guarei_akde95, aes(x=x, y= y),
               fill = "palegreen4", size = 1)+
  geom_polygon(data = guarei_akde70, aes(x=x, y= y),
               fill = "yellowgreen", colour = "black") +
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, size=2))+
  geom_point(data = df_cp_guarei, aes(x = longitude_x_GAL, 
                                      y = latitude_y_GAL), size = 1) +
  coord_fixed()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=2),
        plot.margin=unit(c(0.1,0,0,0),"cm"))+
  theme(axis.text.x = element_text(hjust=0.5, size = 8, colour ="Black"))+
  theme(axis.text.y = element_text(angle=90, hjust=0.5, 
                                   size = 8, colour ="Black"))+
  theme(axis.title.y = element_blank())+
  theme(axis.title.x = element_blank())+
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  annotation_scale(location = "tr", width_hint = 0.2, 
                   pad_x = unit(0.1, "in"), pad_y = unit(0.8, "in"))+
  annotation_north_arrow(location = "tr", which_north = "true", 
                         pad_x = unit(0.3, "in"), pad_y = unit(0.1, "in"),
                         style = north_arrow_fancy_orienteering)
```

## Some more data organization
```{r}
colnames(df_cp_guarei) <- c("date", "time", "y", "x", "behav", "cp_q6")
guarei_akde95[,3:4] <- NULL; head(guarei_akde95)
guarei_akde70[,3:4] <- NULL; head(guarei_akde70)
```

## Determine the crs for the sf objects!
```{r}
crs <- st_crs("+proj=utm +zone=22 +south +datum=WGS84 +units=m +no_defs")
```

## Convert all objects to sf class
```{r}
# Convert guarei_akde95 data frame to an sf object
guarei_akde95_sf <- st_as_sf(guarei_akde95, coords = c("x", "y"), crs = crs)

# Convert guarei_akde70 data frame to an sf object
guarei_akde70_sf <- st_as_sf(guarei_akde70, coords = c("x", "y"), crs = crs)

# Convert df_cp_guarei data frame to an sf object
df_cp_guarei_sf <- st_as_sf(df_cp_guarei, coords = c("x", "y"), crs = crs)
```

## Use the st_nn function (package nngeo) to calculate the distance between the locomotion CP and the nearest point in the borders. 
## First 95% AKDE border
```{r}
# k = 1 for finding 1 nearest neighbor
distances_95 <- nngeo::st_nn(df_cp_guarei_sf, guarei_akde95_sf, k = 1, 
                             returnDist = T)
head(distances_95)
```

## Than 70% AKDE border
```{r}
distances_70 <- nngeo::st_nn(df_cp_guarei_sf, guarei_akde70_sf, k = 1, 
                             returnDist = T)
head(distances_70)
```

## Extract the distance values from the list
```{r}
distance_values_95 <- distances_95$dist
distance_values_70 <- distances_70$dist
```

## Add the distance values as a new column in df_cp_guarei
```{r}
df_cp_guarei$dist70 <- distance_values_70
df_cp_guarei$dist95 <- distance_values_95
df_cp_guarei$dist95 <- as.numeric(df_cp_guarei$dist95)
df_cp_guarei$dist70 <- as.numeric(df_cp_guarei$dist70)
head(df_cp_guarei)
```

## Now, for the plots, I will classify the distances into categories so I can compare frequency proportions of locomotion CPs between distance categories!
```{r}
breaks <- c(0, 20, 40, 60, 80, 100, 120, 140, 160)
```

## Create the distance categories
```{r}
dist_categories_guarei <- cut(df_cp_guarei$dist95, breaks = breaks, 
                              labels = FALSE, include.lowest = TRUE, 
                              right = FALSE)
```

## Rename the category labels
```{r}
category_labels <- paste0("cat", 1:length(breaks[-1]))
```

## Assign category labels to the distance categories
```{r}
dist_categories_guarei <- factor(dist_categories_guarei, 
                                 levels = 1:length(breaks[-1]), 
                                 labels = category_labels)
df_cp_guarei$cat_dist_95 <- dist_categories_guarei
df_cp_guarei
```

## Compute the frequency of each category
```{r}
freq_table_guarei95 <- table(dist_categories_guarei)
```

## Compute the proportion of location points for each distance category
```{r}
proportions_guarei95 <- freq_table_guarei95 / sum(freq_table_guarei95)
plot_data_guarei95 <- data.frame(Proportion = proportions_guarei95,
                                 Count = freq_table_guarei95)
```
## Add row names as a separate column
```{r}
plot_data_guarei95$Category <- row.names(plot_data_guarei95)
```

## Organize rows & columns
```{r}
row.names(plot_data_guarei95) <- NULL
plot_data_guarei95$Proportion.dist_categories_guarei <- NULL
plot_data_guarei95$Count.dist_categories_guarei <- NULL
colnames(plot_data_guarei95) <- c("Proportion", "Count", "Category")
plot_data_guarei95 <- plot_data_guarei95[, c("Category", "Count", "Proportion")]
plot_data_guarei95$Category <- as.numeric(plot_data_guarei95$Category)
plot_data_guarei95
```

## Convert Category column to numeric, with the distances they represent
```{r}
plot_data_guarei95$Category <- as.numeric(c(0, 20, 40, 60, 80, 100, 120, 
                                            140))
plot_data_guarei95
```
## Plot the proportion of location points by distance category to the 95% AKDE border
```{r}
ggplot(plot_data_guarei95, aes(x = Category, y = Proportion, group = 1)) +
  geom_line() +
  theme_bw() +
  xlab("Distance to AKDE 95% border (m)") +
  ylab("Proportion of Locomotion CPs") +
  scale_y_continuous(expand = c(0.05, 0)) +
  scale_x_continuous(breaks = seq(0, 740, by = 20)) +
  ggtitle("Guareí")
```

## I will try some tests to see if there is any relationship
## Kolmogorov-Smirnov test
```{r}
ks.test(df_cp_guarei$dist95, "punif", min = 0, max = max(df_cp_guarei$dist95))
# As expected from the plot, the distribution is not different from a unirform 
# distribution!
```

## Generalized Linear model (poisson distribution)
```{r}
glm_guarei_95 <- glm(Count ~  Category, data = plot_data_guarei95,
                     family = "poisson")
summary(glm_guarei_95)
# No interaction!
```

## *Now, work with the 70% - Core area*

## Create the custom breaks for categories 95%
```{r}
breaks <- c(0, 20, 40, 60, 80, 100)
```

## Create the distance categories
```{r}
dist_categories_guarei <- cut(df_cp_guarei$dist70, breaks = breaks, 
                              labels = FALSE, include.lowest = TRUE, 
                              right = FALSE)
```

## Rename the category labels
```{r}
category_labels_guarei_70 <- paste0("cat", 1:length(breaks[-1]))
```

## Assign category labels to the distance categories
```{r}
dist_categories_guarei <- category_labels_guarei_70[dist_categories_guarei]
category_labels_guarei_70 <- paste0("cat", 1:length(unique(
  dist_categories_guarei)))
df_cp_guarei$cat_dist_70 <- factor(dist_categories_guarei, 
                                   levels = unique(dist_categories_guarei), 
                                   labels = category_labels_guarei_70)
df_cp_guarei
```

## Compute frequencies for each category
```{r}
freq_table_guarei70 <- table(dist_categories_guarei)
```

## Compute the proportion of location points for each distance category
```{r}
proportions_guarei70 <- freq_table_guarei70 / sum(freq_table_guarei70)

plot_data_guarei70 <- data.frame(Proportion = proportions_guarei70,
                                 Count = freq_table_guarei70)
```

## Organize rows & columns
```{r}
plot_data_guarei70$Category <- row.names(plot_data_guarei70)
# Reset row names
row.names(plot_data_guarei70) <- NULL
plot_data_guarei70$Proportion.dist_categories_guarei <- NULL
plot_data_guarei70$Count.dist_categories_guarei <- NULL
# Change the column names of the data frame
colnames(plot_data_guarei70) <- c("Proportion", "Count", "Category")
plot_data_guarei70 <- plot_data_guarei70[, c("Category", "Count", "Proportion")]
plot_data_guarei70$Category <- as.numeric(plot_data_guarei70$Category)
plot_data_guarei70$Category <- as.numeric(c(0, 40, 60, 80))
plot_data_guarei70
```

## Plot the proportion of location points by distance category
```{r}
ggplot(plot_data_guarei70, aes(x = Category, y = Proportion, group = 1)) +
  geom_line() +
  theme_bw() +
  xlab("Distance to AKDE 70% border (m)") +
  ylab("Proportion of Locomotion CPs") +
  scale_y_continuous(expand = c(0.05, 0)) +
  scale_x_continuous(breaks = seq(0, 80, by = 20)) +
  ggtitle("Guareí")
```

## I will try some tests to see if there is any relationship
## Kolmogorov-Smirnov test
```{r}
ks.test(df_cp_guarei$dist70, "punif", min = 0, max = max(df_cp_guarei$dist70))
# As expected from the plot, the distribution is not different from a unirform 
# distribution!
```

## Generalized Linear model (poisson distribution)
```{r}
glm_guarei_70 <- glm(Count ~  Category, data = plot_data_guarei70, 
                    family = "poisson")
summary(glm_guarei_70)
# As expected, does not work
```


# *Taquara!*
## Read files
```{r}
taquara_akde95 <- st_read("taquara_akde95.gpkg")
plot(taquara_akde95)
# convert a polygon geometry to a data frame
taquara_akde95 <- st_coordinates(taquara_akde95)
# Convert to data frame
taquara_akde95 <- as.data.frame(taquara_akde95)
head(taquara_akde95)
# Rename the columns
colnames(taquara_akde95) <- c("x", "y")
head(taquara_akde95)
taquara_akde75 <- st_read("taquara_akde75.gpkg")
plot(taquara_akde75)
# convert a polygon geometry to a data frame
taquara_akde75 <- st_coordinates(taquara_akde75)
# Convert to data frame
taquara_akde75 <- as.data.frame(taquara_akde75)
head(taquara_akde75)
# Rename the columns
colnames(taquara_akde75) <- c("x", "y")
head(taquara_akde75)
df_cp_taquara <- read.csv("CP_distanceborder_taquara.csv")
head(df_cp_taquara)
```
## Filter lines just to have locomotion cps (the ones I want to calculate distance to the borders)
```{r}
df_cp_taquara <- df_cp_taquara[df_cp_taquara$cp_q5 == "Locomotion", ]
head(df_cp_taquara)
```

## Plot the data to understand the distribution of the CPs
```{r}
ggplot()+
  geom_polygon(data = taquara_akde95, aes(x=x, y= y),
               fill = "palegreen4", size = 1)+
  geom_polygon(data = taquara_akde75, aes(x=x, y= y),
               fill = "yellowgreen", colour = "black") +
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, size=2))+
  geom_point(data = df_cp_taquara, aes(x = longitude_x_GAL, 
                                       y = latitude_y_GAL), size = 1) +
  coord_fixed()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=2),
        plot.margin=unit(c(0.1,0,0,0),"cm"))+
  theme(axis.text.x = element_text(hjust=0.5, size = 8, colour ="Black"))+
  theme(axis.text.y = element_text(angle=90, hjust=0.5, 
                                   size = 8, colour ="Black"))+
  theme(axis.title.y = element_blank())+
  theme(axis.title.x = element_blank())+
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  annotation_scale(location = "tr", width_hint = 0.2, 
                   pad_x = unit(0.1, "in"), pad_y = unit(0.8, "in"))+
  annotation_north_arrow(location = "tr", which_north = "true", 
                         pad_x = unit(0.3, "in"), pad_y = unit(0.1, "in"),
                         style = north_arrow_fancy_orienteering)
```

## Some more data organization
```{r}
colnames(df_cp_taquara) <- c("date", "time", "y", "x", "behav", "cp_q5")
taquara_akde95[,3:4] <- NULL; head(taquara_akde95)
taquara_akde75[,3:4] <- NULL; head(taquara_akde75)
```

## Convert data frames into sf objects
```{r}
# Convert taquara_akde95 data frame to an sf object
taquara_akde95_sf <- st_as_sf(taquara_akde95, coords = c("x", "y"), crs = crs)
# Convert taquara_akde70 data frame to an sf object
taquara_akde75_sf <- st_as_sf(taquara_akde75, coords = c("x", "y"), crs = crs)
# Convert df_cp_taquara data frame to an sf object
df_cp_taquara_sf <- st_as_sf(df_cp_taquara, coords = c("x", "y"), crs = crs)
```

## Use the st_nn function (package nngeo) to calculate the distance between the locomotion CP and the nearest point in the borders. 
## First 95% AKDE border
```{r}
# k = 1 for finding 1 nearest neighbor
distances_95_taquara <- nngeo::st_nn(df_cp_taquara_sf, 
                                     taquara_akde95_sf, k = 1, returnDist = T)
head(distances_95_taquara)
```

## Now the 75% core area
```{r}
distances_75_taquara <- nngeo::st_nn(df_cp_taquara_sf, 
                                     taquara_akde75_sf, k = 1, returnDist = T)
head(distances_75_taquara)
```

## Extract the distance values from the list
```{r}
distance_values_95_taquara <- distances_95_taquara$dist
distance_values_75_taquara <- distances_75_taquara$dist
```

## Add the distance values as a new column in df_cp_taquara_sf
```{r}
df_cp_taquara$dist75 <- distance_values_75_taquara
df_cp_taquara$dist95 <- distance_values_95_taquara
df_cp_taquara$dist95 <- as.numeric(df_cp_taquara$dist95)
df_cp_taquara$dist75 <- as.numeric(df_cp_taquara$dist75)
head(df_cp_taquara)
```

## Now, for the plots, I will classify the distances into categories so I can compare frequency proportions of locomotion CPs between distance categories! 
## 95% AKDE first
```{r}
breaks <- c(0, 20, 40, 60, 80, 100, 120, 140, 160, 180, 200, 220, 240, 
            260, 280, 300, 320, 340, 360, 380, 400, 420, 460, 480, 500, 520, 
            540, 560, 580,600, 620, 640, 660, 680, 700, 720, 740, 760)
```

## Create the distance categories
```{r}
dist_categories_taquara <- cut(df_cp_taquara$dist95, breaks = breaks, 
                               labels = FALSE, include.lowest = TRUE, 
                               right = FALSE)
# Rename the category labels
category_labels <- paste0("cat", 1:length(breaks[-1]))
```

## Assign category labels to the distance categories
```{r}
dist_categories_taquara <- factor(dist_categories_taquara, 
                                  levels = 1:length(breaks[-1]), 
                                  labels = category_labels)
df_cp_taquara$cat_dist_95 <- dist_categories_taquara
df_cp_taquara
```

## Compute the frequency of points in each category
```{r}
freq_table_taquara95 <- table(dist_categories_taquara)
```

## Compute the proportion of location points for each distance category
```{r}
proportions_taquara95 <- freq_table_taquara95 / sum(freq_table_taquara95)
plot_data_taquara95 <- data.frame(Proportion = proportions_taquara95,
                                  Count = freq_table_taquara95)
```

## Add row names as a separate column
```{r}
plot_data_taquara95$Category <- row.names(plot_data_taquara95)
```

## Columns and rows organization
```{r}
row.names(plot_data_taquara95) <- NULL
plot_data_taquara95$Proportion.dist_categories_taquara <- NULL
plot_data_taquara95$Count.dist_categories_taquara <- NULL
# Change the column names of the data frame
colnames(plot_data_taquara95) <- c("Proportion", "Count", "Category")
plot_data_taquara95 <- plot_data_taquara95[, c("Category", "Count", 
                                               "Proportion")]
plot_data_taquara95$Category <- as.numeric(plot_data_taquara95$Category)
plot_data_taquara95
```

## Convert Category column to numeric, with the distances they represent
```{r}
plot_data_taquara95$Category <- as.numeric(c(0, 20, 40, 60, 80, 100, 120, 140, 
                                             160, 180, 200, 220, 240, 260, 280, 
                                             300, 320, 340, 360, 380, 400, 420, 
                                             460, 480, 500, 520, 540, 560, 580,
                                             600, 620, 640, 660, 680, 700, 720,
                                             740))

plot_data_taquara95
```

## Plot the proportion of location points by distance category
```{r}
ggplot(plot_data_taquara95, aes(x = Category, y = Proportion, group = 1)) +
  geom_line() +
  theme_bw() +
  xlab("Distance to AKDE 95% border (m)") +
  ylab("Proportion of Locomotion CPs") +
  scale_y_continuous(expand = c(0.05, 0)) +
  scale_x_continuous(breaks = seq(0, 740, by = 20))+
  ggtitle("Taquara")
```

## I will try some tests to see if there is any relationship
## Kolmogorov-Smirnov test
```{r}
ks.test(df_cp_taquara$dist95, "punif", min = 0, max = max(df_cp_taquara$dist95))
# Distribution is not different from a unirform distribution (p>0.05)
```

## Generealized Linear model (poisson distribution)
```{r}
glm_taquara_95 <- glm(Count ~  Category, data = plot_data_taquara95, 
                      family = "poisson")
summary(glm_taquara_95)
# No correlation
```

# *Taquara 70% AKDE - Core area*
## Create the custom breaks for categories
```{r}
breaks <- c(0, 20, 40, 60, 80, 100, 120, 140, 160, 180, 200, 220, 240, 260, 280,
            300, 320, 340, 360, 380, 400, 420, 460, 480, 500, 520, 540)
```

## Create the distance categories
```{r}
dist_categories_taquara75 <- cut(df_cp_taquara$dist75, breaks = breaks, 
                                 labels = FALSE, include.lowest = TRUE, 
                                 right = FALSE)
```

## Rename the category labels
```{r}
category_labels <- paste0("cat", 1:length(breaks[-1]))
```

## Assign category labels to the distance categories
```{r}
dist_categories_taquara75 <- factor(dist_categories_taquara75, 
                                    levels = 1:length(breaks[-1]), 
                                    labels = category_labels)
df_cp_taquara$cat_dist_75 <- dist_categories_taquara75
df_cp_taquara
```
## Compute frequencies of points per distance category
```{r}
freq_table_taquara75 <- table(dist_categories_taquara75)
```
## Compute the proportion of location points for each distance category
```{r}
proportions_taquara75 <- freq_table_taquara75 / sum(freq_table_taquara75)
plot_data_taquara75 <- data.frame(Proportion = proportions_taquara75,
                                  Count = freq_table_taquara75)
```

## Rows and columns organization
```{r}
plot_data_taquara75$Category <- row.names(plot_data_taquara75)
# Reset row names
row.names(plot_data_taquara75) <- NULL
plot_data_taquara75$Proportion.dist_categories_taquara75 <- NULL
plot_data_taquara75$Count.dist_categories_taquara75 <- NULL
# Change the column names of the data frame
colnames(plot_data_taquara75) <- c("Proportion", "Count", "Category")
plot_data_taquara75 <- plot_data_taquara75[, c("Category", "Count", 
                                               "Proportion")]
plot_data_taquara75$Category <- as.numeric(plot_data_taquara75$Category)
plot_data_taquara75$Category <- as.numeric(c(0, 20, 40, 60, 80, 100, 120, 140, 
                                             160, 180, 200, 220, 240, 260, 280, 
                                             300, 320, 340, 360, 380, 400, 420, 
                                             460, 480, 500, 520))
plot_data_taquara75
```

## Plot the proportion of location points by distance category
```{r}
ggplot(plot_data_taquara75, aes(x = Category, y = Proportion, group = 1)) +
  geom_line() +
  theme_bw() +
  xlab("Distance to AKDE 75% border (m)") +
  ylab("Proportion of Locomotion CPs") +
  scale_y_continuous(expand = c(0.05, 0)) +
  scale_x_continuous(breaks = seq(0, 520, by = 20))+
  ggtitle("Taquara")
```

## I will try some tests to see if there is any relationship
## Kolmogorov-Smirnov test
```{r}
ks.test(df_cp_taquara$dist75, "punif", min = 0, max = max(df_cp_taquara$dist75))
# Besides the plot looking like there is a tendency of higher proportion of 
# points closer to the 75% core area border, the test shows no significance..
```

## Generealized Linear model (poisson distribution)
```{r}
glm_taquara_75 <- glm(Count ~  Category, data = plot_data_taquara75, 
                      family = "poisson")
summary(glm_taquara_75)
# No correlation
```

## I will try another king of plot to see the trend
```{r}
plot=ggplot(plot_data_taquara75, aes(y=Proportion , x= Category))+ 
  geom_point()+
  geom_smooth(method = "glm", formula = y ~ x)+
  xlab("Distance to AKDE 75% border (m)") +
  ylab("Proportion of Locomotion CPs") 
plot + theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(), 
              panel.background = element_blank(), 
              axis.line = element_line(colour = "black"))
```
### Does not look as expected, indeed. Actually, the trend is to have a higher density of proportions with increases in distance :(

# *Santa Maria!*
# Read files
```{r}
sma_akde95 <- st_read("sma_akde95.gpkg")
plot(sma_akde95)
# convert a polygon geometry to a data frame
sma_akde95 <- st_coordinates(sma_akde95)
# Convert to data frame
sma_akde95 <- as.data.frame(sma_akde95)
head(sma_akde95)
# Rename the columns
colnames(sma_akde95) <- c("x", "y")
head(sma_akde95)
sma_akde75 <- st_read("sma_akde75.gpkg")
plot(sma_akde75)
# convert a polygon geometry to a data frame
sma_akde75 <- st_coordinates(sma_akde75)
# Convert to data frame
sma_akde75 <- as.data.frame(sma_akde75)
head(sma_akde75)
# Rename the columns
colnames(sma_akde75) <- c("x", "y")
head(sma_akde75)
df_cp_sma <- read.csv("CP_distanceborder_sma.csv")
head(df_cp_sma)
```

## Filter lines just to have locomotion cps (the ones I want to calculate distance to the borders)
```{r}
df_cp_sma <- df_cp_sma[df_cp_sma$cp_q5 == "Locomotion", ]
head(df_cp_sma)
```

## Plot to understand where the CPs are
```{r}
ggplot()+
  geom_polygon(data = sma_akde95, aes(x=x, y= y),
               fill = "palegreen4", size = 1)+
  geom_polygon(data = sma_akde75, aes(x=x, y= y),
               fill = "yellowgreen", colour = "black") +
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, size=2))+
  geom_point(data = df_cp_sma, aes(x = longitude_x_GAL, 
                                   y = latitude_y_GAL), size = 1) +
  coord_fixed()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=2),
        plot.margin=unit(c(0.1,0,0,0),"cm"))+
  theme(axis.text.x = element_text(hjust=0.5, size = 8, colour ="Black"))+
  theme(axis.text.y = element_text(angle=90, hjust=0.5, 
                                   size = 8, colour ="Black"))+
  theme(axis.title.y = element_blank())+
  theme(axis.title.x = element_blank())+
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  annotation_scale(location = "tr", width_hint = 0.2, 
                   pad_x = unit(0.1, "in"), pad_y = unit(0.8, "in"))+
  annotation_north_arrow(location = "tr", which_north = "true", 
                         pad_x = unit(0.3, "in"), pad_y = unit(0.1, "in"),
                         style = north_arrow_fancy_orienteering)
```

## Some more organization in the data frames
```{r}
colnames(df_cp_sma) <- c("date", "time", "y", "x", "behav", "cp_q5")
sma_akde95[,3:4] <- NULL; head(sma_akde95)
sma_akde75[,3:4] <- NULL; head(sma_akde75)
```

## Convert the data frames into sf objects
```{r}
# Convert sma_akde95 data frame to an sf object
sma_akde95_sf <- st_as_sf(sma_akde95, coords = c("x", "y"), crs = crs)
# Convert sma_akde70 data frame to an sf object
sma_akde75_sf <- st_as_sf(sma_akde75, coords = c("x", "y"), crs = crs)
# Convert df_cp_sma data frame to an sf object
df_cp_sma_sf <- st_as_sf(df_cp_sma, coords = c("x", "y"), crs = crs)
```

## Use the st_nn function (package nngeo) to calculate the distance between the locomotion CP and the nearest point in the borders. 
## First 95% AKDE border
```{r}
# k = 1 for finding 1 nearest neighbor
distances_95_sma <- nngeo::st_nn(df_cp_sma_sf, sma_akde95_sf, k = 1, 
                                 returnDist = T)
head(distances_95_sma)
distances_75_sma <- nngeo::st_nn(df_cp_sma_sf, sma_akde75_sf, k = 1, 
                                 returnDist = T)
head(distances_75_sma)
# Extract the distance values from the list
distance_values_95_sma <- distances_95_sma$dist
distance_values_75_sma <- distances_75_sma$dist
# Add the distance values as a new column in df_cp_sma_sf
df_cp_sma$dist75 <- distance_values_75_sma
df_cp_sma$dist95 <- distance_values_95_sma
head(df_cp_sma)
df_cp_sma$dist95 <- as.numeric(df_cp_sma$dist95)
df_cp_sma$dist75 <- as.numeric(df_cp_sma$dist75)
```

## Now, for the plots, I will classify the distances into categories so I can compare frequency proportions of locomotion CPs between distance categories!

## Create the custom breaks for categories -> 95% AKDE
```{r}
breaks <- c(0, 20, 40, 60, 80, 100, 120, 140, 160, 180, 200, 220, 240)
```

## Create the distance categories
```{r}
dist_categories_sma <- cut(df_cp_sma$dist95, breaks = breaks, labels = FALSE, 
                           include.lowest = TRUE, right = FALSE)
```

## Rename the category labels
```{r}
category_labels <- paste0("cat", 1:length(breaks[-1]))
```

## Assign category labels to the distance categories
```{r}
dist_categories_sma <- factor(dist_categories_sma, 
                              levels = 1:length(breaks[-1]), 
                              labels = category_labels)
df_cp_sma$cat_dist_95 <- dist_categories_sma
df_cp_sma
```

## Compute the frequency of CPs in each category
```{r}
freq_table_sma95 <- table(dist_categories_sma)
```

## Compute the proportion of location points for each distance category
```{r}
proportions_sma95 <- freq_table_sma95 / sum(freq_table_sma95)

plot_data_sma95 <- data.frame(Proportion = proportions_sma95,
                              Count = freq_table_sma95)
```

## Some more data organization
```{r}
plot_data_sma95$Category <- row.names(plot_data_sma95)
# Reset row names
row.names(plot_data_sma95) <- NULL
plot_data_sma95$Proportion.dist_categories_sma <- NULL
plot_data_sma95$Count.dist_categories_sma <- NULL
# Change the column names of the data frame
colnames(plot_data_sma95) <- c("Proportion", "Count", "Category")
plot_data_sma95 <- plot_data_sma95[, c("Category", "Count", "Proportion")]
plot_data_sma95$Category <- as.numeric(plot_data_sma95$Category)
# Convert Category column to numeric, with the distances they represent
plot_data_sma95$Category <- as.numeric(c(0, 20, 40, 60, 80, 100, 120, 140, 
                                         160, 180, 200, 220))
plot_data_sma95
```

## Plot the proportion of location points by distance category
```{r}
ggplot(plot_data_sma95, aes(x = Category, y = Proportion, group = 1)) +
  geom_line() +
  theme_bw() +
  xlab("Distance to AKDE 95% border (m)") +
  ylab("Proportion of Locomotion CPs") +
  scale_y_continuous(expand = c(0.05, 0)) +
  scale_x_continuous(breaks = seq(0, 740, by = 20))+
  ggtitle("Santa Maria")
```

## I will try some tests to see if there is any relationship
## Kolmogorov-Smirnov test
```{r}
ks.test(df_cp_sma$dist95, "punif", min = 0, max = max(df_cp_sma$dist95))
# the test shows no significance. 
```

## Generealized Linear model (poisson distribution)
```{r}
glm_sma_95 <- glm(Count ~  Category, data = plot_data_sma95, 
                      family = "poisson")
summary(glm_sma_95)
# No correlation
```

# *San Maria - 75% AKDE- Core area*
## Create the custom breaks for categories
```{r}
breaks <- c(0, 20, 40, 60, 80, 100)
```

## Create the distance categories
```{r}
dist_categories_sma75 <- cut(df_cp_sma$dist75, breaks = breaks, 
                             labels = FALSE, include.lowest = TRUE, 
                             right = FALSE)
```

## Rename the category labels
```{r}
category_labels <- paste0("cat", 1:length(breaks[-1]))
```

## Assign category labels to the distance categories
```{r}
dist_categories_sma75 <- factor(dist_categories_sma75, 
                                levels = 1:length(breaks[-1]), 
                                labels = category_labels)
df_cp_sma$cat_dist_75 <- dist_categories_sma75
df_cp_sma
```

## Compute the frequeicvy of location points according to categories
```{r}
freq_table_sma75 <- table(dist_categories_sma75)
```

## Compute the proportion of location points for each distance category
```{r}
proportions_sma75 <- freq_table_sma75 / sum(freq_table_sma75)
plot_data_sma75 <- data.frame(Proportion = proportions_sma75,
                              Count = freq_table_sma75)
```

## Some more data organization
```{r}
plot_data_sma75$Category <- row.names(plot_data_sma75)
# Reset row names
row.names(plot_data_sma75) <- NULL
plot_data_sma75$Proportion.dist_categories_sma75 <- NULL
plot_data_sma75$Count.dist_categories_sma75 <- NULL
# Change the column names of the data frame
colnames(plot_data_sma75) <- c("Proportion", "Count", "Category")
plot_data_sma75 <- plot_data_sma75[, c("Category", "Count", "Proportion")]
plot_data_sma75$Category <- as.numeric(plot_data_sma75$Category)
# Convert Category column to numeric, with the distances they represent
plot_data_sma75$Category <- as.numeric(c(0, 20, 40, 60, 80))
plot_data_sma75
```

## Plot the proportion of location points by distance category
```{r}
ggplot(plot_data_sma75, aes(x = Category, y = Proportion, group = 1)) +
  geom_line() +
  theme_bw() +
  xlab("Distance to AKDE 75% border (m)") +
  ylab("Proportion of Locomotion CPs") +
  scale_y_continuous(expand = c(0.05, 0)) +
  scale_x_continuous(breaks = seq(0, 520, by = 20))+
  ggtitle("Santa Maria")
```

## I will try some tests to see if there is any relationship
## Kolmogorov-Smirnov test
```{r}
ks.test(df_cp_sma$dist75, "punif", min = 0, max = max(df_cp_sma$dist75))
# the test shows significance. However, we have just 3 points and 2 locations. 
# Too few points for us to see any pattern
```

## Generealized Linear model (poisson distribution)
```{r}
glm_sma_75 <- glm(Count ~  Category, data = plot_data_sma75, 
                      family = "poisson")
summary(glm_sma_75)
# No correlation
```


# *Suzano!*
## Read files
```{r}
suzano_akde95 <- st_read("suzano_akde95.gpkg")
plot(suzano_akde95)
# convert a polygon geometry to a data frame
suzano_akde95 <- st_coordinates(suzano_akde95)
# Convert to data frame
suzano_akde95 <- as.data.frame(suzano_akde95)
head(suzano_akde95)
# Rename the columns
colnames(suzano_akde95) <- c("x", "y")
head(suzano_akde95)
suzano_akde70 <- st_read("suzano_akde70.gpkg")
plot(suzano_akde70)
# convert a polygon geometry to a data frame
suzano_akde70 <- st_coordinates(suzano_akde70)
# Convert to data frame
suzano_akde70 <- as.data.frame(suzano_akde70)
head(suzano_akde70)
# Rename the columns
colnames(suzano_akde70) <- c("x", "y")
head(suzano_akde70)
df_cp_suzano <- read.csv("CP_distanceborder_suzano.csv")
head(df_cp_suzano)
```

## Filter lines just to have locomotion cps (the ones I want to calculate distance to the borders)
```{r}
df_cp_suzano <- df_cp_suzano[df_cp_suzano$cp_q6 == "Locomotion", ]
head(df_cp_suzano)
```

## Plot the data to see the distribution of Locomomotion CPs
```{r, warning=FALSE}
ggplot()+
  geom_polygon(data = suzano_akde95, aes(x=x, y= y),
               fill = "palegreen4", size = 1)+
  geom_polygon(data = suzano_akde70, aes(x=x, y= y),
               fill = "yellowgreen", colour = "black") +
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, size=2))+
  geom_point(data = df_cp_suzano, aes(x = longitude_x_GAL, 
                                      y = latitude_y_GAL), size = 1) +
  coord_fixed()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=2),
        plot.margin=unit(c(0.1,0,0,0),"cm"))+
  theme(axis.text.x = element_text(hjust=0.5, size = 8, colour ="Black"))+
  theme(axis.text.y = element_text(angle=90, hjust=0.5, 
                                   size = 8, colour ="Black"))+
  theme(axis.title.y = element_blank())+
  theme(axis.title.x = element_blank())+
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  annotation_scale(location = "tr", width_hint = 0.2, 
                   pad_x = unit(0.1, "in"), pad_y = unit(0.8, "in"))+
  annotation_north_arrow(location = "tr", which_north = "true", 
                         pad_x = unit(0.3, "in"), pad_y = unit(0.1, "in"),
                         style = north_arrow_fancy_orienteering)
```

## Some more data organization
```{r}
colnames(df_cp_suzano) <- c("date", "time", "y", "x", "behav", "cp_q6")
suzano_akde95[,3:4] <- NULL; head(suzano_akde95)
suzano_akde70[,3:4] <- NULL; head(suzano_akde70)
```

## Convert data frames to sf objects
```{r}
# Convert suzano_akde95 data frame to an sf object
suzano_akde95_sf <- st_as_sf(suzano_akde95, coords = c("x", "y"), crs = crs)

# Convert suzano_akde70 data frame to an sf object
suzano_akde70_sf <- st_as_sf(suzano_akde70, coords = c("x", "y"), crs = crs)

# Convert df_cp_suzano data frame to an sf object
df_cp_suzano_sf <- st_as_sf(df_cp_suzano, coords = c("x", "y"), crs = crs)
```

## Use the st_nn function (package nngeo) to calculate the distance between the locomotion CP and the nearest point in the borders. 
## First 95% AKDE border

```{r}
# k = 1 for finding 1 nearest neighbor
distances_95 <- nngeo::st_nn(df_cp_suzano_sf, suzano_akde95_sf, k = 1, returnDist = T)
head(distances_95)
```

## Than to the 70% AKDE border
```{r}
distances_70 <- nngeo::st_nn(df_cp_suzano_sf, suzano_akde70_sf, k = 1, returnDist = T)
head(distances_70)
```

## Extract the distance values from the list
```{r}
distance_values_95 <- distances_95$dist
distance_values_70 <- distances_70$dist
```

## Add the distance values as a new column in df_cp_suzano_sf
```{r}
df_cp_suzano$dist70 <- distance_values_70
df_cp_suzano$dist95 <- distance_values_95
head(df_cp_suzano)
df_cp_suzano$dist95 <- as.numeric(df_cp_suzano$dist95)
df_cp_suzano$dist70 <- as.numeric(df_cp_suzano$dist70)
```

## Now, for the plots, I will classify the distances into categories so I can compare frequency proportions of locomotion CPs between distance categories!

## Create the custom breaks for categories -> 95% AKDE
```{r}
breaks <- c(0, 20, 40, 60, 80, 100)
```

## Create the distance categories
```{r}
dist_categories_suzano <- cut(df_cp_suzano$dist95, breaks = breaks, 
                              labels = FALSE, include.lowest = TRUE, 
                              right = FALSE)
```

## Rename the category labels
```{r}
category_labels <- paste0("cat", 1:length(breaks[-1]))
```

## Assign category labels to the distance categories
```{r}
dist_categories_suzano <- factor(dist_categories_suzano, 
                                 levels = 1:length(breaks[-1]), 
                                 labels = category_labels)
df_cp_suzano$cat_dist_95 <- dist_categories_suzano
df_cp_suzano
```

## Compute the frequency of points in each category
```{r}
freq_table_suzano95 <- table(dist_categories_suzano)
```

## Compute the proportion of location points for each distance category
```{r}
proportions_suzano95 <- freq_table_suzano95 / sum(freq_table_suzano95)

plot_data_suzano95 <- data.frame(Proportion = proportions_suzano95,
                                 Count = freq_table_suzano95)
```

## More data organization
```{r}
# Add row names as a separate column
plot_data_suzano95$Category <- row.names(plot_data_suzano95)
# Reset row names
row.names(plot_data_suzano95) <- NULL
plot_data_suzano95$Proportion.dist_categories_suzano <- NULL
plot_data_suzano95$Count.dist_categories_suzano <- NULL
# Change the column names of the data frame
colnames(plot_data_suzano95) <- c("Proportion", "Count", "Category")
plot_data_suzano95 <- plot_data_suzano95[, c("Category", "Count", "Proportion")]
plot_data_suzano95$Category <- as.numeric(plot_data_suzano95$Category)
# Convert Category column to numeric, with the distances they represent
plot_data_suzano95$Category <- as.numeric(c(0, 20, 40, 60, 80))
plot_data_suzano95
```

## Plot the proportion of location points by distance category
```{r}
ggplot(plot_data_suzano95, aes(x = Category, y = Proportion, group = 1)) +
  geom_line() +
  theme_bw() +
  xlab("Distance to AKDE 95% border (m)") +
  ylab("Proportion of Locomotion CPs") +
  scale_y_continuous(expand = c(0.05, 0)) +
  scale_x_continuous(breaks = seq(0, 740, by = 20))+
  ggtitle("Suzano")
```

## I will try some tests to see if there is any relationship
## Kolmogorov-Smirnov test
```{r}
ks.test(df_cp_suzano$dist95, "punif", min = 0, max = max(df_cp_suzano$dist95))
# the test almost shows significance. 
```

## Generealized Linear model (poisson distribution)
```{r}
glm_suzano_95 <- glm(Count ~  Category, data = plot_data_suzano95, 
                      family = "poisson")
summary(glm_suzano_95)
# No correlation
```

# *Suzano - 70% AKDE - Core area*
## Create the custom breaks for categories 95%
```{r}
breaks <- c(0, 20, 40, 60, 80)
```

## Create the distance categories
```{r}
dist_categories_suzano <- cut(df_cp_suzano$dist70, breaks = breaks, 
                              labels = FALSE, include.lowest = TRUE, 
                              right = FALSE)
```
## Rename the category labels
```{r}
category_labels_suzano_70 <- paste0("cat", 1:length(breaks[-1]))
```

## Assign category labels to the distance categories
```{r}
dist_categories_suzano <- category_labels_suzano_70[dist_categories_suzano]
category_labels_suzano_70 <- paste0("cat", 1:length(unique(
  dist_categories_suzano)))
df_cp_suzano$cat_dist_70 <- factor(dist_categories_suzano, 
                                   levels = unique(dist_categories_suzano), 
                                   labels = category_labels_suzano_70)
df_cp_suzano
```

## I do not understand why the code is not working correctly now... So I will create a vector with the category for each point manually and than assign to the data frame.
```{r}
df_cp_suzano$cat_dist_70 <- c("cat2","cat2", "cat3", "cat1",  "cat4", "cat4", 
                              "cat3", "cat4", "cat2", "cat2", "cat4", "cat2", 
                              "cat2", "cat3", "cat1", "cat2", "cat2", "cat2")
```

## Compute the frequency of points in each category
```{r}
freq_table_suzano70 <- table(df_cp_suzano$cat_dist_70)
```

## Compute the proportion of location points for each distance category
```{r}
proportions_suzano70 <- freq_table_suzano70 / sum(freq_table_suzano70)
plot_data_suzano70 <- data.frame(Proportion = proportions_suzano70,
                                 Count = freq_table_suzano70)
```

## More data organization
```{r}
## Add row names as a separate column
plot_data_suzano70$Category <- row.names(plot_data_suzano70)
# Reset row names
row.names(plot_data_suzano70) <- NULL
plot_data_suzano70$Count.Var1 <- NULL
plot_data_suzano70$Proportion.Var1 <- NULL
# Change the column names of the data frame
colnames(plot_data_suzano70) <- c("Proportion", "Count", "Category")
plot_data_suzano70 <- plot_data_suzano70[, c("Category", "Count", "Proportion")]
plot_data_suzano70$Category <- as.numeric(plot_data_suzano70$Category)
# Convert Category column to numeric, with the distances they represent
plot_data_suzano70$Category <- as.numeric(c(0, 20, 40, 60))
plot_data_suzano70
```

## Plot the proportion of location points by distance category
```{r}
ggplot(plot_data_suzano70, aes(x = Category, y = Proportion, group = 1)) +
  geom_line() +
  theme_bw() +
  xlab("Distance to AKDE 70% border (m)") +
  ylab("Proportion of Locomotion CPs") +
  scale_y_continuous(expand = c(0.05, 0)) +
  scale_x_continuous(breaks = seq(0, 80, by = 20))+
  ggtitle("Suzano")
```

## I will try some tests to see if there is any relationship
## Kolmogorov-Smirnov test
```{r}
ks.test(df_cp_suzano$dist70, "punif", min = 0, max = max(df_cp_suzano$dist70))
# the test shows no significance. 
```

## Generealized Linear model (poisson distribution)
```{r}
glm_suzano_70 <- glm(Count ~  Category, data = plot_data_suzano70, 
                      family = "poisson")
summary(glm_suzano_70)
# No correlation
```

## I will try another kind of plot to see the trend
```{r}
plot=ggplot(plot_data_suzano70, aes(y=Proportion , x= Category))+ 
  geom_point()+
  geom_smooth(method = "glm", formula = y ~ x)+
  xlab("Distance to AKDE 70% border (m)") +
  ylab("Proportion of Locomotion CPs") 
plot + theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(), 
              panel.background = element_blank(), 
              axis.line = element_line(colour = "black"))
```
### Does not look as expected, indeed :(

*** 

# I will try a last test for Suzano. 
## Since the edges of the forest in Suzano are more restrictive than in the other areas, I will test if the Locomotion CPs are more frequent close to te edges. I considered as an edge the frontier between 1) forest and river; and 2) forest and matrix. 

## Read files
```{r}
edge <- st_read("edge_suzano.gpkg")
plot(edge)
# convert a polygon geometry to a data frame
edge <- st_coordinates(edge)
# Convert to data frame
edge <- as.data.frame(edge)
head(edge)
# Rename the columns
colnames(edge) <- c("x", "y", "lineid")
head(edge)
```

## Convert data frames to sf objects
```{r}
# Convert suzano_akde95 data frame to an sf object
edge_sf <- st_as_sf(edge, coords = c("x", "y"), crs = crs)
```

## Use the st_nn function (package nngeo) to calculate the distance between the locomotion CP and the nearest point in the edges. 
```{r}
# k = 1 for finding 1 nearest neighbor
distances_edge <- nngeo::st_nn(df_cp_suzano_sf, edge_sf, k = 1, returnDist = T)
head(distances_edge)
```

## Extract the distance values from the list
```{r}
distances_edge <- distances_edge$dist
```

## Add the distance values as a new column in df_cp_suzano_sf
```{r}
df_cp_suzano$edge <- distances_edge
head(df_cp_suzano)
df_cp_suzano$edge <- as.numeric(df_cp_suzano$edge)
```

## Now, for the plots, I will classify the distances into categories so I can compare frequency proportions of locomotion CPs between distance categories!

## Create the custom breaks for categories 
```{r}
summary(df_cp_suzano$edge)
breaks <- c(0, 20, 40, 60, 80, 100, 120)
```

## Create the distance categories
```{r}
dist_categories_suzano <- cut(df_cp_suzano$edge, breaks = breaks, 
                              labels = FALSE, include.lowest = TRUE, 
                              right = FALSE)
```

## Rename the category labels
```{r}
category_labels <- paste0("cat", 1:length(breaks[-1]))
```

## Assign category labels to the distance categories
```{r}
dist_categories_suzano <- factor(dist_categories_suzano, 
                                 levels = 1:length(breaks[-1]), 
                                 labels = category_labels)
df_cp_suzano$cat_edge <- dist_categories_suzano
df_cp_suzano
```

## Compute the frequency of points in each category
```{r}
freq_table_suzano_edge <- table(dist_categories_suzano)
```

## Compute the proportion of location points for each distance category
```{r}
proportions_suzano_edge <- freq_table_suzano_edge / sum(freq_table_suzano_edge)

plot_data_suzano_edge <- data.frame(Proportion = proportions_suzano_edge,
                                 Count = freq_table_suzano_edge)
```

## More data organization
```{r}
# Add row names as a separate column
plot_data_suzano_edge$Category <- row.names(plot_data_suzano_edge)
# Reset row names
row.names(plot_data_suzano_edge) <- NULL
plot_data_suzano_edge$Proportion.dist_categories_suzano <- NULL
plot_data_suzano_edge$Count.dist_categories_suzano <- NULL
# Change the column names of the data frame
colnames(plot_data_suzano_edge) <- c("Proportion", "Count", "Category")
plot_data_suzano_edge <- plot_data_suzano_edge[, c("Category", "Count", "Proportion")]
plot_data_suzano_edge$Category <- as.numeric(plot_data_suzano_edge$Category)
# Convert Category column to numeric, with the distances they represent
plot_data_suzano_edge$Category <- as.numeric(c(0, 20, 40, 60, 80,100))
plot_data_suzano_edge
```

## Plot the proportion of location points by distance category
```{r}
ggplot(plot_data_suzano_edge, aes(x = Category, y = Proportion, group = 1)) +
  geom_line() +
  theme_bw() +
  xlab("Distance to edges (m)") +
  ylab("Proportion of Locomotion CPs") +
  #scale_y_continuous(expand = c(0.05, 0)) +
  #scale_x_continuous(breaks = seq(0, 120, by = 20))+
  ggtitle("Suzano")
```

## I will try some tests to see if there is any relationship
## Kolmogorov-Smirnov test
```{r}
ks.test(df_cp_suzano$edge, "punif", min = 0, max = max(df_cp_suzano$edge))
# not significative. 
```

## Generealized Linear model (poisson distribution)
```{r}
glm_suzano_edge <- glm(Count ~  Category, data = plot_data_suzano_edge, 
                      family = "poisson")
summary(glm_suzano_edge)
# No correlation
```

I will try another kind of plot to see the trend
```{r}
plot=ggplot(plot_data_suzano_edge, aes(y=Proportion , x= Category))+
  geom_point()+
  geom_smooth(method = "glm", formula = y ~ x)+
  xlab("Distance to edges (m)") +
  ylab("Proportion of Locomotion CPs")
plot + theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              panel.background = element_blank(),
              axis.line = element_line(colour = "black"))
```

## **I can see that there is no significative relationchip between edeges and Locomotion CPs, despite a small trend to increase frequency of CPs with increases in distance to edges.**