---
title: "Core areas"
author: "Felipe Bufalo"
date: "2023-05-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Script to estimate the core areas from AKDEc home range estimations.
## After estimating the home ranges, I wanted to follow Samuel & Garton 1985 method to get to the core area without choosing in an arbitrary way. The AKDEs were estimated using the shape files of the areas as hard boundaries.

### I had a lot of dificulties to get to the results (core areas). I could not estimate the isopleths without someonelse's assistance with ctmm package (dealing with UD - AKDE objects). So, I performed the most mechanical way possible, extracting manually each home range area estimated by the model for each area and then created the data frames to work with. 

## It is ugly, but it works!

### Load packages
```{r}
library(ctmm)
library(sf)
```

# Starting with Guareí
```{r}
load("AKDE_guarei.rda")
```

## Take a look at the model
```{r}
summary(UDakdedf_ML)
```

## I will plot the akde home range estimation so I can see how it looks:
```{r}
plot(UDakdedf_ML) # Looks good!
```

## I will now plot all isopleths (5% by 5% - from 1% home range estimation to 99%). It was the only way I coul manage this. Basically, I create a shapefile for each isopleth estimation and then extract the area. This allowed me to get the areas estimated in each isopleth.

```{r}
shapefile_polygons99 <- as.sf(UDakdedf_ML, level.UD=0.99, level=0.95)
middle_polygon99 <- shapefile_polygons99[2,]
area_m2_99 <-st_area(middle_polygon99)
area_m2_99/10000

shapefile_polygons95 <- as.sf(UDakdedf_ML, level.UD=0.95, level=0.95)
middle_polygon95 <- shapefile_polygons95[2,]
area_m2_95 <-st_area(middle_polygon95)
area_m2_95/10000

shapefile_polygons90 <- as.sf(UDakdedf_ML, level.UD=0.90, level=0.95)
middle_polygon90 <- shapefile_polygons90[2,]
area_m2_90 <-st_area(middle_polygon90)
area_m2_90/10000

shapefile_polygons85 <- as.sf(UDakdedf_ML, level.UD=0.85, level=0.95)
middle_polygon85 <- shapefile_polygons85[2,]
area_m2_85 <-st_area(middle_polygon85)
area_m2_85/10000

shapefile_polygons80 <- as.sf(UDakdedf_ML, level.UD=0.80, level=0.95)
middle_polygon80 <- shapefile_polygons80[2,]
area_m2_80 <-st_area(middle_polygon80)
area_m2_80/10000

shapefile_polygons75 <- as.sf(UDakdedf_ML, level.UD=0.75, level=0.95)
middle_polygon75 <- shapefile_polygons75[2,]
area_m2_75 <-st_area(middle_polygon75)
area_m2_75/10000

shapefile_polygons70 <- as.sf(UDakdedf_ML, level.UD=0.70, level=0.95)
middle_polygon70 <- shapefile_polygons70[2,]
area_m2_70 <-st_area(middle_polygon70)
area_m2_70/10000

shapefile_polygons65 <- as.sf(UDakdedf_ML, level.UD=0.65, level=0.95)
middle_polygon65 <- shapefile_polygons65[2,]
area_m2_65 <-st_area(middle_polygon65)
area_m2_65/10000

shapefile_polygons60 <- as.sf(UDakdedf_ML, level.UD=0.60, level=0.95)
middle_polygon60 <- shapefile_polygons60[2,]
area_m2_60 <-st_area(middle_polygon60)
area_m2_60/10000

shapefile_polygons55 <- as.sf(UDakdedf_ML, level.UD=0.55, level=0.95)
middle_polygon55 <- shapefile_polygons55[2,]
area_m2_55 <-st_area(middle_polygon55)
area_m2_55/10000

shapefile_polygons50 <- as.sf(UDakdedf_ML, level.UD=0.50, level=0.95)
middle_polygon50 <- shapefile_polygons50[2,]
area_m2_50 <-st_area(middle_polygon50)
area_m2_50/10000

shapefile_polygons45 <- as.sf(UDakdedf_ML, level.UD=0.45, level=0.95)
middle_polygon45 <- shapefile_polygons45[2,]
area_m2_45 <-st_area(middle_polygon45)
area_m2_45/10000

shapefile_polygons40 <- as.sf(UDakdedf_ML, level.UD=0.40, level=0.95)
middle_polygon40 <- shapefile_polygons40[2,]
area_m2_40 <-st_area(middle_polygon40)
area_m2_40/10000

shapefile_polygons35 <- as.sf(UDakdedf_ML, level.UD=0.35, level=0.95)
middle_polygon35 <- shapefile_polygons35[2,]
area_m2_35 <-st_area(middle_polygon35)
area_m2_35/10000

shapefile_polygons30 <- as.sf(UDakdedf_ML, level.UD=0.30, level=0.95)
middle_polygon30 <- shapefile_polygons30[2,]
area_m2_30 <-st_area(middle_polygon30)
area_m2_30/10000

shapefile_polygons25 <- as.sf(UDakdedf_ML, level.UD=0.25, level=0.95)
middle_polygon25 <- shapefile_polygons25[2,]
area_m2_25 <-st_area(middle_polygon25)
area_m2_25/10000

shapefile_polygons20 <- as.sf(UDakdedf_ML, level.UD=0.20, level=0.95)
middle_polygon20 <- shapefile_polygons20[2,]
area_m2_20 <-st_area(middle_polygon20)
area_m2_20/10000

shapefile_polygons15 <- as.sf(UDakdedf_ML, level.UD=0.15, level=0.95)
middle_polygon15 <- shapefile_polygons15[2,]
area_m2_15 <-st_area(middle_polygon15)
area_m2_15/10000

shapefile_polygons10 <- as.sf(UDakdedf_ML, level.UD=0.10, level=0.95)
middle_polygon10 <- shapefile_polygons10[2,]
area_m2_10 <-st_area(middle_polygon10)
area_m2_10/10000

shapefile_polygons05 <- as.sf(UDakdedf_ML, level.UD=0.05, level=0.95)
middle_polygon05 <- shapefile_polygons05[2,]
area_m2_05 <-st_area(middle_polygon05)
area_m2_05/10000

shapefile_polygons01 <- as.sf(UDakdedf_ML, level.UD=0.01, level=0.95)
middle_polygon01 <- shapefile_polygons01[2,]
area_m2_01 <-st_area(middle_polygon01)
area_m2_01/10000
```

## With all areas printed, I manually created the vectors with area and isopleths.

```{r}
area.href <- c(45.87342, 37.96572, 33.5622, 30.26605, 27.43143, 
               24.88615, 22.54775, 20.33878, 18.25306, 16.23031, 14.39388, 
               12.60215, 10.89223, 9.264786, 7.717092, 6.246799, 4.853915, 
               3.541742, 2.290297, 1.101816, 0.197499)

isopleth <- c(99, 95, 90, 85, 80, 75, 70, 65, 60, 55, 50, 45, 40, 35, 30, 25, 20,
              15, 10, 05, 01)
```

## Then, create the data frame 
```{r}
HRsize <- data.frame(area.href = area.href, isopleth = isopleth)
HRsize <- HRsize[order(HRsize$isopleth), ]
head(HRsize)
```

## Calculate the percentage of total home range area represented by each isopleth
```{r}
HRsize$PA <- HRsize$area.href/(max(HRsize$area.href))*100 # add PA
HRsize$IV <- HRsize$isopleth # add IV
```

## Identify the core area as the point of maximum deviation from area/%UD equality line
```{r}
with(HRsize, which((IV-PA) == max(IV-PA)))
```

## Change the value according to the result (IV row number indicated above (15))
```{r}
HRsize$IV[15] # 70% isopleth
```

## Look what is the size of the home range at 70%
```{r}
HRsize # ha 
```

## Plot to see the probability of uniform area use vs empirical area use in Guareí
```{r}
plot(seq(0,100,length = 50), seq(0,100,length = 50), type = "l",lty = 3, 
     lwd=2, xlab="Isopleth (%UD)", ylab="Proportion of total home range area (%)", 
     main="Core area - Guareí")
lines(PA ~ IV, data = HRsize, col = 2, lwd=2)
points(x = 70, y = 49.1521016, type = "p", lwd=2) ## change X according to value 
# identified above and Y to meet the line
```

# Get area estimates and errors:
```{r}
shapefile_polygons70 <- as.sf(UDakdedf_ML, level.UD=0.70, level=0.95)
area_m2_70 <-st_area(shapefile_polygons70)
area_m2_70/10000 # ha already!
```

# Suzano
```{r}
rm(list=ls())
load("AKDE_suzano.rda")
```

## Take a look at the model
```{r}
summary(UDakdedf_suzano_ML)
```

## I will plot the akde home range estimation so I can see how it looks:
```{r}
plot(UDakdedf_suzano_ML) # Looks good!
```

## I will now plot all isopleths (5% by 5% - from 1% home range estimation to 99%). It was the only way I coul manage this. Basically, I create a shapefile for each isopleth estimation and then extract the area. This allowed me to get the areas estimated in each isopleth.

```{r}
shapefile_polygons99 <- as.sf(UDakdedf_suzano_ML, level.UD=0.99, level=0.95)
middle_polygon99 <- shapefile_polygons99[2,]
area_m2_99 <-st_area(middle_polygon99)
area_m2_99/10000

shapefile_polygons95 <- as.sf(UDakdedf_suzano_ML, level.UD=0.95, level=0.95)
middle_polygon95 <- shapefile_polygons95[2,]
area_m2_95 <-st_area(middle_polygon95)
area_m2_95/10000

shapefile_polygons90 <- as.sf(UDakdedf_suzano_ML, level.UD=0.90, level=0.95)
middle_polygon90 <- shapefile_polygons90[2,]
area_m2_90 <-st_area(middle_polygon90)
area_m2_90/10000

shapefile_polygons85 <- as.sf(UDakdedf_suzano_ML, level.UD=0.85, level=0.95)
middle_polygon85 <- shapefile_polygons85[2,]
area_m2_85 <-st_area(middle_polygon85)
area_m2_85/10000

shapefile_polygons80 <- as.sf(UDakdedf_suzano_ML, level.UD=0.80, level=0.95)
middle_polygon80 <- shapefile_polygons80[2,]
area_m2_80 <-st_area(middle_polygon80)
area_m2_80/10000

shapefile_polygons75 <- as.sf(UDakdedf_suzano_ML, level.UD=0.75, level=0.95)
middle_polygon75 <- shapefile_polygons75[2,]
area_m2_75 <-st_area(middle_polygon75)
area_m2_75/10000

shapefile_polygons70 <- as.sf(UDakdedf_suzano_ML, level.UD=0.70, level=0.95)
middle_polygon70 <- shapefile_polygons70[2,]
area_m2_70 <-st_area(middle_polygon70)
area_m2_70/10000

shapefile_polygons65 <- as.sf(UDakdedf_suzano_ML, level.UD=0.65, level=0.95)
middle_polygon65 <- shapefile_polygons65[2,]
area_m2_65 <-st_area(middle_polygon65)
area_m2_65/10000

shapefile_polygons60 <- as.sf(UDakdedf_suzano_ML, level.UD=0.60, level=0.95)
middle_polygon60 <- shapefile_polygons60[2,]
area_m2_60 <-st_area(middle_polygon60)
area_m2_60/10000

shapefile_polygons55 <- as.sf(UDakdedf_suzano_ML, level.UD=0.55, level=0.95)
middle_polygon55 <- shapefile_polygons55[2,]
area_m2_55 <-st_area(middle_polygon55)
area_m2_55/10000

shapefile_polygons50 <- as.sf(UDakdedf_suzano_ML, level.UD=0.50, level=0.95)
middle_polygon50 <- shapefile_polygons50[2,]
area_m2_50 <-st_area(middle_polygon50)
area_m2_50/10000

shapefile_polygons45 <- as.sf(UDakdedf_suzano_ML, level.UD=0.45, level=0.95)
middle_polygon45 <- shapefile_polygons45[2,]
area_m2_45 <-st_area(middle_polygon45)
area_m2_45/10000

shapefile_polygons40 <- as.sf(UDakdedf_suzano_ML, level.UD=0.40, level=0.95)
middle_polygon40 <- shapefile_polygons40[2,]
area_m2_40 <-st_area(middle_polygon40)
area_m2_40/10000

shapefile_polygons35 <- as.sf(UDakdedf_suzano_ML, level.UD=0.35, level=0.95)
middle_polygon35 <- shapefile_polygons35[2,]
area_m2_35 <-st_area(middle_polygon35)
area_m2_35/10000

shapefile_polygons30 <- as.sf(UDakdedf_suzano_ML, level.UD=0.30, level=0.95)
middle_polygon30 <- shapefile_polygons30[2,]
area_m2_30 <-st_area(middle_polygon30)
area_m2_30/10000

shapefile_polygons25 <- as.sf(UDakdedf_suzano_ML, level.UD=0.25, level=0.95)
middle_polygon25 <- shapefile_polygons25[2,]
area_m2_25 <-st_area(middle_polygon25)
area_m2_25/10000

shapefile_polygons20 <- as.sf(UDakdedf_suzano_ML, level.UD=0.20, level=0.95)
middle_polygon20 <- shapefile_polygons20[2,]
area_m2_20 <-st_area(middle_polygon20)
area_m2_20/10000

shapefile_polygons15 <- as.sf(UDakdedf_suzano_ML, level.UD=0.15, level=0.95)
middle_polygon15 <- shapefile_polygons15[2,]
area_m2_15 <-st_area(middle_polygon15)
area_m2_15/10000

shapefile_polygons10 <- as.sf(UDakdedf_suzano_ML, level.UD=0.10, level=0.95)
middle_polygon10 <- shapefile_polygons10[2,]
area_m2_10 <-st_area(middle_polygon10)
area_m2_10/10000

shapefile_polygons05 <- as.sf(UDakdedf_suzano_ML, level.UD=0.05, level=0.95)
middle_polygon05 <- shapefile_polygons05[2,]
area_m2_05 <-st_area(middle_polygon05)
area_m2_05/10000

shapefile_polygons01 <- as.sf(UDakdedf_suzano_ML, level.UD=0.01, level=0.95)
middle_polygon01 <- shapefile_polygons01[2,]
area_m2_01 <-st_area(middle_polygon01)
area_m2_01/10000
```

## With all areas printed, I manually created the vectors with area and isopleths.

```{r}
area.href <- c(0.06404234, 0.4003905, 0.8198001, 1.26308, 1.740751, 2.246062, 
               2.780916, 3.342436, 3.948553, 4.613612, 5.352505, 6.179061, 
               7.104065, 8.145396, 9.330773, 10.69629, 12.27685, 14.19957, 
               16.50785, 19.41044, 23.79421)


isopleth <- c(01, 05, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 
              80, 85, 90, 95, 99)
```

## Then, create the data frame 
```{r}
HRsize <- data.frame(area.href = area.href, isopleth = isopleth)
head(HRsize) 
```

## Calculate the percentage of total home range area represented by each isopleth
```{r}
HRsize$PA <- HRsize$area.href/(max(HRsize$area.href))*100 # add PA
HRsize$IV <- HRsize$isopleth # add IV
```

## Identify the core area as the point of maximum deviation from area/%UD equality line
```{r}
with(HRsize, which((IV-PA) == max(IV-PA)))
```

## Change the value according to the result (IV row number indicated above (15))
```{r}
HRsize$IV[15] # 70% isopleth
```

## Look what is the size of the home range at 70%
```{r}
HRsize #11.16579000 ha 
```

## Plot to see the probability of uniform area use vs empirical area use in Suzano
```{r}
plot(seq(0,100,length = 50), seq(0,100,length = 50), type = "l",lty = 3, 
     lwd=2, xlab="Isopleth (%UD)", ylab="Proportion of total home range area (%)", 
     main="Core area - Suzano")
lines(PA ~ IV, data = HRsize, col = 2, lwd=2)
points(x = 70, y = 39.2144686, type = "p", lwd=2) ## change X according to value 
# identified above and Y to meet the line
```

# Extract area in ha and errors:
```{r}
shapefile_polygons70 <- as.sf(UDakdedf_suzano_ML, level.UD=0.70, level=0.95)
area_m2_70 <-st_area(shapefile_polygons70)
area_m2_70/10000 # ha already!
```

# Taquara
```{r}
rm(list=ls())
load("AKDE_taquara.rda")
```

## Take a look at the model
```{r}
summary(UDakdedf_taquara_ML)
```

## I will plot the akde home range estimation so I can see how it looks:
```{r}
plot(UDakdedf_taquara_ML) # Looks good!
```

## I will now plot all isopleths (5% by 5% - from 1% home range estimation to 99%). It was the only way I coul manage this. Basically, I create a shapefile for each isopleth estimation and then extract the area. This allowed me to get the areas estimated in each isopleth.

```{r}
shapefile_polygons99 <- as.sf(UDakdedf_taquara_ML, level.UD=0.99, level=0.95)
middle_polygon99 <- shapefile_polygons99[2,]
area_m2_99 <-st_area(middle_polygon99)
area_m2_99/10000

shapefile_polygons95 <- as.sf(UDakdedf_taquara_ML, level.UD=0.95, level=0.95)
middle_polygon95 <- shapefile_polygons95[2,]
area_m2_95 <-st_area(middle_polygon95)
area_m2_95/10000

shapefile_polygons90 <- as.sf(UDakdedf_taquara_ML, level.UD=0.90, level=0.95)
middle_polygon90 <- shapefile_polygons90[2,]
area_m2_90 <-st_area(middle_polygon90)
area_m2_90/10000

shapefile_polygons85 <- as.sf(UDakdedf_taquara_ML, level.UD=0.85, level=0.95)
middle_polygon85 <- shapefile_polygons85[2,]
area_m2_85 <-st_area(middle_polygon85)
area_m2_85/10000

shapefile_polygons80 <- as.sf(UDakdedf_taquara_ML, level.UD=0.80, level=0.95)
middle_polygon80 <- shapefile_polygons80[2,]
area_m2_80 <-st_area(middle_polygon80)
area_m2_80/10000

shapefile_polygons75 <- as.sf(UDakdedf_taquara_ML, level.UD=0.75, level=0.95)
middle_polygon75 <- shapefile_polygons75[2,]
area_m2_75 <-st_area(middle_polygon75)
area_m2_75/10000

shapefile_polygons70 <- as.sf(UDakdedf_taquara_ML, level.UD=0.70, level=0.95)
middle_polygon70 <- shapefile_polygons70[2,]
area_m2_70 <-st_area(middle_polygon70)
area_m2_70/10000

shapefile_polygons65 <- as.sf(UDakdedf_taquara_ML, level.UD=0.65, level=0.95)
middle_polygon65 <- shapefile_polygons65[2,]
area_m2_65 <-st_area(middle_polygon65)
area_m2_65/10000

shapefile_polygons60 <- as.sf(UDakdedf_taquara_ML, level.UD=0.60, level=0.95)
middle_polygon60 <- shapefile_polygons60[2,]
area_m2_60 <-st_area(middle_polygon60)
area_m2_60/10000

shapefile_polygons55 <- as.sf(UDakdedf_taquara_ML, level.UD=0.55, level=0.95)
middle_polygon55 <- shapefile_polygons55[2,]
area_m2_55 <-st_area(middle_polygon55)
area_m2_55/10000

shapefile_polygons50 <- as.sf(UDakdedf_taquara_ML, level.UD=0.50, level=0.95)
middle_polygon50 <- shapefile_polygons50[2,]
area_m2_50 <-st_area(middle_polygon50)
area_m2_50/10000

shapefile_polygons45 <- as.sf(UDakdedf_taquara_ML, level.UD=0.45, level=0.95)
middle_polygon45 <- shapefile_polygons45[2,]
area_m2_45 <-st_area(middle_polygon45)
area_m2_45/10000

shapefile_polygons40 <- as.sf(UDakdedf_taquara_ML, level.UD=0.40, level=0.95)
middle_polygon40 <- shapefile_polygons40[2,]
area_m2_40 <-st_area(middle_polygon40)
area_m2_40/10000

shapefile_polygons35 <- as.sf(UDakdedf_taquara_ML, level.UD=0.35, level=0.95)
middle_polygon35 <- shapefile_polygons35[2,]
area_m2_35 <-st_area(middle_polygon35)
area_m2_35/10000

shapefile_polygons30 <- as.sf(UDakdedf_taquara_ML, level.UD=0.30, level=0.95)
middle_polygon30 <- shapefile_polygons30[2,]
area_m2_30 <-st_area(middle_polygon30)
area_m2_30/10000

shapefile_polygons25 <- as.sf(UDakdedf_taquara_ML, level.UD=0.25, level=0.95)
middle_polygon25 <- shapefile_polygons25[2,]
area_m2_25 <-st_area(middle_polygon25)
area_m2_25/10000

shapefile_polygons20 <- as.sf(UDakdedf_taquara_ML, level.UD=0.20, level=0.95)
middle_polygon20 <- shapefile_polygons20[2,]
area_m2_20 <-st_area(middle_polygon20)
area_m2_20/10000

shapefile_polygons15 <- as.sf(UDakdedf_taquara_ML, level.UD=0.15, level=0.95)
middle_polygon15 <- shapefile_polygons15[2,]
area_m2_15 <-st_area(middle_polygon15)
area_m2_15/10000

shapefile_polygons10 <- as.sf(UDakdedf_taquara_ML, level.UD=0.10, level=0.95)
middle_polygon10 <- shapefile_polygons10[2,]
area_m2_10 <-st_area(middle_polygon10)
area_m2_10/10000

shapefile_polygons05 <- as.sf(UDakdedf_taquara_ML, level.UD=0.05, level=0.95)
middle_polygon05 <- shapefile_polygons05[2,]
area_m2_05 <-st_area(middle_polygon05)
area_m2_05/10000

shapefile_polygons01 <- as.sf(UDakdedf_taquara_ML, level.UD=0.01, level=0.95)
middle_polygon01 <- shapefile_polygons01[2,]
area_m2_01 <-st_area(middle_polygon01)
area_m2_01/10000
```

## With all areas printed, I manually created the vectors with area and isopleths.

```{r}
area.href <- c(1.274447, 6.983761, 14.30958, 21.86846, 29.75736, 37.9813, 
               46.65597, 55.83595, 65.6059, 76.09046, 87.44227, 99.81765, 
               113.4802, 128.7344, 146.1137, 166.3771, 190.7048, 221.2212, 
               262.1181, 325.7782, 453.4343)


isopleth <- c(01, 05, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 
              80, 85, 90, 95, 99)
```

## Then, create the data frame 
```{r}
HRsize <- data.frame(area.href = area.href, isopleth = isopleth)
head(HRsize) 
```

## Calculate the percentage of total home range area represented by each isopleth
```{r}
HRsize$PA <- HRsize$area.href/(max(HRsize$area.href))*100 # add PA
HRsize$IV <- HRsize$isopleth # add IV
```

## Identify the core area as the point of maximum deviation from area/%UD equality line
```{r}
with(HRsize, which((IV-PA) == max(IV-PA)))
```

## Change the value according to the result (IV row number indicated above (15))
```{r}
HRsize$IV[16] # 75% isopleth!
```

## Look what is the size of the home range at 75%
```{r}
HRsize # 166.377100ha 
```

## Plot to see the probability of uniform area use vs empirical area use in taquara
```{r}
plot(seq(0,100,length = 50), seq(0,100,length = 50), type = "l",lty = 3, 
     lwd=2, xlab="Isopleth (%UD)", ylab="Proportion of total home range area (%)", 
     main="Core area - taquara")
lines(PA ~ IV, data = HRsize, col = 2, lwd=2)
points(x = 75, y = 36.6926587, type = "p", lwd=2) ## change X according to value 
# identified above and Y to meet the line
```

# Extract area in ha and errors:
```{r}
shapefile_polygons75 <- as.sf(UDakdedf_taquara_ML, level.UD=0.75, level=0.95)
area_m2_75 <-st_area(shapefile_polygons75)
area_m2_75/10000 # ha already!
```

# Santa Maria
```{r}
rm(list=ls())
load("AKDE_sma.rda")
```

## Take a look at the model
```{r}
summary(UDakdedf_sma_ML)
```

## I will plot the akde home range estimation so I can see how it looks:
```{r}
plot(UDakdedf_sma_ML) # Looks good!
```

## I will now plot all isopleths (5% by 5% - from 1% home range estimation to 99%). It was the only way I coul manage this. Basically, I create a shapefile for each isopleth estimation and then extract the area. This allowed me to get the areas estimated in each isopleth.

```{r}
shapefile_polygons99 <- as.sf(UDakdedf_sma_ML, level.UD=0.99, level=0.95)
middle_polygon99 <- shapefile_polygons99[2,]
area_m2_99 <-st_area(middle_polygon99)
area_m2_99/10000

shapefile_polygons95 <- as.sf(UDakdedf_sma_ML, level.UD=0.95, level=0.95)
middle_polygon95 <- shapefile_polygons95[2,]
area_m2_95 <-st_area(middle_polygon95)
area_m2_95/10000

shapefile_polygons90 <- as.sf(UDakdedf_sma_ML, level.UD=0.90, level=0.95)
middle_polygon90 <- shapefile_polygons90[2,]
area_m2_90 <-st_area(middle_polygon90)
area_m2_90/10000

shapefile_polygons85 <- as.sf(UDakdedf_sma_ML, level.UD=0.85, level=0.95)
middle_polygon85 <- shapefile_polygons85[2,]
area_m2_85 <-st_area(middle_polygon85)
area_m2_85/10000

shapefile_polygons80 <- as.sf(UDakdedf_sma_ML, level.UD=0.80, level=0.95)
middle_polygon80 <- shapefile_polygons80[2,]
area_m2_80 <-st_area(middle_polygon80)
area_m2_80/10000

shapefile_polygons75 <- as.sf(UDakdedf_sma_ML, level.UD=0.75, level=0.95)
middle_polygon75 <- shapefile_polygons75[2,]
area_m2_75 <-st_area(middle_polygon75)
area_m2_75/10000

shapefile_polygons70 <- as.sf(UDakdedf_sma_ML, level.UD=0.70, level=0.95)
middle_polygon70 <- shapefile_polygons70[2,]
area_m2_70 <-st_area(middle_polygon70)
area_m2_70/10000

shapefile_polygons65 <- as.sf(UDakdedf_sma_ML, level.UD=0.65, level=0.95)
middle_polygon65 <- shapefile_polygons65[2,]
area_m2_65 <-st_area(middle_polygon65)
area_m2_65/10000

shapefile_polygons60 <- as.sf(UDakdedf_sma_ML, level.UD=0.60, level=0.95)
middle_polygon60 <- shapefile_polygons60[2,]
area_m2_60 <-st_area(middle_polygon60)
area_m2_60/10000

shapefile_polygons55 <- as.sf(UDakdedf_sma_ML, level.UD=0.55, level=0.95)
middle_polygon55 <- shapefile_polygons55[2,]
area_m2_55 <-st_area(middle_polygon55)
area_m2_55/10000

shapefile_polygons50 <- as.sf(UDakdedf_sma_ML, level.UD=0.50, level=0.95)
middle_polygon50 <- shapefile_polygons50[2,]
area_m2_50 <-st_area(middle_polygon50)
area_m2_50/10000

shapefile_polygons45 <- as.sf(UDakdedf_sma_ML, level.UD=0.45, level=0.95)
middle_polygon45 <- shapefile_polygons45[2,]
area_m2_45 <-st_area(middle_polygon45)
area_m2_45/10000

shapefile_polygons40 <- as.sf(UDakdedf_sma_ML, level.UD=0.40, level=0.95)
middle_polygon40 <- shapefile_polygons40[2,]
area_m2_40 <-st_area(middle_polygon40)
area_m2_40/10000

shapefile_polygons35 <- as.sf(UDakdedf_sma_ML, level.UD=0.35, level=0.95)
middle_polygon35 <- shapefile_polygons35[2,]
area_m2_35 <-st_area(middle_polygon35)
area_m2_35/10000

shapefile_polygons30 <- as.sf(UDakdedf_sma_ML, level.UD=0.30, level=0.95)
middle_polygon30 <- shapefile_polygons30[2,]
area_m2_30 <-st_area(middle_polygon30)
area_m2_30/10000

shapefile_polygons25 <- as.sf(UDakdedf_sma_ML, level.UD=0.25, level=0.95)
middle_polygon25 <- shapefile_polygons25[2,]
area_m2_25 <-st_area(middle_polygon25)
area_m2_25/10000

shapefile_polygons20 <- as.sf(UDakdedf_sma_ML, level.UD=0.20, level=0.95)
middle_polygon20 <- shapefile_polygons20[2,]
area_m2_20 <-st_area(middle_polygon20)
area_m2_20/10000

shapefile_polygons15 <- as.sf(UDakdedf_sma_ML, level.UD=0.15, level=0.95)
middle_polygon15 <- shapefile_polygons15[2,]
area_m2_15 <-st_area(middle_polygon15)
area_m2_15/10000

shapefile_polygons10 <- as.sf(UDakdedf_sma_ML, level.UD=0.10, level=0.95)
middle_polygon10 <- shapefile_polygons10[2,]
area_m2_10 <-st_area(middle_polygon10)
area_m2_10/10000

shapefile_polygons05 <- as.sf(UDakdedf_sma_ML, level.UD=0.05, level=0.95)
middle_polygon05 <- shapefile_polygons05[2,]
area_m2_05 <-st_area(middle_polygon05)
area_m2_05/10000

shapefile_polygons01 <- as.sf(UDakdedf_sma_ML, level.UD=0.01, level=0.95)
middle_polygon01 <- shapefile_polygons01[2,]
area_m2_01 <-st_area(middle_polygon01)
area_m2_01/10000
```

## With all areas printed, I manually created the vectors with area and isopleths.

```{r}
area.href <- c(0.448188, 2.523605, 5.2199, 8.087039, 11.19635, 14.27741, 
               17.51229, 20.92389, 24.57936, 28.4499, 32.5851, 37.03654, 
               41.75009, 46.80041, 52.39897, 58.57856, 65.59301, 73.90276, 
               84.39851, 100.0692, 128.0064)


isopleth <- c(01, 05, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 
              80, 85, 90, 95, 99)
```

## Then, create the data frame 
```{r}
HRsize <- data.frame(area.href = area.href, isopleth = isopleth)
head(HRsize) 
```

## Calculate the percentage of total home range area represented by each isopleth
```{r}
HRsize$PA <- HRsize$area.href/(max(HRsize$area.href))*100 # add PA
HRsize$IV <- HRsize$isopleth # add IV
```

## Identify the core area as the point of maximum deviation from area/%UD equality line
```{r}
with(HRsize, which((IV-PA) == max(IV-PA)))
```

## Change the value according to the result (IV row number indicated above (15))
```{r}
HRsize$IV[16] # 75% isopleth!
```

## Look what is the size of the home range at 75%
```{r}
HRsize #58.578560 ha 
```

## Plot to see the probability of uniform area use vs empirical area use in sma
```{r}
plot(seq(0,100,length = 50), seq(0,100,length = 50), type = "l",lty = 3, 
     lwd=2, xlab="Isopleth (%UD)", ylab="Proportion of total home range area (%)", 
     main="Core area - Santa Maria")
lines(PA ~ IV, data = HRsize, col = 2, lwd=2)
points(x = 75, y = 45.7622119, type = "p", lwd=2) ## change X according to value 
# identified above and Y to meet the line
```

# Extract area in ha and errors:
```{r}
shapefile_polygons75 <- as.sf(UDakdedf_sma_ML, level.UD=0.75, level=0.95)
area_m2_75 <-st_area(shapefile_polygons75)
area_m2_75/10000 # ha already!
```