---
title: "Change Point Test_Guareí-Taquara-Suzano_FB_2022_07_d01"
author: "Felipe Bufalo"
date: "7/1/2022"
output: pdf_document
---

# I created this script to understand what are the location points to which black lion tamarins (BLTs) are directing their routes to in different areas (Guareí, Taquara and Suzano). For that, in this script, I will run the Change Point Test (CPT). This test, in an objective way, allows the identification of points of direction change (CPs) along each daily route.

## Load the packages _'RColorBrewer'_, _'maptools'_, _'classInt'_, _'tidyverse'_, _'ggplot2'_, _'ggspatial'_.

```{r, include=FALSE}
library(RColorBrewer)
library(sf)
library(dplyr)
library(classInt)# CPT Test
library(tidyverse)
library(ggplot2)
library(ggspatial)
```

**Load data to work with**

```{r}
load("cpt_guarei.rda")

forest_guarei <- st_read("forest_guarei.gpkg")
forest_guarei <- st_coordinates(forest_guarei)
forest_guarei <- as.data.frame(forest_guarei)

k95 <- st_read("guarei_akde95.gpkg")
k95 <- st_coordinates(k95)
k95 <- as.data.frame(k95)

k70 <- st_read("guarei_akde70.gpkg")
k70 <- st_coordinates(k70)
k70 <- as.data.frame(k70)
```

# I adapted the script available at:

BYRNE, R. W.  et al. How did they get here from there? Detecting changes of direction in terrestrial ranging. Animal Behaviour, v. 77, n. 3, p. 619-631, 2009/03/01/ 2009. ISSN 0003-3472. Disponível em: < http://www.sciencedirect.com/science/article/pii/S0003347208005514 >.

# Starting with Guareí!

# First, I need to understand my data and set the best q value for the CPT to find the largest number of CPs. __The q value represents the number of vectors to be considered by the CPT before each possible CP is found.__ For that, I adapted Byrne et al. 2009 to run in loops the CPT for all q values, from 1 to 10. I considered alpha = 0.01, 10000 permutations and tolerance (tol) = 10 meters in order to remove GPS error at the study site.

## I converted the column "date" to factor to allow grouping the days into levels.
```{r}
data_guarei$date <- as.factor(data_guarei$date)
class(data_guarei$date)
days <- levels(data_guarei$date); head(days)
```


## Next step was creating an output datafame with a a column "q" for printing the "q" values which I am testing.
```{r}
output_guarei <- data_guarei
output_guarei$CP <- ""
output_guarei$q <- ""
output_guarei <- output_guarei[-c(1:nrow(output_guarei)),]; head(output_guarei)
```

## Now I run the test for all "q" values (1:10) and all 25 days. I also used alpha=0.01, N=10000, and tol=0.

```{r}
for(date.mlp in days[1:26]) {
  
  print(date.mlp)
  
  for(q in 1:10) {
    
    print(paste(date.mlp, "q =", q))
    
    datainp_g <- data_guarei[data_guarei$date == date.mlp,]
    
    inp<-matrix(c(0),dim(datainp_g)[1],2)
    inp<-datainp_g[,6:7]
    
    x1<-inp[[1]]
    x2<-inp[[2]]
    
    # Inspect first few rows of the data
    xy <- cbind(x1,x2)
    head(xy)
    
    # INPUTS
    # input alpha
    # alpha = significance level
    # alpha = 0.01 is a convenient default
    alpha <- 0.01
    
    # input q
    # Enter value of q
    # q = 4 is used here just as an example.
    # The user should replace this by a value of q
    # that is appropriate for the species and data
    # considered, e.g. that obtained by running CPT_Rcode
    # on part of the data.
    #q <- 6
    
    # input N
    # N = total number of permutations (1 observed and N-1 simulated) 
    # # N = 10000 is a convenient number
    N <- 10000
    
    # input tol
    # tol = tolerance
    # = maximum distance between indistinguishable positions # tol = 0 is a 
    # convenient default
    # The user may like to replace this by
    # twice the average GPS error in research area
    tol <- 0
    
    # PRELIMINARIES
    # Reverse the time-ordering,
    # so that (bx1[1], bx2[1]) refers to (final)
    #  putative goal
    bx1 <- 0*x1
    bx2 <- 0*x2
    
    n <- length(x1)
    for (j in 1:n){
      bx1[j] <- x1[n-j+1]
      bx2[j] <- x2[n-j+1]
    }
    # Calculate the steps (bxdiff1, bxdiff2)
    bx1diff <- diff(bx1)
    bx2diff <- diff(bx2)
    
    # REMOVE POINTS AT WHICH ANIMAL STAYS STILL
    # ind is (reverse) time ordering of points
    # newp > 0 if point differs from previous point
    tolsq <- tol^2
    ind <- c(1:n)
    newp <- c(1:n)
    for (j in 2:n){
      newp[j] <- ind[j]*( bx1diff[j-1]^2 + bx2diff[j-1]^2 > tolsq)
    }
    
    # bz1, bz2 are coordinates of points(in reverse time order) 
    # at which there is movement
    bz1 <- bx1[newp >0]
    bz2 <- bx2[newp >0]
    nz <- length(bz1)
    
    # We shall apply the CPT to points with coordinates (bz1,bz2)
    # Calculate the steps (bzdiff1, bzdiff2)
    bz1diff <- diff(bz1)
    bz2diff <- diff(bz2)
    # SOME DECLARATIONS
    # goal.no = number (from end) of current putative goal 
    # (with goal.no = 1 for end position
    
    # Thus goal.no = goal.t + 1
    # where goal.t was used in original code to refer to
    # time of current putative goal (with goal.t = 0 
    # for # end position # start with goal.no = 1
    goal.no <- 1
    
    # last.no = number (backwards in time) of last position of 
    # interest # last.no = length(bz1) is a convenient default
    # (and includes all the positions)
    last.no <- length(bz1)
    no.of.nos <- length(bz1)
    Rsumrand  <- rep(0, len=N)
    Rsumrandr <- rep(0, len=N)
    
    # Pr will store observed p-values in a run of r
    Pr <- rep(1, len=no.of.nos)
    
    # sig is a vector indicating whether or not
    # waypoint k is detected as a possible change point: 
    # sig[k] = 1 if change point detected at waypoint k 
    # sig[k] = 0 otherwise
    sig <- rep(0,nz)
    
    # k is number of steps in "k-leg"
    k <- 0
    
    # LOOK (SEQUENTIALLY) FOR NEXT POSSIBLE CHANGE POINT 
    
    while(goal.no < last.no - q){
      
      print(goal.no)
      
      k <- 0
      
      # INCREASE k UNTIL NEXT POSSIBLE CHANGE POINT IS FOUND 
      # Re-initialise Pr
      Pr <- 0*Pr + 1
      P <- 1
      
      while ((P > alpha) && (goal.no + q + k < last.no)){ 
        #B
        
        k <- k + 1
        
        R1 <- sqrt((bz1[goal.no+k] - bz1[goal.no])^2 + (bz2[goal.no+k] - 
                                                          bz2[goal.no])^2) 
        R2 <- sqrt((bz1[goal.no+k+q] - bz1[goal.no+k])^2 + (bz2[goal.no+k+q] - 
                                                              bz2[goal.no+k])^2)
        Rsum <- R1 + R2
        
        # Rsumrand[1] = observed value of statistic R1 + R2 
        Rsumrand[1] <- Rsum
        
        # Now calculate statistic R1 + R2 for a further N-1 
        # random permutations and store in Rsumrand
        for (it in 2:N){
          # start of it loop C
          u <- runif(k+q,0,1)
          perm <- order(u)
          bz1r <- bz1[goal.no]
          bz2r <- bz2[goal.no]
          for (j in 1:k){ # start of j loop D 
            bz1r <- bz1r + bz1diff[goal.no-1+perm[j]] 
            bz2r <- bz2r + bz2diff[goal.no-1+perm[j]]
          } # end of j loop D
          
          R1rand <- sqrt((bz1r - bz1[goal.no])^2 + (bz2r - bz2[goal.no])^2)
          R2rand <- sqrt((bz1[goal.no+k+q] - bz1r)^2 + 
                           (bz2[goal.no+k+q] - bz2r)^2) 
          Rsumrand[it] <- R1rand + R2rand
        } # end of it loop C
        
        P <- sum(Rsumrand >= Rsum)/N
        
        Pr[k] <- P
        
      } # end of "while" on
      # ((P > alpha) && (goal.no + q + k < last.no)) B
      # f is the first value of k in the current run that is significant 
      f <- k
      
      while ((P <= alpha) && (goal.no + q + k < last.no)){ 
        #E
        
        k <- k+1
        R1 <- sqrt((bz1[goal.no+k] - bz1[goal.no])^2 + 
                     (bz2[goal.no+k] - bz2[goal.no])^2)
        
        R2 <- sqrt((bz1[goal.no+k+q] - bz1[goal.no+k])^2 + 
                     (bz2[goal.no+k+q] - bz2[goal.no+k])^2)
        
        Rsum <- R1 + R2
        
        # Rsumrand[1] = observed value of statistic R1 + R2 
        Rsumrand[1] <- Rsum
        
        # Now calculate statistic R1 + R2 for a further N-1 
        # random permutations and store in Rsumrand
        for (it in 2:N){ # start of it loop F
          u <- runif(k+q,0,1)
          perm <- order(u)
          bz1r <- bz1[goal.no]
          bz2r <- bz2[goal.no]
          for (j in 1:k){ # start of j loop G 
            bz1r <- bz1r + bz1diff[goal.no-1+perm[j]] 
            bz2r <- bz2r + bz2diff[goal.no-1+perm[j]]
          } # end of j loop G
          
          R1rand <- sqrt((bz1r - bz1[goal.no])^2 + 
                           (bz2r - bz2[goal.no])^2)
          R2rand <- sqrt((bz1[goal.no+k+q] - bz1r)^2 + 
                           (bz2[goal.no+k+q] - bz2r)^2)      
          Rsumrand[it] <- R1rand + R2rand
        } # end of it loop F
        
        P <- sum(Rsumrand >= Rsum)/N
        
        Pr[k] <- P
        
      }  # end of "while" on
      
      # ((P < alpha) && (goal.no + q + k < last.no)) E
      
      # l is the last value of k in the current run that is significant 
      l <- k - 1
      
      # apply "peak rule"
      Pr[f:l]
      rmin <- min(which(Pr[f:l] == min(Pr[f:l]))) 
      rmin <- if(l >=f) rmin else 0
      
      goal.no <- ifelse(rmin > 0, goal.no + f + rmin - 1, last.no) 
      
      sig[goal.no] <- ifelse(goal.no == last.no,0,1)
      
    } # end of "while" on (goal.no < last.no - q) # A
    
    #  Remove putative goal from list of CP's
    sig[1] <- 0
    
    # OUTPUTS
    
    # OUTPUT OF
    # (first, last) (row nos. of first and last times at 
    # "significant" change points)
    # (east, north) (their coordinates)
    # cp lists indices of waypoints identified as change points
    cp <- which(sig==1)
    
    
    # bsig is sig in reverse time-order
    bsig <- rep(0,nz)
    for (j in 1:nz){
      bsig[j] <- sig[nz-j+1]
    }
    
    # "times" (row nos.) and coordinates of change points 
    newpp <- newp[newp > tol]
    cp.time <- newpp[sig==1]
    cp.bx1 <- bz1[sig==1]
    cp.bx2 <- bz2[sig==1]
    cp.no <- n + 1 - cp.time
    # last is cp.no in reverse order
    last <- 0*cp.no
    cp.leng <- length(cp.no)
    for (j in 1:cp.leng){
      last[j] <- cp.no[cp.leng-j+1]
    }
    # last contains row nos. of (last times at) change points 
    # (east, north) are their coordinates
    east <- x2[last]
    north <- x1[last]
    
    # first contains row nos. of first times at change points 
    first <- 0*last
    for (j in 1:length(last)){
      first[j] <- n+2 - min(which(newp > (n+1-last[j])))
    }
    
    # PLOT WAYPOINTS AND MARK CHANGE POINTS
    # For plotting purposes, reverse the order of
    # bz1, bz2 to get z1, z2
    z1 <- 0*bz1
    z2 <- 0*bz2
    nz <- length(z1)
    for (j in 1:nz){
      z1[j] <- bz1[nz-j+1]
      z2[j] <- bz2[nz-j+1]
    }
    # s is vector to index the points
    s <- c(1:nz)
    
    # Save change-points
    z1bsig <- z1[bsig==1]# Latitude of CPs
    z2bsig <- z2[bsig==1]# Longitude of CPs
    cpsame <- which(first == last)
    cpdiff <- which(first != last)
    lastsame <- last[cpsame]# Number of lines where CPs are located 
    firstdiff <- first[cpdiff]
    lastdiff <- last[cpdiff]
    z1bsame <- z1bsig[cpsame]
    z2bsame <- z2bsig[cpsame]
    z1bdiff <- z1bsig[cpdiff]
    z2bdiff <- z2bsig[cpdiff]
    
    
    # create a new column
    datainp_g$cp <- "NO-CP"; head(datainp_g)
    # Fill CPs in the change point locations
    true.index <- which((datainp_g$longitude %in% z2bsig) & 
                          (datainp_g$latitude %in% z1bsig))
    datainp_g$cp[true.index] <- "CP"
    
    # fill q value
    datainp_g$q <- q
    
    # plot
    plot(datainp_g$longitude, datainp_g$latitude, pch = 19, 
         main = datainp_g$date[1])
    # lines
    lines(datainp_g$longitude, datainp_g$latitude)
    points(datainp_g[datainp_g$cp == "CP",]$longitude, 
           datainp_g[datainp_g$cp == "CP",]$latitude, pch = 19, col = 2)
    
    output_guarei <- rbind(output_guarei, datainp_g)
  }
}
```

## Filtering duplicated lines so I don't count more than once the same "CP"
```{r}
output_guarei_filter<- output_guarei[,-c(1,3,4,5)]; head(output_guarei_filter)
output_guarei_filter<- output_guarei_filter[!duplicated(output_guarei_filter), ]
head(output_guarei_filter)
```

## Summary to understand the number of "CPs" according to value of "q"
```{r}
count_cps_guarei <- output_guarei_filter %>% 
  dplyr::group_by(date, q) %>% 
  dplyr::summarize(number.of.CPs = sum(cp == "CP"))
head(count_cps_guarei)
```

## Now I organize the data into wide format to plot and understand what is the best value of "q" to use:
```{r}
count_cps_guarei_wide <- spread(count_cps_guarei, key = date, 
                                value = number.of.CPs)
# Create "total"TOTAL" column with sum or row values:
count_cps_guarei_wide <- count_cps_guarei_wide %>%
  mutate(TOTAL=rowSums(.[setdiff(names(.),"q")]))
# Converting to data.frame:
count_cps_guarei_wide<-as.data.frame(count_cps_guarei_wide)
```

## Ploting "q" for Guareí:
### Plot of number of CPs per day according to values of "q" from 1 to 10. Light-gray lines represent number of CPs for each value of "q" in a unique day. Thick black line represent the sum of CPs for each value of "q" in all days.
```{r}
par(mfrow=c(1,1))
plot(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"10/06/19"],
     type="o",pch=20,
     col="gray85",xlab="q",ylab ="Number of change- points",
     xlim=c(1,max(count_cps_guarei_wide[,"q"])),
     ylim=c(1,max(count_cps_guarei_wide)),lab=c(10,20,0))
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"11/06/19"],
       type="o",pch=20,col="gray85")
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"11/07/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"11/08/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"12/06/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"12/07/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"12/08/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"13/07/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"13/08/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"14/06/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"14/08/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"15/06/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"15/08/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"16/05/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"16/07/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"17/05/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"17/06/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"17/07/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"18/05/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"18/07/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"19/07/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"19/08/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"20/05/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"20/07/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"20/08/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"21/08/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_guarei_wide[,"q"],count_cps_guarei_wide[,"TOTAL"],
       type="o",pch=20,col="black", lwd=3)
```

## It looks like q = 7 is the best for running the analysis, since it is the last value at which the number of CPs increases continuously. 

## After finding q, I just have to create a subset of my "output_guarei" file with q = 7.
```{r}
cpt_guarei <- subset(output_guarei, q == 7)
cpt_guarei <- subset(cpt_guarei, select = -id)
cpt_guarei$id <- 1:nrow(cpt_guarei)
cpt_guarei<-cpt_guarei[,c(9,1,2,3,4,5,6,7,8)]
head(cpt_guarei)
```

***
# Here I present the plot with all identified CPs and associated behaviors in Guareí. Also, I plot here 3 different zones (territory and fragment borders and core area). For this, I have worked in the dataframes in excel previously to organize the data acoordingly.

### Identification of behaviors associated with direction change (CPs) along black lion tamarins' daily routes. Pink triangles = frugivory (67%), cyan diamonds = foraging (24%), red circles = locomotion (5%), orange upside-down triangles = encounter (3%), light-gray circle = fur-rubbing (1%). Little black dots and lines repesent GPS points recorded every 5 minutes.
```{r}
#tiff(".tiff", width = 11000, height = 6500, res = 1000)
ggplot()+
  geom_polygon(data = forest_guarei, aes(x=X, y= Y),
               fill = "darkgreen", colour = "black", size = 1)+
  geom_polygon(data = k95, aes(x=X, y= Y),
               fill = "palegreen4", colour = "black")+
  geom_polygon(data = k70, aes(x=X, y= Y),
               fill = "yellowgreen", colour = "black") +
  #geom_polygon(data = k50, aes(x=long, y= lat, group=group),
               #fill = "yellowgreen", colour = "black") +
  #geom_polygon(data = k50, aes(x=long, y= lat, group=group),
               #fill = "mediumseagreen", colour = "black")+
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, size=2))+
  geom_point(data = output.guarei1, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.guarei1, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.guarei2, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.guarei2, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.guarei3, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.guarei3, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.guarei4, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.guarei4, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.guarei5, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.guarei5, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.guarei6, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.guarei6, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.guarei7, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.guarei7, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.guarei8, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.guarei8, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.guarei9, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.guarei9, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.guarei10, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei10, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei11, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei11, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei12, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei12, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei13, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei13, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei14, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei14, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei15, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei15, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei16, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei16, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei17, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei17, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei18, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei18, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei19, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei19, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei20, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei20, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei21, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei21, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei22, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei22, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei23, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei23, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei24, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei24, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei25, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei25, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei26, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.guarei26, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.guarei[output.guarei$cpffru == "x",], 
             aes(x = longitude, y = latitude), fill = "pink", 
             shape = 24, size = 4, stroke = 1)+
  geom_point(data = output.guarei[output.guarei$cplocom == "x",], 
             aes(x = longitude, y = latitude), fill = "red", 
             shape = 21, size = 4, stroke = 1)+
  geom_point(data = output.guarei[output.guarei$cpfor == "x",], 
             aes(x = longitude, y = latitude), fill = "cyan", 
             shape = 23, size = 4, stroke = 1)+
  geom_point(data = output.guarei[output.guarei$cpagonin == "x",], 
             aes(x = longitude, y = latitude), fill = "coral", 
             shape = 25, size = 4, stroke = 1)+
  geom_point(data = output.guarei[output.guarei$cprubb == "x",], 
             aes(x = longitude, y = latitude), fill = "aliceblue", 
             shape = 21, size = 4, stroke = 1) +
  coord_fixed()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=2),
        plot.margin=unit(c(0.1,0,0,0),"cm"))+
  theme(axis.text.x = element_text(hjust=0.5, size = 8, colour ="Black"))+
  theme(axis.text.y = element_text(angle=90, hjust=0.5, 
                                   size = 8, colour ="Black"))+
  theme(axis.title.y = element_blank())+
  theme(axis.title.x = element_blank())+
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  annotation_scale(location = "tl", width_hint = 0.2, 
                   pad_x = unit(0.1, "in"), pad_y = unit(0.8, "in"))+
  annotation_north_arrow(location = "tl", which_north = "true", 
                         pad_x = unit(0.3, "in"), pad_y = unit(0.1, "in"),
                         style = north_arrow_fancy_orienteering)
#dev.off()
```

*** 

# Taquara

```{r}
load("cpt_taquara.rda")
```

# Finding the best value of "q". 

## Converting "data" to factor:
```{r}
data_taquara$date <- as.factor(data_taquara$date)
class(data_taquara$date)
days <- levels(data_taquara$date); head(days)
```


## Creating an output datafame with a a column "q" for printing the "q" values which I am testing.
```{r}
output_taquara <- data_taquara
output_taquara$CP <- ""
output_taquara$q <- ""
output_taquara <- output_taquara[-c(1:nrow(output_taquara)),]
head(output_taquara)
```

## Now I run the test for all "q" values (1:10) and all 11 days. I also used alpha=0.01, N=10000, and tol=0.

```{r}
for(date.mlp in days[1:11]) {
  
  print(date.mlp)
  
  for(q in 1:10) {
    
    print(paste(date.mlp, "q =", q))
    
    datainp_t <- data_taquara[data_taquara$date == date.mlp,]
    
    inp<-matrix(c(0),dim(datainp_t)[1],2)
    inp<-datainp_t[,6:7]
    
    x1<-inp[[1]]
    x2<-inp[[2]]
    
    # Inspect first few rows of the data
    xy <- cbind(x1,x2)
    head(xy)
    
    # INPUTS
    # input alpha
    # alpha = significance level
    # alpha = 0.01 is a convenient default
    alpha <- 0.01
    
    # input q
    # Enter value of q
    # q = 4 is used here just as an example.
    # The user should replace this by a value of q
    # that is appropriate for the species and data
    # considered, e.g. that obtained by running CPT_Rcode
    # on part of the data.
    #q <- 6
    
    # input N
    # N = total number of permutations (1 observed and N-1 simulated) 
    # # N = 10000 is a convenient number
    N <- 10000
    
    # input tol
    # tol = tolerance
    # = maximum distance between indistinguishable positions # tol = 0 is a 
    # convenient default
    # The user may like to replace this by
    # twice the average GPS error in research area
    tol <- 0
    
    # PRELIMINARIES
    # Reverse the time-ordering,
    # so that (bx1[1], bx2[1]) refers to (final)
    #  putative goal
    bx1 <- 0*x1
    bx2 <- 0*x2
    
    n <- length(x1)
    for (j in 1:n){
      bx1[j] <- x1[n-j+1]
      bx2[j] <- x2[n-j+1]
    }
    # Calculate the steps (bxdiff1, bxdiff2)
    bx1diff <- diff(bx1)
    bx2diff <- diff(bx2)
    
    # REMOVE POINTS AT WHICH ANIMAL STAYS STILL
    # ind is (reverse) time ordering of points
    # newp > 0 if point differs from previous point
    tolsq <- tol^2
    ind <- c(1:n)
    newp <- c(1:n)
    for (j in 2:n){
      newp[j] <- ind[j]*( bx1diff[j-1]^2 + bx2diff[j-1]^2 > tolsq)
    }
    
    # bz1, bz2 are coordinates of points(in reverse time order) 
    # at which there is movement
    bz1 <- bx1[newp >0]
    bz2 <- bx2[newp >0]
    nz <- length(bz1)
    
    # We shall apply the CPT to points with coordinates (bz1,bz2)
    # Calculate the steps (bzdiff1, bzdiff2)
    bz1diff <- diff(bz1)
    bz2diff <- diff(bz2)
    # SOME DECLARATIONS
    # goal.no = number (from end) of current putative goal 
    # (with goal.no = 1 for end position
    
    # Thus goal.no = goal.t + 1
    # where goal.t was used in original code to refer to
    # time of current putative goal (with goal.t = 0 
    # for # end position # start with goal.no = 1
    goal.no <- 1
    
    # last.no = number (backwards in time) of last position of 
    # interest # last.no = length(bz1) is a convenient default
    # (and includes all the positions)
    last.no <- length(bz1)
    no.of.nos <- length(bz1)
    Rsumrand  <- rep(0, len=N)
    Rsumrandr <- rep(0, len=N)
    
    # Pr will store observed p-values in a run of r
    Pr <- rep(1, len=no.of.nos)
    
    # sig is a vector indicating whether or not
    # waypoint k is detected as a possible change point: 
    # sig[k] = 1 if change point detected at waypoint k 
    # sig[k] = 0 otherwise
    sig <- rep(0,nz)
    
    # k is number of steps in "k-leg"
    k <- 0
    
    # LOOK (SEQUENTIALLY) FOR NEXT POSSIBLE CHANGE POINT 
    
    while(goal.no < last.no - q){
      
      print(goal.no)
      
      k <- 0
      
      # INCREASE k UNTIL NEXT POSSIBLE CHANGE POINT IS FOUND 
      # Re-initialise Pr
      Pr <- 0*Pr + 1
      P <- 1
      
      while ((P > alpha) && (goal.no + q + k < last.no)){ 
        #B
        
        k <- k + 1
        
        R1 <- sqrt((bz1[goal.no+k] - bz1[goal.no])^2 + (bz2[goal.no+k] - 
                                                          bz2[goal.no])^2) 
        R2 <- sqrt((bz1[goal.no+k+q] - bz1[goal.no+k])^2 + (bz2[goal.no+k+q] - 
                                                              bz2[goal.no+k])^2)
        Rsum <- R1 + R2
        
        # Rsumrand[1] = observed value of statistic R1 + R2 
        Rsumrand[1] <- Rsum
        
        # Now calculate statistic R1 + R2 for a further N-1 
        # random permutations and store in Rsumrand
        for (it in 2:N){
          # start of it loop C
          u <- runif(k+q,0,1)
          perm <- order(u)
          bz1r <- bz1[goal.no]
          bz2r <- bz2[goal.no]
          for (j in 1:k){ # start of j loop D 
            bz1r <- bz1r + bz1diff[goal.no-1+perm[j]] 
            bz2r <- bz2r + bz2diff[goal.no-1+perm[j]]
          } # end of j loop D
          
          R1rand <- sqrt((bz1r - bz1[goal.no])^2 + (bz2r - bz2[goal.no])^2)
          R2rand <- sqrt((bz1[goal.no+k+q] - bz1r)^2 + 
                           (bz2[goal.no+k+q] - bz2r)^2) 
          Rsumrand[it] <- R1rand + R2rand
        } # end of it loop C
        
        P <- sum(Rsumrand >= Rsum)/N
        
        Pr[k] <- P
        
      } # end of "while" on
      # ((P > alpha) && (goal.no + q + k < last.no)) B
      # f is the first value of k in the current run that is significant 
      f <- k
      
      while ((P <= alpha) && (goal.no + q + k < last.no)){ 
        #E
        
        k <- k+1
        R1 <- sqrt((bz1[goal.no+k] - bz1[goal.no])^2 + 
                     (bz2[goal.no+k] - bz2[goal.no])^2)
        
        R2 <- sqrt((bz1[goal.no+k+q] - bz1[goal.no+k])^2 + 
                     (bz2[goal.no+k+q] - bz2[goal.no+k])^2)
        
        Rsum <- R1 + R2
        
        # Rsumrand[1] = observed value of statistic R1 + R2 
        Rsumrand[1] <- Rsum
        
        # Now calculate statistic R1 + R2 for a further N-1 
        # random permutations and store in Rsumrand
        for (it in 2:N){ # start of it loop F
          u <- runif(k+q,0,1)
          perm <- order(u)
          bz1r <- bz1[goal.no]
          bz2r <- bz2[goal.no]
          for (j in 1:k){ # start of j loop G 
            bz1r <- bz1r + bz1diff[goal.no-1+perm[j]] 
            bz2r <- bz2r + bz2diff[goal.no-1+perm[j]]
          } # end of j loop G
          
          R1rand <- sqrt((bz1r - bz1[goal.no])^2 + 
                           (bz2r - bz2[goal.no])^2)
          R2rand <- sqrt((bz1[goal.no+k+q] - bz1r)^2 + 
                           (bz2[goal.no+k+q] - bz2r)^2)      
          Rsumrand[it] <- R1rand + R2rand
        } # end of it loop F
        
        P <- sum(Rsumrand >= Rsum)/N
        
        Pr[k] <- P
        
      }  # end of "while" on
      
      # ((P < alpha) && (goal.no + q + k < last.no)) E
      
      # l is the last value of k in the current run that is significant 
      l <- k - 1
      
      # apply "peak rule"
      Pr[f:l]
      rmin <- min(which(Pr[f:l] == min(Pr[f:l]))) 
      rmin <- if(l >=f) rmin else 0
      
      goal.no <- ifelse(rmin > 0, goal.no + f + rmin - 1, last.no) 
      
      sig[goal.no] <- ifelse(goal.no == last.no,0,1)
      
    } # end of "while" on (goal.no < last.no - q) # A
    
    #  Remove putative goal from list of CP's
    sig[1] <- 0
    
    # OUTPUTS
    
    # OUTPUT OF
    # (first, last) (row nos. of first and last times at 
    # "significant" change points)
    # (east, north) (their coordinates)
    # cp lists indices of waypoints identified as change points
    cp <- which(sig==1)
    
    
    # bsig is sig in reverse time-order
    bsig <- rep(0,nz)
    for (j in 1:nz){
      bsig[j] <- sig[nz-j+1]
    }
    
    # "times" (row nos.) and coordinates of change points 
    newpp <- newp[newp > tol]
    cp.time <- newpp[sig==1]
    cp.bx1 <- bz1[sig==1]
    cp.bx2 <- bz2[sig==1]
    cp.no <- n + 1 - cp.time
    # last is cp.no in reverse order
    last <- 0*cp.no
    cp.leng <- length(cp.no)
    for (j in 1:cp.leng){
      last[j] <- cp.no[cp.leng-j+1]
    }
    # last contains row nos. of (last times at) change points 
    # (east, north) are their coordinates
    east <- x2[last]
    north <- x1[last]
    
    # first contains row nos. of first times at change points 
    first <- 0*last
    for (j in 1:length(last)){
      first[j] <- n+2 - min(which(newp > (n+1-last[j])))
    }
    
    # PLOT WAYPOINTS AND MARK CHANGE POINTS
    # For plotting purposes, reverse the order of
    # bz1, bz2 to get z1, z2
    z1 <- 0*bz1
    z2 <- 0*bz2
    nz <- length(z1)
    for (j in 1:nz){
      z1[j] <- bz1[nz-j+1]
      z2[j] <- bz2[nz-j+1]
    }
    # s is vector to index the points
    s <- c(1:nz)
    
    # Save change-points
    z1bsig <- z1[bsig==1]# Latitude of CPs
    z2bsig <- z2[bsig==1]# Longitude of CPs
    cpsame <- which(first == last)
    cpdiff <- which(first != last)
    lastsame <- last[cpsame]# Number of lines where CPs are located 
    firstdiff <- first[cpdiff]
    lastdiff <- last[cpdiff]
    z1bsame <- z1bsig[cpsame]
    z2bsame <- z2bsig[cpsame]
    z1bdiff <- z1bsig[cpdiff]
    z2bdiff <- z2bsig[cpdiff]
    
    
    # create a new column
    datainp_t$cp <- "NO-CP"; head(datainp_t)
    # Fill CPs in the change point locations
    true.index <- which((datainp_t$longitude %in% z2bsig) & 
                          (datainp_t$latitude %in% z1bsig))
    datainp_t$cp[true.index] <- "CP"
    
    # fill q value
    datainp_t$q <- q
    
    # plot
    plot(datainp_t$longitude, datainp_t$latitude, pch = 19, 
         main = datainp_t$date[1])
    # lines
    lines(datainp_t$longitude, datainp_t$latitude)
    points(datainp_t[datainp_t$cp == "CP",]$longitude, 
           datainp_t[datainp_t$cp == "CP",]$latitude, pch = 19, col = 2)
    
    output_taquara <- rbind(output_taquara, datainp_t)
  }
}
```

## Filtering duplicated lines so I don't count more than once the same "CP"
```{r}
output_taquara_filter<- output_taquara[,-c(1,3,4,5)]
head(output_taquara_filter)
output_taquara_filter<- output_taquara_filter[!duplicated
                                              (output_taquara_filter), ]
head(output_taquara_filter)
```

## Summary to understand the number of "CPs" according to value of "q"
```{r}
count_cps_taquara <- output_taquara_filter %>% 
  dplyr::group_by(date, q) %>% 
  dplyr::summarize(number.of.CPs = sum(cp == "CP"))
head(count_cps_taquara)
```

## Now I organize the data into wide format to plot and understand what is the best value of "q" to use:
```{r}
count_cps_taquara_wide <- spread(count_cps_taquara, key = date, 
                                value = number.of.CPs)
# Create "total"TOTAL" column with sum or row values:
count_cps_taquara_wide <- count_cps_taquara_wide %>%
  mutate(TOTAL=rowSums(.[setdiff(names(.),"q")]))
# Converting to data.frame:
count_cps_taquara_wide<-as.data.frame(count_cps_taquara_wide)
```

## Ploting "q" for Taquara:

### Plot of number of CPs per day according to values of "q" from 1 to 10. Light-gray lines represent number of CPs for each value of "q" in a unique day. Thick black line represent the sum of CPs for each value of "q" in all days.
```{r}
par(mfrow=c(1,1))
plot(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"03/12/17"],
     type="o",pch=20,
     col="gray85",xlab="q",ylab ="Number of change- points",
     xlim=c(1,max(count_cps_taquara_wide[,"q"])),
     ylim=c(1,max(count_cps_taquara_wide)),lab=c(10,20,0))
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"27/01/18"],
       type="o",pch=20,col="gray85")
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"28/01/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"29/01/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"30/01/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"31/01/18"],
       type="o",pch=20,col="gray85")
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"03/02/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"02/05/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"04/07/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"05/07/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"18/09/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_taquara_wide[,"q"],count_cps_taquara_wide[,"TOTAL"],
       type="o",pch=20,col="black", lwd=3)
```

## It looks like q = 5 is the best for running the analysis, since it is the last value at which the number of CPs increases continuously. 

## After finding q, I just have to create a subset of my "output_taquara" file with q = 5.
```{r}
cpt_taquara <- subset(output_taquara, q == 5)
cpt_taquara <- subset(cpt_taquara, select = -id)
cpt_taquara$id <- 1:nrow(cpt_taquara)
cpt_taquara<-cpt_taquara[,c(9,1,2,3,4,5,6,7,8)]
head(cpt_taquara)
```

***

# Here I present the plot with all identified CPs and associated behaviors in Morro do Diabo (Taquara) Also, I plot here 2 different zones (territory border and core area). For this, I have worked in the dataframes in excel previously to organize the data acoordingly.

### Identification of behaviors associated with direction change (CPs) along black lion tamarins' daily routes in Morro do Diabo (Taquara). Red circles = locomotion (44%), pink triangles = frugivory (24%), green squares = resting (15%), cyan diamonds = foraging (8%), purple squares = gummivory (7%), light-yellow star = inactive (2%). Little black dots and lines repesent GPS points recorded every 5 minutes.

```{r}
#tiff(".tiff", width = 11000, height = 6500, res = 1000)
ggplot()+
  geom_polygon(data = forest_taquara, aes(x=long, y= lat, group=group),
               fill = "darkgreen", size = 1)+
  geom_polygon(data = hr_taquara, aes(x=long, y= lat, group=group),
               fill = "palegreen4", colour = "black", size = 1)+
  geom_polygon(data = frabordf_taquara, aes(x=long, y= lat, group=group),
               fill = "mediumseagreen", colour = "black", size = .5)+
  geom_polygon(data = terbordf_taquara, aes(x=long, y= lat, group=group),
               fill = "yellowgreen", colour = "black") +
  geom_polygon(data = river_taquara, aes(x=long, y= lat, group=group),
               fill = "cadetblue1", colour = "black", size = 1) +
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, size=2))+
  geom_point(data = output.taquara1, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.taquara1, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.taquara2, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.taquara2, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.taquara3, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.taquara3, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.taquara4, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.taquara4, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.taquara5, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.taquara5, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.taquara6, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.taquara6, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.taquara7, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.taquara7, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.taquara8, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.taquara8, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.taquara9, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.taquara9, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.taquara10, aes(x = longitude, 
                                          y = latitude), size = 1) +
  geom_path(data = output.taquara10, aes(x = longitude, 
                                         y = latitude), size = 0.2) +
  geom_point(data = output.taquara11, aes(x = longitude, 
                                          y = latitude), size = 1) +
  geom_path(data = output.taquara11, aes(x = longitude, 
                                         y = latitude), size = 0.2) +
  geom_point(data = output.taquara[output.taquara$cpout == "x",], 
             aes(x = longitude, y = latitude), col = "darkorchid2", 
             shape = 3, size = 3.5, stroke = 1.5 )+
  geom_point(data = output.taquara[output.taquara$cpidle == "x",], 
             aes(x = longitude, y = latitude), col = "cornsilk", 
             shape = 8, size = 4, stroke = 1)+
  geom_point(data = output.taquara[output.taquara$cpres == "x",], 
             aes(x = longitude, y = latitude), fill = "chartreuse2", 
             shape = 22, size = 4, stroke = 1)+
  geom_point(data = output.taquara[output.taquara$cpagonin == "x",], 
             aes(x = longitude, y = latitude), fill = "coral", 
             shape = 25, size = 4, stroke = 1)+
  geom_point(data = output.taquara[output.taquara$cpgum == "x",], 
             aes(x = longitude, y = latitude), fill = "purple", 
             shape = 22, size = 4, stroke = 1) +
  geom_point(data = output.taquara[output.taquara$cplocom == "x",], 
             aes(x = longitude, y = latitude), fill = "red", 
             shape = 21, size = 4, stroke = 1)+
  geom_point(data = output.taquara[output.taquara$cpfor == "x",], 
             aes(x = longitude, y = latitude), fill = "cyan", 
             shape = 23, size = 4, stroke = 1)+
  geom_point(data = output.taquara[output.taquara$cpffru == "x",], 
             aes(x = longitude, y = latitude), fill = "pink", 
             shape = 24, size = 4, stroke = 1)+
  geom_point(data = output.taquara[output.taquara$cprubb == "x",], 
             aes(x = longitude, y = latitude), fill = "aliceblue", 
             shape = 21, size = 4, stroke = 1) +
  coord_fixed()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=2),
        plot.margin=unit(c(0.1,0,0,0),"cm"))+
  theme(axis.text.x = element_text(hjust=0.5, size = 8, colour ="Black"))+
  theme(axis.text.y = element_text(angle=90, hjust=0.5, 
                                   size = 8, colour ="Black"))+
  theme(axis.title.y = element_blank())+
  theme(axis.title.x = element_blank())+
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  annotation_scale(location = "tr", width_hint = 0.2, 
                   pad_x = unit(0.1, "in"), pad_y = unit(0.8, "in"))+
  annotation_north_arrow(location = "tr", which_north = "true", 
                         pad_x = unit(0.3, "in"), pad_y = unit(0.1, "in"),
                         style = north_arrow_fancy_orienteering)
#dev.off()
```


***

# Suzano

```{r}
load("cpt_suzano.rda")
```

# Finding the best value of "q". 

## Converting "data" to factor:
```{r}
data_suzano$date <- as.factor(data_suzano$date)
class(data_suzano$date)
days <- levels(data_suzano$date); head(days)
```


## Creating an output datafame with a a column "q" for printing the "q" values which I am testing.
```{r}
output_suzano <- data_suzano
output_suzano$CP <- ""
output_suzano$q <- ""
output_suzano <- output_suzano[-c(1:nrow(output_suzano)),]
head(output_suzano)
```

## Now I run the test for all "q" values (1:10) and all 11 days. I also used alpha=0.01, N=10000, and tol=0.

```{r}
for(date.mlp in days[1:13]) {
  
  print(date.mlp)
  
  for(q in 1:10) {
    
    print(paste(date.mlp, "q =", q))
    
    datainp_s <- data_suzano[data_suzano$date == date.mlp,]
    
    inp<-matrix(c(0),dim(datainp_s)[1],2)
    inp<-datainp_s[,6:7]
    
    x1<-inp[[1]]
    x2<-inp[[2]]
    
    # Inspect first few rows of the data
    xy <- cbind(x1,x2)
    head(xy)
    
    # INPUTS
    # input alpha
    # alpha = significance level
    # alpha = 0.01 is a convenient default
    alpha <- 0.01
    
    # input q
    # Enter value of q
    # q = 4 is used here just as an example.
    # The user should replace this by a value of q
    # that is appropriate for the species and data
    # considered, e.g. that obtained by running CPT_Rcode
    # on part of the data.
    #q <- 6
    
    # input N
    # N = total number of permutations (1 observed and N-1 simulated) 
    # # N = 10000 is a convenient number
    N <- 10000
    
    # input tol
    # tol = tolerance
    # = maximum distance between indistinguishable positions # tol = 0 is a 
    # convenient default
    # The user may like to replace this by
    # twice the average GPS error in research area
    tol <- 0
    
    # PRELIMINARIES
    # Reverse the time-ordering,
    # so that (bx1[1], bx2[1]) refers to (final)
    #  putative goal
    bx1 <- 0*x1
    bx2 <- 0*x2
    
    n <- length(x1)
    for (j in 1:n){
      bx1[j] <- x1[n-j+1]
      bx2[j] <- x2[n-j+1]
    }
    # Calculate the steps (bxdiff1, bxdiff2)
    bx1diff <- diff(bx1)
    bx2diff <- diff(bx2)
    
    # REMOVE POINTS AT WHICH ANIMAL STAYS STILL
    # ind is (reverse) time ordering of points
    # newp > 0 if point differs from previous point
    tolsq <- tol^2
    ind <- c(1:n)
    newp <- c(1:n)
    for (j in 2:n){
      newp[j] <- ind[j]*( bx1diff[j-1]^2 + bx2diff[j-1]^2 > tolsq)
    }
    
    # bz1, bz2 are coordinates of points(in reverse time order) 
    # at which there is movement
    bz1 <- bx1[newp >0]
    bz2 <- bx2[newp >0]
    nz <- length(bz1)
    
    # We shall apply the CPT to points with coordinates (bz1,bz2)
    # Calculate the steps (bzdiff1, bzdiff2)
    bz1diff <- diff(bz1)
    bz2diff <- diff(bz2)
    # SOME DECLARATIONS
    # goal.no = number (from end) of current putative goal 
    # (with goal.no = 1 for end position
    
    # Thus goal.no = goal.t + 1
    # where goal.t was used in original code to refer to
    # time of current putative goal (with goal.t = 0 
    # for # end position # start with goal.no = 1
    goal.no <- 1
    
    # last.no = number (backwards in time) of last position of 
    # interest # last.no = length(bz1) is a convenient default
    # (and includes all the positions)
    last.no <- length(bz1)
    no.of.nos <- length(bz1)
    Rsumrand  <- rep(0, len=N)
    Rsumrandr <- rep(0, len=N)
    
    # Pr will store observed p-values in a run of r
    Pr <- rep(1, len=no.of.nos)
    
    # sig is a vector indicating whether or not
    # waypoint k is detected as a possible change point: 
    # sig[k] = 1 if change point detected at waypoint k 
    # sig[k] = 0 otherwise
    sig <- rep(0,nz)
    
    # k is number of steps in "k-leg"
    k <- 0
    
    # LOOK (SEQUENTIALLY) FOR NEXT POSSIBLE CHANGE POINT 
    
    while(goal.no < last.no - q){
      
      print(goal.no)
      
      k <- 0
      
      # INCREASE k UNTIL NEXT POSSIBLE CHANGE POINT IS FOUND 
      # Re-initialise Pr
      Pr <- 0*Pr + 1
      P <- 1
      
      while ((P > alpha) && (goal.no + q + k < last.no)){ 
        #B
        
        k <- k + 1
        
        R1 <- sqrt((bz1[goal.no+k] - bz1[goal.no])^2 + (bz2[goal.no+k] - 
                                                          bz2[goal.no])^2) 
        R2 <- sqrt((bz1[goal.no+k+q] - bz1[goal.no+k])^2 + (bz2[goal.no+k+q] - 
                                                              bz2[goal.no+k])^2)
        Rsum <- R1 + R2
        
        # Rsumrand[1] = observed value of statistic R1 + R2 
        Rsumrand[1] <- Rsum
        
        # Now calculate statistic R1 + R2 for a further N-1 
        # random permutations and store in Rsumrand
        for (it in 2:N){
          # start of it loop C
          u <- runif(k+q,0,1)
          perm <- order(u)
          bz1r <- bz1[goal.no]
          bz2r <- bz2[goal.no]
          for (j in 1:k){ # start of j loop D 
            bz1r <- bz1r + bz1diff[goal.no-1+perm[j]] 
            bz2r <- bz2r + bz2diff[goal.no-1+perm[j]]
          } # end of j loop D
          
          R1rand <- sqrt((bz1r - bz1[goal.no])^2 + (bz2r - bz2[goal.no])^2)
          R2rand <- sqrt((bz1[goal.no+k+q] - bz1r)^2 + 
                           (bz2[goal.no+k+q] - bz2r)^2) 
          Rsumrand[it] <- R1rand + R2rand
        } # end of it loop C
        
        P <- sum(Rsumrand >= Rsum)/N
        
        Pr[k] <- P
        
      } # end of "while" on
      # ((P > alpha) && (goal.no + q + k < last.no)) B
      # f is the first value of k in the current run that is significant 
      f <- k
      
      while ((P <= alpha) && (goal.no + q + k < last.no)){ 
        #E
        
        k <- k+1
        R1 <- sqrt((bz1[goal.no+k] - bz1[goal.no])^2 + 
                     (bz2[goal.no+k] - bz2[goal.no])^2)
        
        R2 <- sqrt((bz1[goal.no+k+q] - bz1[goal.no+k])^2 + 
                     (bz2[goal.no+k+q] - bz2[goal.no+k])^2)
        
        Rsum <- R1 + R2
        
        # Rsumrand[1] = observed value of statistic R1 + R2 
        Rsumrand[1] <- Rsum
        
        # Now calculate statistic R1 + R2 for a further N-1 
        # random permutations and store in Rsumrand
        for (it in 2:N){ # start of it loop F
          u <- runif(k+q,0,1)
          perm <- order(u)
          bz1r <- bz1[goal.no]
          bz2r <- bz2[goal.no]
          for (j in 1:k){ # start of j loop G 
            bz1r <- bz1r + bz1diff[goal.no-1+perm[j]] 
            bz2r <- bz2r + bz2diff[goal.no-1+perm[j]]
          } # end of j loop G
          
          R1rand <- sqrt((bz1r - bz1[goal.no])^2 + 
                           (bz2r - bz2[goal.no])^2)
          R2rand <- sqrt((bz1[goal.no+k+q] - bz1r)^2 + 
                           (bz2[goal.no+k+q] - bz2r)^2)      
          Rsumrand[it] <- R1rand + R2rand
        } # end of it loop F
        
        P <- sum(Rsumrand >= Rsum)/N
        
        Pr[k] <- P
        
      }  # end of "while" on
      
      # ((P < alpha) && (goal.no + q + k < last.no)) E
      
      # l is the last value of k in the current run that is significant 
      l <- k - 1
      
      # apply "peak rule"
      Pr[f:l]
      rmin <- min(which(Pr[f:l] == min(Pr[f:l]))) 
      rmin <- if(l >=f) rmin else 0
      
      goal.no <- ifelse(rmin > 0, goal.no + f + rmin - 1, last.no) 
      
      sig[goal.no] <- ifelse(goal.no == last.no,0,1)
      
    } # end of "while" on (goal.no < last.no - q) # A
    
    #  Remove putative goal from list of CP's
    sig[1] <- 0
    
    # OUTPUTS
    
    # OUTPUT OF
    # (first, last) (row nos. of first and last times at 
    # "significant" change points)
    # (east, north) (their coordinates)
    # cp lists indices of waypoints identified as change points
    cp <- which(sig==1)
    
    
    # bsig is sig in reverse time-order
    bsig <- rep(0,nz)
    for (j in 1:nz){
      bsig[j] <- sig[nz-j+1]
    }
    
    # "times" (row nos.) and coordinates of change points 
    newpp <- newp[newp > tol]
    cp.time <- newpp[sig==1]
    cp.bx1 <- bz1[sig==1]
    cp.bx2 <- bz2[sig==1]
    cp.no <- n + 1 - cp.time
    # last is cp.no in reverse order
    last <- 0*cp.no
    cp.leng <- length(cp.no)
    for (j in 1:cp.leng){
      last[j] <- cp.no[cp.leng-j+1]
    }
    # last contains row nos. of (last times at) change points 
    # (east, north) are their coordinates
    east <- x2[last]
    north <- x1[last]
    
    # first contains row nos. of first times at change points 
    first <- 0*last
    for (j in 1:length(last)){
      first[j] <- n+2 - min(which(newp > (n+1-last[j])))
    }
    
    # PLOT WAYPOINTS AND MARK CHANGE POINTS
    # For plotting purposes, reverse the order of
    # bz1, bz2 to get z1, z2
    z1 <- 0*bz1
    z2 <- 0*bz2
    nz <- length(z1)
    for (j in 1:nz){
      z1[j] <- bz1[nz-j+1]
      z2[j] <- bz2[nz-j+1]
    }
    # s is vector to index the points
    s <- c(1:nz)
    
    # Save change-points
    z1bsig <- z1[bsig==1]# Latitude of CPs
    z2bsig <- z2[bsig==1]# Longitude of CPs
    cpsame <- which(first == last)
    cpdiff <- which(first != last)
    lastsame <- last[cpsame]# Number of lines where CPs are located 
    firstdiff <- first[cpdiff]
    lastdiff <- last[cpdiff]
    z1bsame <- z1bsig[cpsame]
    z2bsame <- z2bsig[cpsame]
    z1bdiff <- z1bsig[cpdiff]
    z2bdiff <- z2bsig[cpdiff]
    
    
    # create a new column
    datainp_s$cp <- "NO-CP"; head(datainp_s)
    # Fill CPs in the change point locations
    true.index <- which((datainp_s$longitude %in% z2bsig) & 
                          (datainp_s$latitude %in% z1bsig))
    datainp_s$cp[true.index] <- "CP"
    
    # fill q value
    datainp_s$q <- q
    
    # plot
    plot(datainp_s$longitude, datainp_s$latitude, pch = 19, 
         main = datainp_s$date[1])
    # lines
    lines(datainp_s$longitude, datainp_s$latitude)
    points(datainp_s[datainp_s$cp == "CP",]$longitude, 
           datainp_s[datainp_s$cp == "CP",]$latitude, pch = 19, col = 2)
    
    output_suzano <- rbind(output_suzano, datainp_s)
  }
}
```

## Filtering duplicated lines so I don't count more than once the same "CP"
```{r}
output_suzano_filter<- output_suzano[,-c(1,3,4,5)]
head(output_suzano_filter)
output_suzano_filter<- output_suzano_filter[!duplicated
                                              (output_suzano_filter), ]
head(output_suzano_filter)
```

## Summary to understand the number of "CPs" according to value of "q"
```{r}
count_cps_suzano <- output_suzano_filter %>% 
  dplyr::group_by(date, q) %>% 
  dplyr::summarize(number.of.CPs = sum(cp == "CP"))
head(count_cps_suzano)
```

## Now I organize the data into wide format to plot and understand what is the best value of "q" to use:
```{r}
count_cps_suzano_wide <- spread(count_cps_suzano, key = date, 
                                value = number.of.CPs)
# Create "total"TOTAL" column with sum or row values:
count_cps_suzano_wide <- count_cps_suzano_wide %>%
  mutate(TOTAL=rowSums(.[setdiff(names(.),"q")]))
# Converting to data.frame:
count_cps_suzano_wide<-as.data.frame(count_cps_suzano_wide)
```

## Ploting "q" for Suzano:

### Plot of number of CPs per day according to values of "q" from 1 to 10. Light-gray lines represent number of CPs for each value of "q" in a unique day. Thick black line represent the sum of CPs for each value of "q" in all days.

```{r}
par(mfrow=c(1,1))
plot(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"09/09/18"],
     type="o",pch=20,
     col="gray85",xlab="q",ylab ="Number of change- points - Suzano",
     xlim=c(1,max(count_cps_suzano_wide[,"q"])),
     ylim=c(1,max(count_cps_suzano_wide)),lab=c(10,20,0))
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"10/09/18"],
       type="o",pch=20,col="gray85")
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"11/09/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"06/12/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"07/12/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"08/12/18"],
       type="o",pch=20,col="gray85")
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"09/12/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"10/12/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"11/12/18"],
       type="o",pch=20,col="gray85") 
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"19/02/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"20/02/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"06/04/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"12/04/19"],
       type="o",pch=20,col="gray85") 
points(count_cps_suzano_wide[,"q"],count_cps_suzano_wide[,"TOTAL"],
       type="o",pch=20,col="black", lwd=3)
```

## It looks like q = 5 is the best for running the analysis, since it is the last value at which the number of CPs increases continuously. 

## After finding q, I just have to create a subset of my "output_suzano" file with q = 6.
```{r}
cpt_suzano <- subset(output_suzano, q == 5)
cpt_suzano <- subset(cpt_suzano, select = -id)
cpt_suzano$id <- 1:nrow(cpt_suzano)
cpt_suzano<-cpt_suzano[,c(9,1,2,3,4,5,6,7,8)]
head(cpt_suzano)
```

***

# Here I present the plot with all identified CPs and associated behaviors in Suzano. Also, I plot here 3 different zones (territory and fragment border and core area). For this, I have worked in the dataframes in excel previously to organize the data acoordingly.

### Identification of behaviors associated with direction change (CPs) along black lion tamarins' daily routes in Suzano. Red circles = locomotion (40%), pink triangles = frugivory (32%), green squares = resting (26%), purple squares = gummivory (2%). Little black dots and lines repesent GPS points recorded every 5 minutes.

```{r}
#tiff("zuzano_k95_k70.tiff", width = 11000, height = 6500, res = 1000)
ggplot()+
  #geom_polygon(data = plantations_suzano, aes(x=long, y= lat, group=group),
              # fill = "white", colour = "black", size = 1.5)+
  geom_polygon(data = forest_suzano, aes(x=long, y= lat, group=group),
               fill = "darkgreen", colour = "black",size = 1)+
  geom_polygon(data = river_suzano, aes(x=long, y= lat, group=group),
               fill = "cadetblue1", colour = "black", size = .6) +
  #geom_polygon(data = frabordf_suzano, aes(x=long, y= lat, group=group),
               #fill = "mediumseagreen", colour = "black", size = .5)+
  geom_polygon(data = k95_suz, aes(x=long, y= lat, group=group),
               fill = "palegreen4", colour = "black", size = .5)+
  geom_polygon(data = k70_suz, aes(x=long, y= lat, group=group),
               fill = "yellowgreen", colour = "black") +
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, size=2))+
  geom_point(data = output.suzano1, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.suzano1, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.suzano2, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.suzano2, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.suzano3, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.suzano3, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.suzano4, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.suzano4, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.suzano5, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.suzano5, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.suzano6, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.suzano6, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.suzano7, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.suzano7, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.suzano8, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.suzano8, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.suzano9, aes(x = longitude, 
                                        y = latitude), size = 1) +
  geom_path(data = output.suzano9, aes(x = longitude, 
                                       y = latitude), size = 0.2) +
  geom_point(data = output.suzano10, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.suzano10, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.suzano11, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.suzano11, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.suzano12, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.suzano12, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.suzano13, aes(x = longitude, 
                                         y = latitude), size = 1) +
  geom_path(data = output.suzano13, aes(x = longitude, 
                                        y = latitude), size = 0.2) +
  geom_point(data = output.suzano[output.suzano$cplocom == "x",], 
             aes(x = longitude, y = latitude), fill = "red", 
             shape = 21, size = 4, stroke = 1)+
  geom_point(data = output.suzano[output.suzano$cpffru == "x",], 
             aes(x = longitude, y = latitude), fill = "pink", 
             shape = 24, size = 4, stroke = 1)+
  geom_point(data = output.suzano[output.suzano$cpres == "x",], 
             aes(x = longitude, y = latitude), fill = "chartreuse2", 
             shape = 22, size = 4, stroke = 1)+  
  geom_point(data = output.suzano[output.suzano$cpgum == "x",], 
             aes(x = longitude, y = latitude), fill = "purple", 
             shape = 22, size = 4, stroke = 1) +
  coord_fixed()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=2),
        plot.margin=unit(c(0.1,0,0,0),"cm"))+
  theme(axis.text.x = element_text(hjust=0.5, size = 8, colour ="Black"))+
  theme(axis.text.y = element_text(angle=90, hjust=0.5, 
                                   size = 8, colour ="Black"))+
  theme(axis.title.y = element_blank())+
  theme(axis.title.x = element_blank())+
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  annotation_scale(location = "tl", width_hint = 0.2, 
                   pad_x = unit(0.1, "in"), pad_y = unit(0.8, "in"))+
  annotation_north_arrow(location = "tl", which_north = "true", 
                         pad_x = unit(0.3, "in"), pad_y = unit(0.1, "in"),
                         style = north_arrow_fancy_orienteering)
#dev.off()
```

# Santa Maria

```{r}
load("cpt_sma.rda")

forest_sma <- st_read("SantaMaria_FB_plot_2024_02_d12.gpkg")
forest_sma <- st_coordinates(forest_sma)
forest_sma <- as.data.frame(forest_sma)

k95_sma <- st_read("sma_akde95.gpkg")
k95_sma <- st_coordinates(k95_sma)
k95_sma <- as.data.frame(k95_sma)

k70_sma <- st_read("sma_akde75.gpkg")
k70_sma <- st_coordinates(k70_sma)
k70_sma <- as.data.frame(k70_sma)
```

# Finding the best value of "q". 

## Converting "data" to factor:
```{r}
data_sma$date <- as.factor(data_sma$date)
class(data_sma$date)
days <- levels(data_sma$date); head(days)
```


## Creating an output datafame with a a column "q" for printing the "q" values which I am testing.
```{r}
output_sma <- data_sma
output_sma$CP <- ""
output_sma$q <- ""
output_sma <- output_sma[-c(1:nrow(output_sma)),]
head(output_sma)
```

## Now I run the test for all "q" values (1:10) and all 11 days. I also used alpha=0.01, N=10000, and tol=0.

```{r}
for(date.mlp in days[1:11]) {
  
  print(date.mlp)
  
  for(q in 1:10) {
    
    print(paste(date.mlp, "q =", q))
    
    datainp_sma <- data_sma[data_sma$date == date.mlp,]
    
    inp<-matrix(c(0),dim(datainp_sma)[1],2)
    inp<-datainp_sma[,6:7]
    
    x1<-inp[[1]]
    x2<-inp[[2]]
    
    # Inspect first few rows of the data
    xy <- cbind(x1,x2)
    head(xy)
    
    # INPUTS
    # input alpha
    # alpha = significance level
    # alpha = 0.01 is a convenient default
    alpha <- 0.01
    
    # input q
    # Enter value of q
    # q = 4 is used here just as an example.
    # The user should replace this by a value of q
    # that is appropriate for the species and data
    # considered, e.g. that obtained by running CPT_Rcode
    # on part of the data.
    #q <- 6
    
    # input N
    # N = total number of permutations (1 observed and N-1 simulated) 
    # # N = 10000 is a convenient number
    N <- 10000
    
    # input tol
    # tol = tolerance
    # = maximum distance between indistinguishable positions # tol = 0 is a 
    # convenient default
    # The user may like to replace this by
    # twice the average GPS error in research area
    tol <- 0
    
    # PRELIMINARIES
    # Reverse the time-ordering,
    # so that (bx1[1], bx2[1]) refers to (final)
    #  putative goal
    bx1 <- 0*x1
    bx2 <- 0*x2
    
    n <- length(x1)
    for (j in 1:n){
      bx1[j] <- x1[n-j+1]
      bx2[j] <- x2[n-j+1]
    }
    # Calculate the steps (bxdiff1, bxdiff2)
    bx1diff <- diff(bx1)
    bx2diff <- diff(bx2)
    
    # REMOVE POINTS AT WHICH ANIMAL STAYS STILL
    # ind is (reverse) time ordering of points
    # newp > 0 if point differs from previous point
    tolsq <- tol^2
    ind <- c(1:n)
    newp <- c(1:n)
    for (j in 2:n){
      newp[j] <- ind[j]*( bx1diff[j-1]^2 + bx2diff[j-1]^2 > tolsq)
    }
    
    # bz1, bz2 are coordinates of points(in reverse time order) 
    # at which there is movement
    bz1 <- bx1[newp >0]
    bz2 <- bx2[newp >0]
    nz <- length(bz1)
    
    # We shall apply the CPT to points with coordinates (bz1,bz2)
    # Calculate the steps (bzdiff1, bzdiff2)
    bz1diff <- diff(bz1)
    bz2diff <- diff(bz2)
    # SOME DECLARATIONS
    # goal.no = number (from end) of current putative goal 
    # (with goal.no = 1 for end position
    
    # Thus goal.no = goal.t + 1
    # where goal.t was used in original code to refer to
    # time of current putative goal (with goal.t = 0 
    # for # end position # start with goal.no = 1
    goal.no <- 1
    
    # last.no = number (backwards in time) of last position of 
    # interest # last.no = length(bz1) is a convenient default
    # (and includes all the positions)
    last.no <- length(bz1)
    no.of.nos <- length(bz1)
    Rsumrand  <- rep(0, len=N)
    Rsumrandr <- rep(0, len=N)
    
    # Pr will store observed p-values in a run of r
    Pr <- rep(1, len=no.of.nos)
    
    # sig is a vector indicating whether or not
    # waypoint k is detected as a possible change point: 
    # sig[k] = 1 if change point detected at waypoint k 
    # sig[k] = 0 otherwise
    sig <- rep(0,nz)
    
    # k is number of steps in "k-leg"
    k <- 0
    
    # LOOK (SEQUENTIALLY) FOR NEXT POSSIBLE CHANGE POINT 
    
    while(goal.no < last.no - q){
      
      print(goal.no)
      
      k <- 0
      
      # INCREASE k UNTIL NEXT POSSIBLE CHANGE POINT IS FOUND 
      # Re-initialise Pr
      Pr <- 0*Pr + 1
      P <- 1
      
      while ((P > alpha) && (goal.no + q + k < last.no)){ 
        #B
        
        k <- k + 1
        
        R1 <- sqrt((bz1[goal.no+k] - bz1[goal.no])^2 + (bz2[goal.no+k] - 
                                                          bz2[goal.no])^2) 
        R2 <- sqrt((bz1[goal.no+k+q] - bz1[goal.no+k])^2 + (bz2[goal.no+k+q] - 
                                                              bz2[goal.no+k])^2)
        Rsum <- R1 + R2
        
        # Rsumrand[1] = observed value of statistic R1 + R2 
        Rsumrand[1] <- Rsum
        
        # Now calculate statistic R1 + R2 for a further N-1 
        # random permutations and store in Rsumrand
        for (it in 2:N){
          # start of it loop C
          u <- runif(k+q,0,1)
          perm <- order(u)
          bz1r <- bz1[goal.no]
          bz2r <- bz2[goal.no]
          for (j in 1:k){ # start of j loop D 
            bz1r <- bz1r + bz1diff[goal.no-1+perm[j]] 
            bz2r <- bz2r + bz2diff[goal.no-1+perm[j]]
          } # end of j loop D
          
          R1rand <- sqrt((bz1r - bz1[goal.no])^2 + (bz2r - bz2[goal.no])^2)
          R2rand <- sqrt((bz1[goal.no+k+q] - bz1r)^2 + 
                           (bz2[goal.no+k+q] - bz2r)^2) 
          Rsumrand[it] <- R1rand + R2rand
        } # end of it loop C
        
        P <- sum(Rsumrand >= Rsum)/N
        
        Pr[k] <- P
        
      } # end of "while" on
      # ((P > alpha) && (goal.no + q + k < last.no)) B
      # f is the first value of k in the current run that is significant 
      f <- k
      
      while ((P <= alpha) && (goal.no + q + k < last.no)){ 
        #E
        
        k <- k+1
        R1 <- sqrt((bz1[goal.no+k] - bz1[goal.no])^2 + 
                     (bz2[goal.no+k] - bz2[goal.no])^2)
        
        R2 <- sqrt((bz1[goal.no+k+q] - bz1[goal.no+k])^2 + 
                     (bz2[goal.no+k+q] - bz2[goal.no+k])^2)
        
        Rsum <- R1 + R2
        
        # Rsumrand[1] = observed value of statistic R1 + R2 
        Rsumrand[1] <- Rsum
        
        # Now calculate statistic R1 + R2 for a further N-1 
        # random permutations and store in Rsumrand
        for (it in 2:N){ # start of it loop F
          u <- runif(k+q,0,1)
          perm <- order(u)
          bz1r <- bz1[goal.no]
          bz2r <- bz2[goal.no]
          for (j in 1:k){ # start of j loop G 
            bz1r <- bz1r + bz1diff[goal.no-1+perm[j]] 
            bz2r <- bz2r + bz2diff[goal.no-1+perm[j]]
          } # end of j loop G
          
          R1rand <- sqrt((bz1r - bz1[goal.no])^2 + 
                           (bz2r - bz2[goal.no])^2)
          R2rand <- sqrt((bz1[goal.no+k+q] - bz1r)^2 + 
                           (bz2[goal.no+k+q] - bz2r)^2)      
          Rsumrand[it] <- R1rand + R2rand
        } # end of it loop F
        
        P <- sum(Rsumrand >= Rsum)/N
        
        Pr[k] <- P
        
      }  # end of "while" on
      
      # ((P < alpha) && (goal.no + q + k < last.no)) E
      
      # l is the last value of k in the current run that is significant 
      l <- k - 1
      
      # apply "peak rule"
      Pr[f:l]
      rmin <- min(which(Pr[f:l] == min(Pr[f:l]))) 
      rmin <- if(l >=f) rmin else 0
      
      goal.no <- ifelse(rmin > 0, goal.no + f + rmin - 1, last.no) 
      
      sig[goal.no] <- ifelse(goal.no == last.no,0,1)
      
    } # end of "while" on (goal.no < last.no - q) # A
    
    #  Remove putative goal from list of CP's
    sig[1] <- 0
    
    # OUTPUTS
    
    # OUTPUT OF
    # (first, last) (row nos. of first and last times at 
    # "significant" change points)
    # (east, north) (their coordinates)
    # cp lists indices of waypoints identified as change points
    cp <- which(sig==1)
    
    
    # bsig is sig in reverse time-order
    bsig <- rep(0,nz)
    for (j in 1:nz){
      bsig[j] <- sig[nz-j+1]
    }
    
    # "times" (row nos.) and coordinates of change points 
    newpp <- newp[newp > tol]
    cp.time <- newpp[sig==1]
    cp.bx1 <- bz1[sig==1]
    cp.bx2 <- bz2[sig==1]
    cp.no <- n + 1 - cp.time
    # last is cp.no in reverse order
    last <- 0*cp.no
    cp.leng <- length(cp.no)
    for (j in 1:cp.leng){
      last[j] <- cp.no[cp.leng-j+1]
    }
    # last contains row nos. of (last times at) change points 
    # (east, north) are their coordinates
    east <- x2[last]
    north <- x1[last]
    
    # first contains row nos. of first times at change points 
    first <- 0*last
    for (j in 1:length(last)){
      first[j] <- n+2 - min(which(newp > (n+1-last[j])))
    }
    
    # PLOT WAYPOINTS AND MARK CHANGE POINTS
    # For plotting purposes, reverse the order of
    # bz1, bz2 to get z1, z2
    z1 <- 0*bz1
    z2 <- 0*bz2
    nz <- length(z1)
    for (j in 1:nz){
      z1[j] <- bz1[nz-j+1]
      z2[j] <- bz2[nz-j+1]
    }
    # s is vector to index the points
    s <- c(1:nz)
    
    # Save change-points
    z1bsig <- z1[bsig==1]# Latitude of CPs
    z2bsig <- z2[bsig==1]# Longitude of CPs
    cpsame <- which(first == last)
    cpdiff <- which(first != last)
    lastsame <- last[cpsame]# Number of lines where CPs are located 
    firstdiff <- first[cpdiff]
    lastdiff <- last[cpdiff]
    z1bsame <- z1bsig[cpsame]
    z2bsame <- z2bsig[cpsame]
    z1bdiff <- z1bsig[cpdiff]
    z2bdiff <- z2bsig[cpdiff]
    
    
    # create a new column
    datainp_sma$cp <- "NO-CP"; head(datainp_sma)
    # Fill CPs in the change point locations
    true.index <- which((datainp_sma$longitude %in% z2bsig) & 
                          (datainp_sma$latitude %in% z1bsig))
    datainp_sma$cp[true.index] <- "CP"
    
    # fill q value
    datainp_sma$q <- q
    
    # plot
    plot(datainp_sma$longitude, datainp_sma$latitude, pch = 19, 
         main = datainp_sma$date[1])
    # lines
    lines(datainp_sma$longitude, datainp_sma$latitude)
    points(datainp_sma[datainp_sma$cp == "CP",]$longitude, 
           datainp_sma[datainp_sma$cp == "CP",]$latitude, pch = 19, col = 2)
    
    output_sma <- rbind(output_sma, datainp_sma)
  }
}
```

## Filtering duplicated lines so I don't count more than once the same "CP"
```{r}
output_sma_filter<- output_sma[,-c(1,3,4,5)]
head(output_sma_filter)
output_sma_filter<- output_sma_filter[!duplicated
                                            (output_sma_filter), ]
head(output_sma_filter)
```

## Summary to understand the number of "CPs" according to value of "q"
```{r}
count_cps_sma <- output_sma_filter %>% 
  dplyr::group_by(date, q) %>% 
  dplyr::summarize(number.of.CPs = sum(cp == "CP"))
head(count_cps_sma)
```

## Now I organize the data into wide format to plot and understand what is the best value of "q" to use:
```{r}
count_cps_sma_wide <- spread(count_cps_sma, key = date, 
                                value = number.of.CPs)
# Create "total"TOTAL" column with sum or row values:
count_cps_sma_wide <- count_cps_sma_wide %>%
  mutate(TOTAL=rowSums(.[setdiff(names(.),"q")]))
# Converting to data.frame:
count_cps_sma_wide<-as.data.frame(count_cps_sma_wide)
```

## Ploting "q" for Santa Maria:

### Plot of number of CPs per day according to values of "q" from 1 to 10. Light-gray lines represent number of CPs for each value of "q" in a unique day. Thick black line represent the sum of CPs for each value of "q" in all days.

```{r}
#tiff(".tiff", width = 11000, height = 6500, res = 1000)
par(mfrow=c(1,1))
plot(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"06/05/15"],
     type="o",pch=20,
     col="gray85",xlab="q",ylab ="Number of change- points - sma",
     xlim=c(1,max(count_cps_sma_wide[,"q"])),
     ylim=c(1,max(count_cps_sma_wide)),lab=c(10,20,0))
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"09/04/15"],
       type="o",pch=20,col="gray85")
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"11/04/15"],
       type="o",pch=20,col="gray85") 
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"12/04/15"],
       type="o",pch=20,col="gray85") 
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"15/03/15"],
       type="o",pch=20,col="gray85") 
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"15/04/15"],
       type="o",pch=20,col="gray85")
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"18/03/15"],
       type="o",pch=20,col="gray85") 
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"20/03/15"],
       type="o",pch=20,col="gray85") 
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"21/03/15"],
       type="o",pch=20,col="gray85") 
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"29/03/15"],
       type="o",pch=20,col="gray85") 
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"31/03/15"],
       type="o",pch=20,col="gray85") 
points(count_cps_sma_wide[,"q"],count_cps_sma_wide[,"TOTAL"],
       type="o",pch=20,col="black", lwd=3)
#dev.off()
```

## It looks like q = 5 is the best for running the analysis, since it is the last value at which the number of CPs increases continuously. 

## After finding q, I just have to create a subset of my "output_suzano" file with q = 5.
```{r}
cpt_sma <- subset(output_sma, q == 5)
cpt_sma <- subset(cpt_sma, select = -id)
cpt_sma$id <- 1:nrow(cpt_sma)
cpt_sma<-cpt_sma[,c(9,1,2,3,4,5,6,7,8)]
head(cpt_sma)
```

***

# Here I present the plot with all identified CPs and associated behaviors in Santa Maia. Also, I plot here 3 different zones (territory and fragment border and core area). For this, I have worked in the dataframes in excel previously to organize the data acoordingly.

### Identification of behaviors associated with direction change (CPs) along black lion tamarins' daily routes in Santa Maria. Pink triangles = frugivory (53%), blue diamonds = foraging (18%), green squares = resting (10%), brown upside-down triangules = long call (8%), red circles = locomotion (5%), beige stars = inactive (5%), purple cross = out of sight (3%). Little black dots and lines repesent GPS points recorded every 5 minutes.

```{r}
#tiff("sma_k95_k70.tiff", width = 11000, height = 6500, res = 1000)
ggplot()+
  geom_polygon(data = forest_sma, aes(x=X, y= Y),
               fill = "darkgreen", colour = "black", size = 1)+
  geom_polygon(data = k95_sma, aes(x=X, y= Y),
               fill = "palegreen4", colour = "black", size = 1)+
  geom_polygon(data = k70_sma, aes(x=X, y= Y),
               fill = "yellowgreen", colour = "black", size = .5)+
  #geom_polygon(data = terbordf_taquara, aes(x=long, y= lat, group=group),
  #fill = "mediumseagreen", colour = "black") +
  #geom_polygon(data = river_taquara, aes(x=long, y= lat, group=group),
  #fill = "cadetblue1", colour = "black", size = 1) +
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, size=2))+
  geom_point(data = output.sma1, aes(x = longitude, 
                                     y = latitude), size = 1) +
  geom_path(data = output.sma1, aes(x = longitude, 
                                    y = latitude), size = 0.2) +
  geom_point(data = output.sma2, aes(x = longitude, 
                                     y = latitude), size = 1) +
  geom_path(data = output.sma2, aes(x = longitude, 
                                    y = latitude), size = 0.2) +
  geom_point(data = output.sma3, aes(x = longitude, 
                                     y = latitude), size = 1) +
  geom_path(data = output.sma3, aes(x = longitude, 
                                    y = latitude), size = 0.2) +
  geom_point(data = output.sma4, aes(x = longitude, 
                                     y = latitude), size = 1) +
  geom_path(data = output.sma4, aes(x = longitude, 
                                    y = latitude), size = 0.2) +
  geom_point(data = output.sma5, aes(x = longitude, 
                                     y = latitude), size = 1) +
  geom_path(data = output.sma5, aes(x = longitude, 
                                    y = latitude), size = 0.2) +
  geom_point(data = output.sma6, aes(x = longitude, 
                                     y = latitude), size = 1) +
  geom_path(data = output.sma6, aes(x = longitude, 
                                    y = latitude), size = 0.2) +
  geom_point(data = output.sma7, aes(x = longitude, 
                                     y = latitude), size = 1) +
  geom_path(data = output.sma7, aes(x = longitude, 
                                    y = latitude), size = 0.2) +
  geom_point(data = output.sma8, aes(x = longitude, 
                                     y = latitude), size = 1) +
  geom_path(data = output.sma8, aes(x = longitude, 
                                    y = latitude), size = 0.2) +
  geom_point(data = output.sma9, aes(x = longitude, 
                                     y = latitude), size = 1) +
  geom_path(data = output.sma9, aes(x = longitude, 
                                    y = latitude), size = 0.2) +
  geom_point(data = output.sma10, aes(x = longitude, 
                                      y = latitude), size = 1) +
  geom_path(data = output.sma10, aes(x = longitude, 
                                     y = latitude), size = 0.2) +
  geom_point(data = output.sma11, aes(x = longitude, 
                                      y = latitude), size = 1) +
  geom_path(data = output.sma11, aes(x = longitude, 
                                     y = latitude), size = 0.2) +
  geom_point(data = output.sma[output.sma$cpout == "x",], 
             aes(x = longitude, y = latitude), col = "darkorchid2", 
             shape = 3, size = 3.5, stroke = 1.5 )+
  geom_point(data = output.sma[output.sma$cpinac == "x",], 
             aes(x = longitude, y = latitude), col = "cornsilk", 
             shape = 8, size = 4, stroke = 1)+
  geom_point(data = output.sma[output.sma$cpres == "x",], 
             aes(x = longitude, y = latitude), fill = "chartreuse2", 
             shape = 22, size = 4, stroke = 1)+
  geom_point(data = output.sma[output.sma$cpagonin == "x",], 
             aes(x = longitude, y = latitude), fill = "coral", 
             shape = 25, size = 4, stroke = 1)+
  #geom_point(data = output.sma[output.sma$cpgum == "x",], 
  #aes(x = longitude, y = latitude), fill = "purple", 
  #shape = 22, size = 4, stroke = 1) +
  geom_point(data = output.sma[output.sma$cplocom == "x",], 
             aes(x = longitude, y = latitude), fill = "red", 
             shape = 21, size = 4, stroke = 1)+
  geom_point(data = output.sma[output.sma$cpfor == "x",], 
             aes(x = longitude, y = latitude), fill = "cyan", 
             shape = 23, size = 4, stroke = 1)+
  geom_point(data = output.sma[output.sma$cplong == "x",], 
             aes(x = longitude, y = latitude), fill = "chocolate4", 
             shape = 25, size = 4, stroke = 1)+
  geom_point(data = output.sma[output.sma$cpffru == "x",], 
             aes(x = longitude, y = latitude), fill = "pink", 
             shape = 24, size = 4, stroke = 1)+
  #geom_point(data = output.sma[output.sma$cprubb == "x",], 
  #aes(x = longitude, y = latitude), fill = "aliceblue", 
  #shape = 21, size = 4, stroke = 1) +
  coord_fixed()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=2),
        plot.margin=unit(c(0.1,0,0,0),"cm"))+
  theme(axis.text.x = element_text(hjust=0.5, size = 8, colour ="Black"))+
  theme(axis.text.y = element_text(angle=90, hjust=0.5, 
                                   size = 8, colour ="Black"))+
  theme(axis.title.y = element_blank())+
  theme(axis.title.x = element_blank())+
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  annotation_scale(location = "tr", width_hint = 0.2, 
                   pad_x = unit(0.1, "in"), pad_y = unit(0.8, "in"))+
  annotation_north_arrow(location = "tr", which_north = "true", 
                         pad_x = unit(0.3, "in"), pad_y = unit(0.1, "in"),
                         style = north_arrow_fancy_orienteering)
#dev.off()
```
